---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# solomonR

<!-- badges: start -->
[![R-CMD-check](https://github.com/JUhalt/solomonR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JUhalt/solomonR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/JUhalt/solomonR/actions/workflows/pkgdown.yaml/badge.svg)](https://juhalt.github.io/solomonR/)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

The goal of solomonR is to ...

# solomonR

Tools to analyze **Solomon Four-Group Designs** (SFGD) with both the classic teaching workflow and a modern unified GLM:
- Classic flow: 2×2 posttest ANOVA → (if no sensitization) ANCOVA on pretested cells + posttest-only t, optional Stouffer combine (Braver & Braver, 1988).  
- Modern flow: one GLM with robust SEs + key contrasts, permutation p-values, effect sizes, and diagnostic helpers.

## Installation

```{r}
# install.packages("pak")
pak::pak("JUhalt/solomonR")
```

## Quick Start

```{r}
library(solomonR)

set.seed(1)
n <- 20
pretested <- c(rep(1, 2*n), rep(0, 2*n))
treat     <- c(rep(1,n), rep(0,n), rep(1,n), rep(0,n))
y_pre     <- c(rnorm(n, 50,10), rnorm(n, 50,10), rep(NA, 2*n))

# Data-generating process: treatment bumps posttest by 0.4 SD
eps <- rnorm(4*n, 0, 10)
y_post <- 50 + 0.5*ifelse(pretested==1 & !is.na(y_pre), scale(y_pre), 0) + 4*treat + eps

# Modern unified GLM
fit_g <- fit_solomon_glm(y = y_post, treat = treat, pretested = pretested, pretest_score = y_pre)
print(fit_g)            # t.test/ANOVA-like summary with key contrasts
print(summary(fit_g))   # structured summary
plot_solomon_gg(y_post, treat, pretested)  # quick visual

# Classic teaching flow
fit_c <- fit_solomon_classic(y_post, treat, pretested, y_pre)
print(fit_c)            # ANOVA → ANCOVA / Welch t → optional Stouffer Z

# Assumptions & diagnostics
check_solomon_assumptions(y_post, treat, pretested, y_pre)
```

## What you get
- **Key contrasts** (ATE, pretest×treat, simple effects) with **robust SEs** and **semi-partial R² (±CI)**
- **Hedges’ g (±CI)** for Groups 3 vs 4 (posttest-only)
- **Permutation tests** for contrasts, **diagnostics** (Brown–Forsythe, Shapiro by cell, ANCOVA slope homogeneity)
- **Base plot + ggplot** cell-means CIs

See the vignettes:

- Classic Solomon Analysis (Teaching Flow)
- Unified GLM for Solomon Four-Group Designs
