---
title: "The Classic Solomon Four-Group Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Classic Solomon Four-Group Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(solomonR)

data(
  "solomon_example",
  package = "solomonR"
)
```

## Why preserve the classic analysis?

The Solomon four-group design has a long methodological history. Earlier
approaches analyzed different pieces of the design sequentially, using
tests that came to be described as Tests A through I.

`solomonR` reproduces this historical workflow for two reasons:

1. it remains useful for teaching how the Solomon design developed; and
2. it allows researchers to reproduce analyses based on historical
   recommendations.

The classic workflow should not be interpreted as the default modern
analysis merely because it is historically important. `solomonR`
therefore distinguishes the historical procedure from the package's
modern model-based and randomization-based approaches.

The implementation draws especially on the historical discussions of
Huck and Sandler (1973), Walton Braver and Braver (1988), and later simulation
work examining the operating characteristics of conditional Solomon
procedures.

## The four groups

A Solomon four-group design contains:

| Group | Pretest | Treatment | Posttest |
|:-----:|:-------:|:---------:|:--------:|
| 1 | Yes | Yes | Yes |
| 2 | Yes | No | Yes |
| 3 | No | Yes | Yes |
| 4 | No | No | Yes |

Groups 1 and 2 permit a pretest-adjusted treatment comparison.

Groups 3 and 4 provide a treatment comparison that cannot itself have
been influenced by administration of the pretest.

Together, the four groups permit investigation of both the treatment
effect and possible **pretest sensitization**.

## Example data

`solomonR` includes an example dataset with all four Solomon groups.

```{r example-data}
library(solomonR)

data(solomon_example)

demo_preview <- head(
  solomon_example,
  6
)

demo_preview$y_post <- round(
  demo_preview$y_post,
  2
)

demo_preview$y_pre <- round(
  demo_preview$y_pre,
  2
)

knitr::kable(
  demo_preview,
  align = "rrrr"
)
```

The pretest variable is structurally absent for participants who were
not pretested. Those missing values are therefore part of the design,
not ordinary missing observations.

## Running the historical analysis

The complete historical analysis is available through
`fit_solomon_classic()`:

```{r}
classic <- with(
  solomon_example,
  fit_solomon_classic(
    y_post,
    treat,
    pretested,
    y_pre
  )
)

classic
```

The printed output reports all of the historical tests but marks the
tests actually reached by the historical decision sequence with
`[PATH]`.

This distinction is important.

A test can be **calculated** for teaching or inspection without having
been **reached** under the historical conditional procedure.

`plot_classic_flow()` draws the conditional sequence as a decision tree and
highlights the route these data took, with the p-value of each test reached:

```{r classic-flow, fig.width = 7, fig.height = 5.5, fig.alt = "Decision tree of the historical Test A to I sequence, with the path taken by the example data highlighted."}
plot_classic_flow(classic)
```

The caption repeats the caution discussed below: the sequence ending in
Test I is kept for teaching and replication, not as a recommended analysis.

## Tests A-D: the four-group posttest model

The first part of the classic workflow is based on the four-group
posttest model.

### Test A: Pretest x Treatment interaction

Test A asks whether the treatment effect differs between pretested and
unpretested participants.

In modern language, this is the pretest-by-treatment interaction and is
the primary statistical representation of pretest sensitization.

### Tests B and C: simple treatment effects

Test B estimates the treatment effect among pretested participants.

Test C estimates the treatment effect among unpretested participants.

### Test D: treatment main effect

Test D evaluates the treatment effect averaged equally across the
pretested and unpretested conditions.

In `solomonR`, this is represented by the contrast

\[
\beta_{\mathrm{treat}} +
\frac{1}{2}\beta_{\mathrm{treat}\times\mathrm{pretest}}.
\]

This equal-weighted contrast is important because the treatment
coefficient by itself represents the treatment effect only in the
reference pretest condition.

The individual historical results are available from the fitted object.
For example:

```{r classic-ad-access, results='hide'}
classic$tests$A$result
classic$tests$D$result
```

## Tests E-G: analyses within the pretested groups

Historical approaches proposed several ways to make use of the baseline
measurement available in Groups 1 and 2.

### Test E: ANCOVA

Test E compares treatment and control among pretested participants while
adjusting posttest scores for the pretest.

This is the default pretested-group method in
`fit_solomon_classic()`.

```{r classic-e-access, results='hide'}
classic$tests$E$result
```

### Test F: gain-score analysis

Test F compares change scores between the two pretested groups.

```{r classic-f-access, results='hide'}
classic$tests$F$result
```

The change being compared is visible in the pretested groups' pretest and
posttest means. `plot_solomon_change()` shows them beside the unpretested
groups, which are observed at posttest only:

```{r change-plot, fig.width = 7, fig.height = 4.5, fig.alt = "Pretest and posttest means of the two pretested groups, beside the posttest means of the two unpretested groups."}
with(solomon_example, plot_solomon_change(y_post, treat, pretested, y_pre))
```

### Test G: repeated-measures formulation

Test G represents the corresponding two-wave treatment-by-time
comparison.

For two measurement occasions, the inferential contrast is
algebraically connected to the gain-score comparison. `solomonR`
retains Test G as a separate historical label because the distinction
is useful pedagogically.

```{r classic-g-access, results='hide'}
classic$tests$G$result
```

The pretested-group analysis can also be selected explicitly when
reproducing a particular historical strategy.

```{r eval=FALSE}
fit_solomon_classic(
  y_post,
  treat,
  pretested,
  y_pre,
  pretested_test = "gain"
)
```

## Test H: posttest-only comparison

Test H compares Groups 3 and 4, which did not receive the pretest.

```{r classic-h-access, results='hide'}
classic$tests$H$result
```

Because these groups were never pretested, this comparison provides a
direct treatment contrast uncontaminated by administration of a
baseline measure.

`solomonR` also reports Hedges' \(g\) for this posttest-only
comparison when available.

```{r classic-g}
cat(
  sprintf(
    "Hedges' g = %.2f, 95%% CI [%.2f, %.2f]",
    classic$g_post["g"],
    classic$g_post["lower"],
    classic$g_post["upper"]
  )
)
```

## Test I: Walton Braver & Braver (1988) Stouffer combination

Walton Braver and Braver (1988) proposed combining evidence from the
pretested and unpretested treatment comparisons using Stouffer's
method.

When requested, `solomonR` reproduces this historical analysis using
**directional one-tailed p-values aligned with the same treatment
direction**.

```{r classic-test-i}
cat(
  sprintf(
    "Stouffer Z = %.2f, p = %s",
    classic$tests$I$result$z,
    if (
      classic$tests$I$result$p.value < .001
    ) {
      "< .001"
    } else {
      sub(
        "^0",
        "",
        sprintf(
          "%.3f",
          classic$tests$I$result$p.value
        )
      )
    }
  )
)
```

This detail matters. A directional one-tailed p-value is not obtained
correctly by mechanically dividing every two-sided p-value by two
without considering the sign of the effect. The package's replication of
the published error rates found that the published simulations appear to
have combined the tests without regard to sign. `solomonR` refers the
combined z to its two-tailed p-value, as Walton Braver and Braver report it
in their worked example (1988, p. 153). That is why they report
far fewer Test I rejections than the procedure as defined produces (see
[Historical Tests: Replicating the Published Error
Rates](https://juhalt.github.io/solomonR/articles/classic-validation.html)).

## Versions of the decision sequence

The sequence changed after it was published, and `fit_solomon_classic()`
reproduces each version through its `flow` argument.

- **1988, the default.** In the original sequence, testing stops at the
  first significant test among D, the selected pretested-groups test, and H;
  Test I is reached only when all of them are nonsignificant (Walton Braver
  & Braver, 1988, pp. 151–153).
- **1990.** Replying to Sawilowsky and Markman (1990a), the authors agreed
  that using the meta-analysis selectively introduces selection bias. Once
  Tests A and D are nonsignificant, the analyst "should complete all the
  tests through Test I" (Braver & Walton Braver, 1990, p. 322), and Test I is
  regarded as the most definitive test.
- **1995.** A later revision removed Test D. It is unpublished and known
  from Sawilowsky (1996, p. 2), so it is cited as Walton Braver and Braver
  (1995, as cited in Sawilowsky, 1996). The remaining tests follow the 1988
  rule, as in Sawilowsky's simulation of the revised sequence.

```{r flow-1990}
classic_1990 <- with(
  solomon_example,
  fit_solomon_classic(y_post, treat, pretested, y_pre, flow = "1990")
)

classic_1990$path_string
classic_1990$conclusion
```

The tests themselves are identical in every version; only the path and the
conclusion differ. For the 1995 version, `alpha_allocation` applies the
test-wise significance levels that Sawilowsky (1996, Table 4) proposed to
control the experiment-wise error rate:

```{r flow-1995}
classic_1995 <- with(
  solomon_example,
  fit_solomon_classic(y_post, treat, pretested, y_pre, flow = "1995",
                      alpha_allocation = "method2_conservative")
)

classic_1995$path_string
```

## History and maturation

The four-group design also allows a check that does not involve the
treatment.
The unpretested control group's posttest (O6) was measured at the same time
as the other posttests, but its participants received neither the pretest
nor the treatment. Comparing it with the pretests of the pretested groups
(O1 and O3), which were measured before any treatment, estimates the
combined effect of history and maturation between the two occasions. Mai et
al. (2020, p. 8) report these comparisons as independent-samples t tests,
which `fit_solomon_classic()` returns as `history`:

```{r history}
classic$history
```

The check assumes random assignment and a measure that is comparable at the
two occasions.

## Why Test I is treated cautiously

The Stouffer procedure is included because it is part of the history of
the Solomon design and may be required when reproducing historical
analyses.

It is not the default modern recommendation in `solomonR`.

Later simulation work showed that conditional analysis sequences can
have undesirable experiment-wise Type I error properties. In
particular, choosing later tests based on the statistical significance
of earlier tests changes the operating characteristics of the overall
procedure. Sawilowsky et al. (1994) found that the 1988 sequence falsely
declared an effect about 14% of the time at a nominal 5% per test.

The package's own replication found similar rates for the sequence as
Walton Braver and Braver define it. With Test I judged two-tailed, as in
their worked example, the rate was 13.5% to 13.7% for the 1988 and 1995
versions, and Sawilowsky's (1996) alpha allocations stayed above their
targets.

For contemporary applied work, researchers should therefore consider a
single prespecified model or a randomization-based analysis rather than
automatically following a significance-driven testing tree.

See the unified GLM vignette for the primary modern observed-variable
workflow:

```{r eval=FALSE}
vignette("glm-solomon", package = "solomonR")
```

## Historical analysis versus modern analysis

The two approaches answer related questions but serve different
purposes.

| Goal | Suggested approach |
|---|---|
| Teach the historical Solomon procedure | `fit_solomon_classic()` |
| Reproduce a historical Tests A-I analysis | `fit_solomon_classic()` |
| Estimate Solomon effects in one model | `fit_solomon_glm()` |
| Conduct randomization-based inference | `perm_solomon()` |
| Use full-information likelihood | `fit_solomon_ml()` |
| Model latent outcomes | `fit_solomon_sem_latent()` |

The historical procedure is therefore preserved rather than erased,
while modern methods are available alongside it. [A History of the Solomon
Design and Its Analysis](https://juhalt.github.io/solomonR/articles/history.html) tells how each version arose
and how the debate over it unfolded.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](https://juhalt.github.io/solomonR/articles/references.html) page.

## References
Braver, S. L., & Walton Braver, M. C. (1990). Meta-analysis for Solomon
four-group designs reconsidered: A reply to Sawilowsky and Markman.
*Perceptual and Motor Skills, 71*(1), 321–322.
https://doi.org/10.2466/pms.1990.71.1.321

Huck, S. W., & Sandler, H. M. (1973). A note on the Solomon 4-group design:
Appropriate statistical analyses. *The Journal of Experimental Education,
42*(2), 54–55. https://doi.org/10.1080/00220973.1973.11011460

Mai, N. N., Takahashi, Y., & Oo, M. M. (2020). Testing the effectiveness of
transfer interventions using Solomon four-group designs. *Education
Sciences, 10*(4), Article 92. https://doi.org/10.3390/educsci10040092

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
http://digitalcommons.wayne.edu/coe_tbf/29

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S. (1994).
Meta-analysis and the Solomon four-group design. *The Journal of
Experimental Education, 62*(4), 361–376.
https://doi.org/10.1080/00220973.1994.9944140

Sawilowsky, S. S., & Markman, B. S. (1990a). Another look at the power of
meta-analysis in the Solomon four-group design. *Perceptual and Motor
Skills, 71*(1), 177–178. https://doi.org/10.2466/pms.1990.71.1.177

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150
