---
title: "The Classic Solomon Four-Group Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Classic Solomon Four-Group Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(solomonR)

data(
  "solomon_example",
  package = "solomonR"
)
```

## Why preserve the classic analysis?

The Solomon four-group design has a long methodological history. Earlier
approaches analyzed different pieces of the design sequentially, using
tests that came to be described as Tests A through I.

`solomonR` reproduces this historical workflow for two reasons:

1. it remains useful for teaching how the Solomon design developed; and
2. it allows researchers to reproduce analyses based on historical
   recommendations.

The classic workflow should not be interpreted as the default modern
analysis merely because it is historically important. `solomonR`
therefore distinguishes the historical procedure from the package's
modern model-based and randomization-based approaches.

The implementation draws especially on the historical discussions of
Huck and Sandler (1973), Braver and Braver (1988), and later simulation
work examining the operating characteristics of conditional Solomon
procedures.

## The four groups

A Solomon four-group design contains:

| Group | Pretest | Treatment | Posttest |
|:-----:|:-------:|:---------:|:--------:|
| 1 | Yes | Yes | Yes |
| 2 | Yes | No | Yes |
| 3 | No | Yes | Yes |
| 4 | No | No | Yes |

Groups 1 and 2 permit a pretest-adjusted treatment comparison.

Groups 3 and 4 provide a treatment comparison that cannot itself have
been influenced by administration of the pretest.

Together, the four groups permit investigation of both the treatment
effect and possible **pretest sensitization**.

## Example data

`solomonR` includes an example dataset with all four Solomon groups.

```{r example-data}
library(solomonR)

data(solomon_example)

demo_preview <- head(
  solomon_example,
  6
)

demo_preview$y_post <- round(
  demo_preview$y_post,
  2
)

demo_preview$y_pre <- round(
  demo_preview$y_pre,
  2
)

knitr::kable(
  demo_preview,
  align = "rrrr"
)
```

The pretest variable is structurally absent for participants who were
not pretested. Those missing values are therefore part of the design,
not ordinary missing observations.

## Running the historical analysis

The complete historical analysis is available through
`fit_solomon_classic()`:

```{r}
classic <- with(
  solomon_example,
  fit_solomon_classic(
    y_post,
    treat,
    pretested,
    y_pre
  )
)

classic
```

The printed output reports all of the historical tests but marks the
tests actually reached by the historical decision sequence with
`[PATH]`.

This distinction is important.

A test can be **calculated** for teaching or inspection without having
been **reached** under the historical conditional procedure.

## Tests A-D: the four-group posttest model

The first part of the classic workflow is based on the four-group
posttest model.

### Test A: Pretest x Treatment interaction

Test A asks whether the treatment effect differs between pretested and
unpretested participants.

In modern language, this is the pretest-by-treatment interaction and is
the primary statistical representation of pretest sensitization.

### Tests B and C: simple treatment effects

Test B estimates the treatment effect among pretested participants.

Test C estimates the treatment effect among unpretested participants.

### Test D: treatment main effect

Test D evaluates the treatment effect averaged equally across the
pretested and unpretested conditions.

In `solomonR`, this is represented by the contrast

\[
\beta_{\mathrm{treat}} +
\frac{1}{2}\beta_{\mathrm{treat}\times\mathrm{pretest}}.
\]

This equal-weighted contrast is important because the treatment
coefficient by itself represents the treatment effect only in the
reference pretest condition.

The individual historical results are available from the fitted object.
For example:

```{r classic-ad-access, results='hide'}
classic$tests$A$result
classic$tests$D$result
```

## Tests E-G: analyses within the pretested groups

Historical approaches proposed several ways to make use of the baseline
measurement available in Groups 1 and 2.

### Test E: ANCOVA

Test E compares treatment and control among pretested participants while
adjusting posttest scores for the pretest.

This is the default pretested-group method in
`fit_solomon_classic()`.

```{r classic-e-access, results='hide'}
classic$tests$E$result
```

### Test F: gain-score analysis

Test F compares change scores between the two pretested groups.

```{r classic-f-access, results='hide'}
classic$tests$F$result
```

### Test G: repeated-measures formulation

Test G represents the corresponding two-wave treatment-by-time
comparison.

For two measurement occasions, the inferential contrast is
algebraically connected to the gain-score comparison. `solomonR`
retains Test G as a separate historical label because the distinction
is useful pedagogically.

```{r classic-g-access, results='hide'}
classic$tests$G$result
```

The pretested-group analysis can also be selected explicitly when
reproducing a particular historical strategy.

```{r eval=FALSE}
fit_solomon_classic(
  y_post,
  treat,
  pretested,
  y_pre,
  pretested_test = "gain"
)
```

## Test H: posttest-only comparison

Test H compares Groups 3 and 4, which did not receive the pretest.

```{r classic-h-access, results='hide'}
classic$tests$H$result
```

Because these groups were never pretested, this comparison provides a
direct treatment contrast uncontaminated by administration of a
baseline measure.

`solomonR` also reports Hedges' \(g\) for this posttest-only
comparison when available.

```{r classic-g}
cat(
  sprintf(
    "Hedges' g = %.2f, 95%% CI [%.2f, %.2f]",
    classic$g_post["g"],
    classic$g_post["lower"],
    classic$g_post["upper"]
  )
)
```

## Test I: Braver & Braver (1988) Stouffer combination

Braver and Braver (1988) proposed combining evidence from the
pretested and unpretested treatment comparisons using Stouffer's
method.

When requested, `solomonR` reproduces this historical analysis using
**directional one-tailed p-values aligned with the same treatment
direction**.

```{r classic-test-i}
cat(
  sprintf(
    "Stouffer Z = %.2f, p = %s",
    classic$tests$I$result$z,
    if (
      classic$tests$I$result$p.value < .001
    ) {
      "< .001"
    } else {
      sub(
        "^0",
        "",
        sprintf(
          "%.3f",
          classic$tests$I$result$p.value
        )
      )
    }
  )
)
```

This detail matters. A directional one-tailed p-value is not obtained
correctly by mechanically dividing every two-sided p-value by two
without considering the sign of the effect.

## Why Test I is treated cautiously

The Stouffer procedure is included because it is part of the history of
the Solomon design and may be required when reproducing historical
analyses.

It is not the default modern recommendation in `solomonR`.

Later simulation work showed that conditional analysis sequences can
have undesirable experiment-wise Type I error properties. In
particular, choosing later tests based on the statistical significance
of earlier tests changes the operating characteristics of the overall
procedure.

For contemporary applied work, researchers should therefore consider a
single prespecified model or a randomization-based analysis rather than
automatically following a significance-driven testing tree.

See the unified GLM vignette for the primary modern observed-variable
workflow:

```{r eval=FALSE}
vignette("glm-solomon", package = "solomonR")
```

## Historical analysis versus modern analysis

The two approaches answer related questions but serve different
purposes.

| Goal | Suggested approach |
|---|---|
| Teach the historical Solomon procedure | `fit_solomon_classic()` |
| Reproduce a historical Tests A-I analysis | `fit_solomon_classic()` |
| Estimate Solomon effects in one model | `fit_solomon_glm()` |
| Conduct randomization-based inference | `perm_solomon()` |
| Use full-information likelihood | `fit_solomon_ml()` |
| Model latent outcomes | `fit_solomon_sem_latent()` |

The historical procedure is therefore preserved rather than erased,
while modern methods are available alongside it.

## References

Braver, M. W., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*, 150-154.

Huck, S. W., & Sandler, H. M. (1973). A note on the Solomon 4-group
design: Appropriate statistical analyses. *The Journal of Experimental
Education, 42*, 54-55.

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S.
(1994). Meta-analysis and the Solomon four-group design. *The Journal
of Experimental Education, 62*, 361-376.
