---
title: "Modern Analysis of the Solomon Four-Group Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modern Analysis of the Solomon Four-Group Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## A unified approach

The Solomon four-group design is often introduced through a sequence
of separate analyses.

A modern alternative is to represent the four-group design in a single
model and estimate the scientifically relevant questions as explicit
contrasts.

For a continuous posttest outcome, the core model in `solomonR` is

\[
Y =
\beta_0 +
\beta_T T +
\beta_P P +
\beta_{TP}(T \times P) +
\beta_X X_{\mathrm{obs}} +
\varepsilon,
\]

where:

- \(T\) indicates treatment assignment;
- \(P\) indicates whether a participant was pretested;
- \(T \times P\) represents pretest sensitization; and
- \(X_{\mathrm{obs}}\) incorporates the pretest information available
  in the pretested groups.

`solomonR` then estimates four predefined Solomon contrasts from this
model.

## Structural missingness of the pretest

One unusual feature of the Solomon design deserves special attention.

Participants in Groups 3 and 4 were deliberately **not pretested**.
Their absent pretest values are therefore structurally missing by
design.

Simply placing the raw pretest variable into an ordinary regression
formula can cause complete-case deletion of Groups 3 and 4, destroying
the four-group design.

`fit_solomon_glm()` avoids this problem internally. For the optional
pretest covariate, the model uses the observed pretest score among
pretested participants and a design-safe value among participants for
whom the pretest was never administered.

The pretest indicator remains in the model, so this coding does not
pretend that Groups 3 and 4 actually had baseline scores of zero.

## Example data

```{r example-data}
library(solomonR)

data(solomon_demo)

demo_preview <- head(
  solomon_demo,
  6
)

demo_preview$y_post <- round(
  demo_preview$y_post,
  2
)

demo_preview$y_pre <- round(
  demo_preview$y_pre,
  2
)

knitr::kable(
  demo_preview,
  align = "rrrr"
)
```

## Fit the model

A useful default for continuous outcomes is the unified model with HC3
heteroskedasticity-robust covariance estimation.

```{r}
fit <- with(
  solomon_demo,
  fit_solomon_glm(
    y = y_post,
    treat = treat,
    pretested = pretested,
    pretest_score = y_pre,
    robust = "HC3"
  )
)

fit
```

The output contains model coefficients followed by the four principal
Solomon contrasts.

The contrasts can also be accessed directly:

```{r glm-effects}
effects_display <- fit$effects[
  ,
  c(
    "contrast",
    "estimate",
    "std.error",
    "statistic",
    "p.value",
    "r2"
  )
]

effects_display$estimate <- sprintf(
  "%.2f",
  effects_display$estimate
)

effects_display$std.error <- sprintf(
  "%.2f",
  effects_display$std.error
)

effects_display$statistic <- sprintf(
  "%.2f",
  effects_display$statistic
)

effects_display$p.value <- ifelse(
  effects_display$p.value < .001,
  "< .001",
  sub(
    "^0",
    "",
    sprintf(
      "%.3f",
      effects_display$p.value
    )
  )
)

effects_display$r2 <- sub(
  "^0",
  "",
  sprintf(
    "%.3f",
    effects_display$r2
  )
)

names(effects_display) <- c(
  "Contrast",
  "Estimate",
  "SE",
  "z",
  "p",
  "Wald R2"
)

knitr::kable(
  effects_display,
  align = c(
    "l",
    "r",
    "r",
    "r",
    "r",
    "r"
  )
)
```

## The four Solomon estimands

### Average treatment effect

The reported ATE is the treatment effect averaged equally over the two
pretest conditions:

\[
ATE =
\beta_T +
\frac{1}{2}\beta_{TP}.
\]

This distinction is important.

With the unpretested condition used as the reference group,
\(\beta_T\) by itself is the treatment effect among **unpretested**
participants. It is not the equal-weighted Solomon ATE.

### Pretest x Treatment

The sensitization contrast is

\[
\beta_{TP}.
\]

It asks whether the treatment effect differs according to whether the
pretest was administered.

A statistically nonsignificant interaction should not be interpreted
as proof that sensitization is absent. Formal equivalence procedures
for that stronger question are planned for a future version of
`solomonR`.

### Treatment effect among pretested participants

The treatment effect among participants who received the pretest is

\[
\beta_T + \beta_{TP}.
\]

### Treatment effect among unpretested participants

The treatment effect among participants who did not receive the
pretest is

\[
\beta_T.
\]

Together, these contrasts describe both the treatment effect and its
possible modification by pretesting.

## Why HC3?

The `robust` argument controls covariance estimation.

```{r eval=FALSE}
fit_solomon_glm(
  y_post,
  treat,
  pretested,
  y_pre,
  robust = "none"
)

fit_solomon_glm(
  y_post,
  treat,
  pretested,
  y_pre,
  robust = "HC3"
)
```

HC3 provides heteroskedasticity-robust standard errors and is a useful
default when equal residual variances are uncertain.

A CR2 covariance option is also available for clustered data:

```{r eval=FALSE}
fit_solomon_glm(
  y_post,
  treat,
  pretested,
  y_pre,
  robust = "CR2",
  cluster = cluster_id
)
```

CR2 combines the bias-reduced cluster-robust covariance estimator (Bell &
McCaffrey, 2002) with Satterthwaite degrees of freedom for each
coefficient and contrast (Pustejovsky & Tipton, 2018). The degrees of
freedom are reported in the `df` column and can be small when there are
few clusters.

## Wald-based partial R-squared

For Gaussian models, `fit_solomon_glm()` reports a Wald-based partial
\(R^2\) for each one-degree-of-freedom contrast.

With conventional Gaussian OLS covariance,

\[
R^2_{\mathrm{partial}} =
\frac{t^2}{t^2 + df_{\mathrm{residual}}}.
\]

When a robust covariance estimator such as HC3 is used, `solomonR`
applies the same transformation to the robust Wald statistic. In that
case the result should be interpreted as a **descriptive Wald-based
approximation**, not an exact decomposition of model variance.

Confidence intervals for this quantity are not currently reported.

```{r glm-r2}
r2_display <- fit$effects[
  ,
  c(
    "contrast",
    "r2"
  )
]

r2_display$r2 <- sub(
  "^0",
  "",
  sprintf(
    "%.3f",
    r2_display$r2
  )
)

names(r2_display) <- c(
  "Contrast",
  "Wald R2"
)

knitr::kable(
  r2_display,
  align = c("l", "r")
)
```

## Diagnostics

The package provides design-relevant diagnostics through
`check_solomon_assumptions()`.

```{r}
checks <- with(
  solomon_demo,
  check_solomon_assumptions(
    y_post,
    treat,
    pretested,
    y_pre
  )
)

checks
```

These include variance checks, cell-level distribution summaries, and
assessment of ANCOVA slope homogeneity.

Diagnostics should inform analysis choices rather than be treated as a
mechanical series of pass/fail gates.

## Randomization-based inference

If treatment assignment was genuinely randomized within the Solomon
pretesting conditions, treatment-label permutation provides another
inferential approach.

```{r}
perm <- perm_solomon(
  fit,
  contrast = "ATE (avg over pretest)",
  reps = 1000,
  seed = 42,
  return_dist = TRUE
)

perm
```

`perm_solomon()` shuffles treatment labels **within pretest strata**.
This preserves the number of treated and control participants within
the pretested and unpretested portions of the design.

The test uses an HC3-studentized statistic.

Its Monte Carlo p-value uses a finite-simulation correction:

\[
p =
\frac{B_{\mathrm{extreme}} + 1}
     {B_{\mathrm{valid}} + 1}.
\]

This prevents an estimated permutation p-value of exactly zero.

The null distribution can be visualized with:

```{r fig.width=7, fig.height=4.5}
plot_perm(perm)
```

Randomization inference is justified by the assignment mechanism, not
merely by a small sample size. If treatment was randomized at the
cluster level, permutation must likewise occur at the cluster level.
`perm_solomon()` permutes individuals, so it refuses fits that include a
clustering variable; use the CR2 tests above for clustered designs.

## A likelihood-based alternative

`fit_solomon_ml()` provides a full-information maximum-likelihood
analysis inspired by van Engelenburg's treatment of the Solomon
four-group design.

```{r}
ml <- with(
  solomon_demo,
  fit_solomon_ml(
    y_post,
    treat,
    pretested,
    y_pre
  )
)

ml
```

A useful feature of the likelihood formulation is that the structurally
absent pretests in Groups 3 and 4 are treated as part of the design
rather than as ordinary missing baseline observations.

The ML and GLM approaches target closely related Solomon effects but use
different assumptions for uncertainty estimation.

## Historical analysis

For teaching or reproducing the classical Tests A-I workflow, use:

```{r}
classic <- with(
  solomon_demo,
  fit_solomon_classic(
    y_post,
    treat,
    pretested,
    y_pre
  )
)

classic
```

See:

```{r eval=FALSE}
vignette("classic-solomon", package = "solomonR")
```

for a detailed explanation of the historical sequence and its
limitations.

## SEM extensions

`solomonR` also includes observed- and latent-variable SEM approaches.

An observed four-group mean model can be fit with:

```{r eval=FALSE}
fit_solomon_sem(
  y_post,
  treat,
  pretested
)
```

The unrestricted four-group observed mean model is saturated, so global
fit indices are not diagnostic for that model.

For multi-item outcomes, `fit_solomon_sem_latent()` provides latent
Solomon contrasts. Latent mean comparisons require scalar measurement
invariance across the four Solomon groups.

These SEM tools extend the same basic estimands used throughout the
package:

- ATE;
- pretest sensitization;
- treatment effect among pretested participants; and
- treatment effect among unpretested participants.

## Choosing an analysis

For many randomized Solomon studies with continuous outcomes, a useful
workflow is:

1. describe the four groups and inspect the data;
2. fit the unified GLM with prespecified contrasts;
3. examine the pretest-by-treatment contrast directly;
4. inspect diagnostics;
5. use randomization inference when justified by the treatment
   assignment mechanism; and
6. use ML or SEM when the scientific question or measurement structure
   warrants those approaches.

The historical decision tree remains valuable for teaching, but a
modern analysis does not need to choose its inferential procedure based
on whether an earlier significance test crossed \(p=.05\).

## Where solomonR is heading

The package is under active development.

Planned additions include:

- equivalence testing for pretest sensitization;
- automated design validation;
- structural-versus-incidental missingness diagnostics;
- method-comparison tools;
- Solomon-specific visualization;
- redesigned power and sample-size planning;
- generalized outcomes;
- clustered and longitudinal extensions; and
- expanded SEM workflows.

See `ROADMAP.md` in the package repository for the current development
plan.
