---
title: "Binary Outcomes: Validating marginal_solomon()"
description: >
  Simulation study of the marginal Solomon contrasts for binary outcomes, the
  conditional logistic contrast, and the historical categorical rule.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("binary-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("binary-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

scale_label <- c(difference = "Risk difference", ratio = "Risk ratio", odds_ratio = "Odds ratio")
method_label <- c(bootstrap = "Bootstrap (percentile)", delta = "Delta method (HC3)",
                  unadjusted = "Unadjusted, delta method (HC3)")

sc <- unique(p[, c("scenario", "n", "control_risk", "bX", "pattern", "fit_failures",
                   "unadjusted_failures", "interval_failures")])

m <- p[p$method %in% names(method_label), ]
m$failed <- ifelse(m$method == "bootstrap", m$fit_failures + m$interval_failures,
                   ifelse(m$method == "delta", m$fit_failures, m$unadjusted_failures))
m$supported <- m$failed <= 0.05 * reps
m$null <- abs(m$true_value) < 1e-8
# Tolerances declared on issue #43 before any run.
m$bias_ok <- abs(m$bias) <= 2 * m$bias_mcse
m$cov_ok <- m$coverage >= 0.94 & m$coverage <= 0.96
m$se_ok <- abs(m$relerr) <= 10
m$t1_ok <- ifelse(m$null, m$rejection >= 0.04 & m$rejection <= 0.06, TRUE)
m$all_ok <- m$bias_ok & m$cov_ok & m$se_ok & m$t1_ok
s <- m[m$supported, ]

pct <- function(x) sprintf("%.1f%%", 100 * x)
rng <- function(x, d = 3) if (length(x)) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x)) else "--"
```

This article reports the simulation validation of `marginal_solomon()`, the
package's estimator of the Solomon contrasts for binary outcomes, together
with two comparisons that motivated it. The aims, data-generating mechanisms,
estimands, methods, performance measures, and tolerances were posted on
[issue #43](https://github.com/JUhalt/solomonR/issues/43) before any run,
following the ADEMP structure of Morris et al. (2019). One amendment, a fuller
definition of a failed logistic fit, was posted before any run.

The first full run stopped with a numerical error in the bootstrap refit and
produced no results. The error was fixed, a regression test was added, and
the study was rerun unchanged; both events are logged on the issue. The
results below come from package commit `r substr(run$commit, 1, 7)`,
`r reps` replications per scenario, and `r run$bootstrap_R` bootstrap
resamples per dataset.

## Design

**Data.** Each participant has a latent baseline X ~ N(0, 1), observed only
in the pretested groups, and a binary outcome with

logit Pr(Y = 1) = b0 + bX X + bT T + bP P + bTP T P.

The 48 scenarios cross participants per cell (20, 50, 100), the control risk
(0.1, 0.3), the association of the pretest with the outcome (bX = 0 or 1 on
the log-odds scale), and four effect patterns:

- no effects;
- a treatment effect only (bT = log 2);
- a treatment effect and a pretest effect (bT = log 2, bP = log 1.5);
- sensitization only (bTP = log 2).

**Estimands.** The marginal risk in each cell, averaged over X by numerical
integration, gives the four Solomon contrasts on three scales: risk
difference, log risk ratio, and log odds ratio. Sensitization is zero on every
scale under the first two patterns. Under the third it is nonzero on the
difference and ratio scales, which shows that sensitization depends on the
scale.

**Methods.** `marginal_solomon()` on the pretest-adjusted logistic fit, with
bootstrap and with delta-method intervals; the same contrasts without the
pretest (the unadjusted comparator); the logit-scale contrasts of
`fit_solomon_glm(family = binomial())`; and the historical rule of
`fisher_solomon()` (El Karkri et al., 2025b).

**Tolerances.** Bias within 2 Monte Carlo standard errors (MCSE); coverage of
95% intervals from 0.940 to 0.960; model standard error within 10% of the
empirical standard error; and, where the true value is zero, a rejection rate
from 0.040 to 0.060. A scenario in which more than 5% of datasets fail is
outside the method's supported range.

## Failures and the supported range

With few events, a Solomon cell can have no events at all, and the logistic
model then has no estimate. The bootstrap fails more often, because a
resample can empty a cell that the data did not.

```{r failures}
f <- aggregate(cbind(fit = fit_failures, boot = fit_failures + interval_failures) ~ n + control_risk,
               sc, function(x) max(x) / reps)
f <- f[order(f$control_risk, f$n), ]
knitr::kable(data.frame(
  `Per cell` = f$n, `Control risk` = f$control_risk,
  `Datasets with a failed fit (max)` = pct(f$fit),
  `Datasets without a bootstrap interval (max)` = pct(f$boot),
  check.names = FALSE
), align = "rrrr")
```

With 20 participants per cell and a control risk of 0.1, up to
`r pct(max(sc$fit_failures[sc$n == 20 & sc$control_risk == 0.1]) / reps)` of
datasets have no estimate, so these scenarios are outside every method's
supported range. The bootstrap is also outside its range with 20 per cell and
a control risk of 0.3, and with 50 per cell and a control risk of 0.1.

## Risk differences

```{r rd}
rd <- s[s$scale == "difference", ]
tab <- aggregate(cbind(cells = 1, pass = all_ok) ~ method + n + control_risk, rd, sum)
tab <- tab[order(tab$method, tab$control_risk, tab$n), ]
knitr::kable(data.frame(
  Method = method_label[tab$method], `Per cell` = tab$n, `Control risk` = tab$control_risk,
  `Contrasts meeting every tolerance` = sprintf("%d of %d", tab$pass, tab$cells),
  check.names = FALSE
), row.names = FALSE, align = "lrrr")
```

On the risk-difference scale the delta method met every tolerance in
`r sum(rd$all_ok[rd$method == "delta"])` of `r sum(rd$method == "delta")`
supported scenario-contrasts. Its coverage ranged from
`r rng(rd$coverage[rd$method == "delta"])` and its Type I error, where the
true value is zero, from `r rng(rd$rejection[rd$method == "delta" & rd$null])`.
The bootstrap's coverage ranged from
`r rng(rd$coverage[rd$method == "bootstrap"])`, slightly below nominal in
some scenarios. The expectation stated in the protocol, that the delta method
would under-cover in small samples with a low control risk (Localio et al.,
2007), was not borne out: with the HC3 covariance the package uses, its
coverage fell below 0.940 in
`r sum(rd$method == "delta" & rd$coverage < 0.94)` supported scenario-contrast
(lowest `r sprintf("%.3f", min(rd$coverage[rd$method == "delta"]))`), and its
Type I error exceeded 0.060 in
`r sum(rd$method == "delta" & rd$null & rd$rejection > 0.06)`
(highest `r sprintf("%.3f", max(rd$rejection[rd$method == "delta" & rd$null]))`).

## Risk ratios and odds ratios

```{r ratios}
ra <- s[s$scale != "difference", ]
tab <- aggregate(cbind(cells = 1, pass = all_ok, over = coverage > 0.96,
                       low_t1 = null & rejection < 0.04) ~ method + scale + n + control_risk, ra, sum)
tab <- tab[order(tab$scale, tab$method, tab$control_risk, tab$n), ]
knitr::kable(data.frame(
  Scale = scale_label[tab$scale], Method = method_label[tab$method],
  `Per cell` = tab$n, `Control risk` = tab$control_risk,
  `Meeting every tolerance` = sprintf("%d of %d", tab$pass, tab$cells),
  `Coverage above 0.96` = tab$over,
  `Type I error below 0.04` = tab$low_t1,
  check.names = FALSE
), row.names = FALSE, align = "llrrrrr")
```

On the ratio scales every method was conservative in small samples: the
delta-method intervals covered more than 96% of the time in many scenarios
(up to `r sprintf("%.3f", max(ra$coverage[ra$method == "delta"]))`), and
tests rejected true nulls less often than 4%. The log-scale estimators also
showed small-sample bias. Performance approached the tolerances with 100
participants per cell and a control risk of 0.3. The bootstrap percentile
intervals covered closer to nominal on these scales (from
`r rng(ra$coverage[ra$method == "bootstrap"])`), but its p-values, which use
the bootstrap standard error, were conservative, because that standard error
exceeded the empirical one by up to
`r sprintf("%.0f%%", max(ra$relerr[ra$method == "bootstrap"]))`.

## The conditional logistic contrast

`fit_solomon_glm(family = binomial())` with a pretest covariate compares a
treatment effect conditional on the pretest (pretested participants) with a
marginal one (unpretested participants). Odds ratios are noncollapsible, so
the two differ whenever the pretest predicts the outcome (Daniel et al.,
2021). The table shows its Pretest x Treatment contrast when there is no
sensitization.

```{r conditional}
cl <- p[p$method == "conditional_logit" & p$contrast == "Pretest x Treatment" &
          p$pattern %in% c("treatment", "pretest_and_treatment") & p$fit_failures <= 0.05 * reps, ]
cl <- cl[order(cl$bX, cl$pattern, cl$control_risk, cl$n), ]
knitr::kable(data.frame(
  `Pretest log-odds` = cl$bX, Pattern = ifelse(cl$pattern == "treatment", "Treatment only",
                                               "Treatment and pretest"),
  `Per cell` = cl$n, `Control risk` = cl$control_risk,
  `Mean estimate (log odds)` = sprintf("%.3f", cl$mean_estimate),
  `In MCSE units` = sprintf("%.1f", cl$mean_estimate / (cl$empse / sqrt(cl$n_used))),
  `Rejection rate` = sprintf("%.3f", cl$rejection),
  check.names = FALSE
), row.names = FALSE, align = "rlrrrrr")
```

When the pretest predicts the outcome, the contrast averaged 0.08 to 0.21 on
the log-odds scale with no sensitization present, 5 to 13 Monte Carlo
standard errors from zero. At these sample sizes the bias is small relative
to the contrast's standard error, so its test still rejected at close to the
nominal rate (at most `r sprintf("%.3f", max(cl$rejection[cl$bX == 1]))`);
the expectation that it would reject clearly more often than 5% was not borne
out here, although the bias does not shrink as samples grow. As decided in
the protocol, `fit_solomon_glm()` now warns about such fits
(`solomonR_noncollapsible_warning`) and points to `marginal_solomon()`.

## The historical categorical rule

`fisher_solomon()` reproduces the rule described by El Karkri et al. (2025b):
sensitization is declared when the treatment comparison is significant among
pretested participants and not among unpretested participants.

```{r rule}
hr <- p[p$method == "historical_fisher_rule", ]
tab <- aggregate(rejection ~ pattern + n, hr, function(x) rng(x))
tab$pattern <- factor(tab$pattern, c("null", "treatment", "pretest_and_treatment", "sensitization"),
                      c("No effects", "Treatment only (no sensitization)",
                        "Treatment and pretest (no sensitization on the log-odds scale)",
                        "Sensitization"))
tab <- tab[order(tab$pattern, tab$n), ]
knitr::kable(data.frame(Pattern = tab$pattern, `Per cell` = tab$n,
                        `Declares sensitization (range)` = tab$rejection, check.names = FALSE),
             row.names = FALSE, align = "lrr")
```

With a treatment effect and no sensitization, the rule declared sensitization
in up to a quarter of datasets, because it fires whenever the pretested
comparison reaches significance and the unpretested one does not. As Gelman
and Stern (2006) put it, "even large changes in significance levels can
correspond to small, nonsignificant changes in the underlying quantities"
(p. 328).

## Adjusting for the pretest

```{r efficiency}
e <- merge(m[m$method == "delta" & m$scale == "difference", c("scenario", "contrast", "empse", "supported")],
           m[m$method == "unadjusted" & m$scale == "difference", c("scenario", "contrast", "empse")],
           by = c("scenario", "contrast"), suffixes = c("_adj", "_un"))
e <- merge(e, sc[, c("scenario", "bX")], by = "scenario")
e <- e[e$supported, ]
e$ratio <- e$empse_adj / e$empse_un
tab <- aggregate(ratio ~ bX + contrast, e, function(x) rng(x))
knitr::kable(data.frame(`Pretest log-odds` = tab$bX, Contrast = tab$contrast,
                        `Empirical SE, adjusted / unadjusted` = tab$ratio, check.names = FALSE),
             row.names = FALSE, align = "rlr")
```

Standardizing over the pretest made the contrasts involving pretested
participants more precise when the pretest predicts the outcome, with no
meaningful loss when it does not, as Daniel et al. (2021) describe for
covariate adjustment in randomized trials.

## Summary

- Risk differences from `marginal_solomon()` are validated across the
  supported range; the delta method met every tolerance more often than the
  bootstrap, whose intervals were slightly narrow in some scenarios.
- Risk ratios and odds ratios are conservative with 20 to 50 participants per
  cell and approach the tolerances with 100 per cell; their delta-method
  intervals are wider than necessary, not too narrow.
- With 20 participants per cell and a control risk of 0.1, no method is
  supported: too many datasets have a cell without events.
- The logit-scale Pretest x Treatment contrast from a pretest-adjusted
  logistic model is biased away from zero without sensitization.
- The historical rule declares sensitization far more often than 5% when
  there is none.

## Reproducibility

The script, `binary-validation/binary-simulation.R`, and its results,
`performance.csv` and `run-information.csv`, are in the package repository.
The run took from `r run$started` to `r run$finished` on
`r run$workers` workers (`r run$R_version`).

## References

Daniel, R., Zhang, J., & Farewell, D. (2021). Making apples from oranges:
Comparing noncollapsible effect estimators and their standard errors after
adjustment for different covariate sets. *Biometrical Journal, 63*(3),
528–557. https://doi.org/10.1002/bimj.201900297

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025b). Methodological
aspects of the Solomon four-group design: Detecting pre-test sensitisation
and analysing qualitative and quantitative variables in education research.
*Review of Education, 13*(1), Article e70050.
https://doi.org/10.1002/rev3.70050

Gelman, A., & Stern, H. (2006). The difference between "significant" and
"not significant" is not itself statistically significant. *The American
Statistician, 60*(4), 328–331. https://doi.org/10.1198/000313006X152649

Localio, A. R., Margolis, D. J., & Berlin, J. A. (2007). Relative risks and
confidence intervals were easily computed indirectly from multivariable
logistic regression. *Journal of Clinical Epidemiology, 60*(9), 874–882.
https://doi.org/10.1016/j.jclinepi.2006.12.001

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086
