---
title: "How to Cite solomonR and the Methods It Implements"
description: >
  Cite the package and the published methods behind each analysis.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

When you use solomonR, cite both the package and the published methods
behind the analysis you ran. The package is software, and the methods are
the scholarship its analyses rest on.

## The package

```{r citation}
citation("solomonR")
```

Include the version you used. It is printed above, and `packageVersion("solomonR")`
reports it as well.

## The methods

The simplest way to cite the methods of a particular analysis is
`report_solomon()`. It lists the APA 7 references for exactly the methods
and options your fit used. For example, cluster-robust standard errors add
their sources, and a historical analysis names the version of the test
sequence:

```{r report-refs}
fit <- with(solomon_example, fit_solomon_glm(y_post, treat, pretested, y_pre))
report_solomon(fit)$references
```

The table lists the references each analysis function rests on, whatever
options are chosen. Options add references of their own; `report_solomon()`
includes them, and each function's help page lists them all.

```{r registry, echo = FALSE}
refs <- solomonR:::.solomon_function_refs
text <- solomonR:::.solomon_reference_text
cite <- function(keys) {
  if (!length(keys)) return("Its model and inference options (see its help page)")
  short <- vapply(text[keys], function(e) {
    authors <- sub(" \\([0-9]{4}.*$", "", e)
    year <- sub("^[^(]*\\(([0-9]{4}[a-z]?).*$", "\\1", e)
    original <- regmatches(e, regexpr("Original work published [0-9]{4}", e))
    if (length(original)) year <- paste0(sub("^.* ", "", original), "/", year)
    surnames <- sub(",.*$", "", strsplit(authors, "\\.,\\s(&\\s)?", perl = TRUE)[[1]])
    who <- if (length(surnames) == 1L) surnames else if (length(surnames) == 2L) {
      paste(surnames, collapse = " & ")
    } else {
      paste(surnames[1], "et al.")
    }
    paste0(who, " (", year, ")")
  }, "")
  paste(short, collapse = "; ")
}
knitr::kable(data.frame(
  Function = sprintf("`%s()`", names(refs)),
  `Core references` = vapply(refs, cite, ""),
  check.names = FALSE
), row.names = FALSE, align = "ll")
```

The full entries, with notes on how the package uses each work, are on the
[References](references.html) page.

## Historical procedures

If you reproduce a historical analysis, cite its original source. When the
package's replication bears on your use of it, cite that too. The historical
procedures are labeled by author and year throughout the package:

- the Tests A–I sequence and its versions (Walton Braver & Braver, 1988;
  Braver & Walton Braver, 1990);
- the 1995 revision (Walton Braver & Braver, 1995, as cited in Sawilowsky,
  1996).

The article "Historical Tests: Replicating the Published Error Rates" reports
what the package's replication found about their error rates.

## References

Braver, S. L., & Walton Braver, M. C. (1990). Meta-analysis for Solomon
four-group designs reconsidered: A reply to Sawilowsky and Markman.
*Perceptual and Motor Skills, 71*(1), 321–322.
https://doi.org/10.2466/pms.1990.71.1.321

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
https://digitalcommons.wayne.edu/coe_tbf/29/

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150
