---
title: "Historical Tests: Replicating the Published Error Rates"
description: >
  Replication of the published Type I error rates of the Walton Braver and
  Braver sequence of Tests A-I, its 1995 revision, and Sawilowsky's alpha
  allocations, with fit_solomon_classic().
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("classic-validation/performance.csv", stringsAsFactors = FALSE)
pub <- utils::read.csv("classic-validation/published.csv", stringsAsFactors = FALSE)
inv <- utils::read.csv("classic-validation/test-i-investigation.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("classic-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)
p$r <- round(p$r, 2)
pub$r <- round(pub$r, 2)

# Agreement rule fixed in the protocol on issue #51.
m <- merge(pub, p, by = c("flow", "allocation", "dist", "n", "r", "measure"))
m$used <- m$replications - m$failures
m$se <- sqrt(m$rate * (1 - m$rate) / m$used + m$value * (1 - m$value) / 10000)
m$z <- (m$rate - m$value) / m$se
m$agree <- abs(m$z) <= 2
one <- m[m$criterion == "one_tailed", ]
two <- m[m$criterion == "two_tailed", ]

dist_label <- c(normal = "Normal", uniform = "Uniform", exponential = "Exponential",
                chisq1 = "Chi-square (1)", gamma3 = "Gamma (3)", weibull2 = "Weibull (2)",
                cauchy = "Cauchy", t3 = "t (3)")
num <- function(x, d = 4) formatC(x, format = "f", digits = d)
count <- function(ok) sprintf("%d of %d", sum(ok), length(ok))
rate_of <- function(flow, measure, crit = "one_tailed", dist = "normal", n = 30, r = 0, alloc = "none") {
  p$rate[p$flow == flow & p$measure == measure & p$criterion == crit & p$dist == dist &
           p$n == n & p$r == r & p$allocation == alloc]
}
```

The historical analysis of the Solomon design is a sequence of tests, Tests
A to I, proposed by Walton Braver and Braver (1988). Two Monte Carlo studies
estimated how often the sequence declares an effect when there is none
(Sawilowsky et al., 1994; Sawilowsky, 1996). This article reports solomonR's
replication of those published rates with `fit_solomon_classic()`. The
protocol was posted on
[issue #51](https://github.com/JUhalt/solomonR/issues/51) before any run,
following the ADEMP structure of Morris et al. (2019). It covers the 1988
sequence, the 1995 revision without Test D, and Sawilowsky's (1996) alpha
allocations. The results come from package commit
`r substr(run$commit, 1, 7)` and `r format(reps, big.mark = ",")`
replications per condition.

## Design

**Data.** All four groups have the same distribution, so there is no
treatment effect and no pretest effect. The published studies are
replicated in three sets of conditions:

- **Set 1.** Eight distributions with 30 per group and independent pretest
  and posttest: Sawilowsky et al. (1994, Table 2) and Sawilowsky (1996,
  Tables 1 and 4).
- **Set 2.** Normal data with pretest–posttest correlations from .05 to .95
  (Sawilowsky, 1996, Table 3).
- **Set 3.** Normal data with 3, 10, or 20 per group (Sawilowsky, 1996,
  Table 3).

**Methods.** Each dataset was analyzed once with `fit_solomon_classic()`.
The paths of every flow and allocation were derived from its p-values. For
the first 200 datasets of each condition, the fitted function was also run
with each flow and allocation, and its path had to match; all
`r format(sum(unique(p[, c("condition", "paths_checked")])$paths_checked), big.mark = ",")`
checks matched. Test I, the Stouffer combination, was judged by two
criteria:

- the one-tailed p of the combined z, which the package used when the
  study ran;
- the two-tailed p that Walton Braver and Braver report in their worked
  example (1988, p. 153: z = 2.05, p = .040), which the package uses now
  (see the decisions below).

**Agreement.** A replicated rate agrees with a published one if they differ
by at most two standard errors of the difference. That standard error
allows for the Monte Carlo error of both studies; the published studies
used 10,000 replications. A published table is reproduced if at least 90%
of its values agree and none differs by more than four standard errors.

## Tests A to H agree with the published rates

For normal data, the rate at which each classical test is reached and
rejects agrees with the published values:

```{r abh}
ah <- one[one$dist == "normal" & one$n == 30 & one$r == 0 & one$allocation == "none" &
            one$measure %in% c("A", "D", "E", "H"), ]
ah <- ah[order(ah$source, ah$flow, match(ah$measure, c("A", "D", "E", "H"))), ]
knitr::kable(data.frame(
  Source = paste(ah$source, ah$table), Flow = ah$flow, Test = ah$measure,
  Published = num(ah$value), Replicated = num(ah$rate), `Difference (SE)` = sprintf("%.1f", ah$z),
  check.names = FALSE
), row.names = FALSE, align = "llcrrr",
caption = "Rate at which each test is reached and rejects, normal data, 30 per group.")
```

```{r abh-dist}
ah_all <- one[one$table == "Table 1" & one$measure %in% c("A", "D", "E", "H"), ]
ah_miss <- ah_all[!ah_all$agree, ]
```

Across the eight distributions, Tests A to H agreed in
`r count(ah_all$agree)` comparisons. Of the `r nrow(ah_miss)`
disagreements, `r sum(ah_miss$dist %in% c("uniform", "gamma3"))` were in
the uniform and gamma conditions. The published Test A rates for uniform
(`r num(pub$value[pub$dist == "uniform" & pub$measure == "A" & pub$flow == "1988"])`)
and gamma
(`r num(pub$value[pub$dist == "gamma3" & pub$measure == "A" & pub$flow == "1988"])`)
data could not be reproduced. The replicated rates for those distributions,
`r num(rate_of("1988", "A", dist = "uniform"))` and
`r num(rate_of("1988", "A", dist = "gamma3"))`, are close to the nominal .05
and to the rates for the other distributions, and no explanation for the
published values was found. Because the later tests are reached only when
Test A is nonsignificant, their published rates in those two conditions
differ as well.

## Test I does not agree, and why

The published conditional rates of Test I are far lower than the
replication's under either criterion:

```{r testi}
ti <- m[m$measure == "I" & m$dist == "normal" & m$n == 30 & m$r == 0, ]
ti <- ti[order(ti$source, ti$flow, ti$allocation, ti$criterion), ]
w <- stats::reshape(ti[, c("source", "table", "flow", "allocation", "value", "criterion", "rate")],
                    idvar = c("source", "table", "flow", "allocation", "value"),
                    timevar = "criterion", direction = "wide")
knitr::kable(data.frame(
  Source = paste(w$source, w$table), Flow = w$flow, Allocation = w$allocation,
  Published = num(w$value), `One-tailed` = num(w$rate.one_tailed),
  `Two-tailed` = num(w$rate.two_tailed), check.names = FALSE
), row.names = FALSE, align = "lllrrr",
caption = "Rate at which Test I is reached and rejects, normal data, 30 per group.")
```

Test I agreed with the published value in
`r count(one$agree[one$measure == "I"])` comparisons under the one-tailed
criterion and `r count(two$agree[two$measure == "I"])` under the two-tailed
criterion. Under the protocol's decision rule 2, the published rates
support neither criterion, so the rule does not decide it (see the
decisions below).

**Investigation (post hoc).** The disagreement was investigated after the
run. The investigation was not part of the protocol. A separate
implementation of the sequence, written without solomonR, reproduced the
package's Test I rates. It also reproduced the published ones, but only
under a different reading of the Stouffer step:

- **Walton Braver and Braver (1988, p. 152).** Their Stouffer step combines
  the z values of the *one-tailed* p-values of Tests E and H, in the
  direction of the hypothesized effect.
- **The reading that fits.** Each test's *two-sided* p-value is converted to
  z as if it were one-tailed, so the sign of the effect is ignored, and the
  sum is compared with the one-tailed critical value.

```{r investigation}
iv <- inv[inv$measure %in% c("I", "any"), ]
iv$key <- paste(iv$flow, iv$allocation)
pubi <- pub[pub$dist == "normal" & pub$n == 30 & pub$r == 0 & pub$measure %in% c("I", "any") &
              (pub$table != "Table 2"), ]
pubi$key <- paste(pubi$flow, pubi$allocation)
w <- stats::reshape(iv[, c("key", "measure", "reading", "rate")], idvar = c("key", "measure"),
                    timevar = "reading", direction = "wide")
w <- merge(w, pubi[, c("key", "measure", "value")], by = c("key", "measure"))
w <- w[order(w$measure == "any", w$key), ]
knitr::kable(data.frame(
  `Flow and allocation` = w$key, Measure = ifelse(w$measure == "I", "Test I", "Any rejection"),
  Published = num(w$value),
  `Directional, one-tailed` = num(w$rate.directional_one_tailed),
  `Directional, two-tailed` = num(w$rate.directional_two_tailed),
  `Two-sided p read as one-tailed` = num(w$rate.two_sided_p_as_one_tailed),
  check.names = FALSE
), row.names = FALSE, align = "llrrrr",
caption = sprintf("Independent implementation, normal data, 30 per group, %s replications.",
                  format(inv$replications[1], big.mark = ",")))
```

That reading reproduces every published normal-data rate of Test I, and
the experiment-wise rates, within Monte Carlo error. This suggests that the
published studies computed Test I differently from its definition. The
suggestion is an inference from the numbers: the published reports do not
describe their computation in that detail.

## The error rate of the sequence as defined

For the procedure as Walton Braver and Braver define it, the error rates
are higher than those published:

| Sequence | Published | Replicated, one-tailed Test I | Replicated, two-tailed Test I |
|---|---|---|---|
| 1988 (with Test D) | `r num(pub$value[pub$source == "Sawilowsky (1996)" & pub$flow == "1988" & pub$dist == "normal" & pub$measure == "any"])` | `r num(rate_of("1988", "any"))` | `r num(rate_of("1988", "any", "two_tailed"))` |
| 1995 (without Test D) | `r num(pub$value[pub$flow == "1995" & pub$table == "Table 1" & pub$dist == "normal" & pub$measure == "any"])` | `r num(rate_of("1995", "any"))` | `r num(rate_of("1995", "any", "two_tailed"))` |

Both sequences falsely declare an effect nearly three times as often as the
nominal 5%. The conclusion of Sawilowsky et al. (1994) stands: the sequence
is not a safe analysis.

Across pretest–posttest correlations and group sizes (Sets 2 and 3), the
1995 sequence's error rate ranged from
`r num(min(p$rate[p$set %in% c(2, 3) & p$flow == "1995" & p$allocation == "none" & p$measure == "any" & p$criterion == "two_tailed"]))`
to
`r num(max(p$rate[p$set %in% c(2, 3) & p$flow == "1995" & p$allocation == "none" & p$measure == "any" & p$criterion == "two_tailed"]))`
with the two-tailed criterion. The published values from Sawilowsky's (1996)
Table 3 agreed in
`r count(one$agree[one$table == "Table 3"])` comparisons. Those values are
totals that include Test I.

## Sawilowsky's alpha allocations

Sawilowsky (1996, Table 4) chose test-wise levels so that the
experiment-wise error rate of the 1995 sequence stays within the
conservative (.055) or liberal (.075) robustness limit of Bradley (1968, as
cited in Sawilowsky, 1996):

```{r allocations}
al <- c("method1_conservative", "method1_liberal", "method2_conservative", "method2_liberal")
limit <- c(0.055, 0.075, 0.055, 0.075)
tab <- data.frame(
  Allocation = al, Limit = num(limit, 3),
  Published = num(sapply(al, function(a) pub$value[pub$allocation == a & pub$measure == "any"])),
  `Replicated, one-tailed` = num(sapply(al, function(a) rate_of("1995", "any", alloc = a))),
  `Replicated, two-tailed` = num(sapply(al, function(a) rate_of("1995", "any", "two_tailed", alloc = a))),
  check.names = FALSE
)
knitr::kable(tab, row.names = FALSE, align = "lrrrr",
             caption = "Experiment-wise error rate under each allocation, normal data, 30 per group.")
exceed <- sum(sapply(seq_along(al), function(i) rate_of("1995", "any", "two_tailed", alloc = al[i]) > limit[i]))
```

The allocations reproduce the published rates only under the reading of
Test I described above. With Test I as defined, `r exceed` of the 4
allocations exceed their limit under the two-tailed criterion. As the
protocol's decision rule 3 requires, `fit_solomon_classic()` implements
Sawilowsky's published levels, and its help page reports these replicated
rates next to his.

## The 1990 amendment

The 1990 flow has no published error rates. With Test I regarded as
definitive, a significant Test A, Test D, or Test I declared an effect in
`r sprintf("%.1f%%", 100 * rate_of("1990", "decisive"))` of datasets
(one-tailed Test I). Any significant test on the path did so in
`r sprintf("%.1f%%", 100 * rate_of("1990", "any"))`.

## Decisions under the protocol

1. **Flows.** The published tables were not reproduced as a whole:
   `r count(one$agree)` published values agreed under the one-tailed
   criterion and `r count(two$agree)` under the two-tailed criterion. By
   rule 1, the flows are not described as validated against the published
   tables. The investigation locates the disagreement:
   - Tests A to H and the flow logic agree for normal data, and the fitted
     function's paths matched in every check.
   - The published Test I rates follow a computation that differs from
     Walton Braver and Braver's definition.
   - The published Test A rates for uniform and gamma data, and the rates
     of the tests that follow them, remain unexplained.
2. **Test I criterion.** Neither criterion reproduces the published Test I
   rates, so rule 2 did not decide between them, and the question went to
   the maintainer. The maintainer's decision was to follow the literature.
   Walton Braver and Braver's worked example reports the two-tailed p of the
   combined z (1988, p. 153), so `fit_solomon_classic()` and
   `power_solomon()` now judge Test I by it. The one-tailed value is still
   returned. The replicated error rates above are given under both
   criteria.
3. **Allocations.** The option implements the published levels. The help
   page reports the replicated rates and that `r exceed` of the 4 exceed
   their limits with Test I as defined.

## Reproducibility

The files are in the package repository:

- the pre-registered script, `classic-validation/classic-simulation.R`;
- the published targets, `published.csv`, with table and page;
- the results, `performance.csv` and `run-information.csv`;
- the post hoc investigation, `test-i-investigation.R` and its output.

The run took from `r run$started` to `r run$finished` on `r run$workers`
workers (`r run$R_version`).

## References

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
http://digitalcommons.wayne.edu/coe_tbf/29

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S. (1994).
Meta-analysis and the Solomon four-group design. *The Journal of
Experimental Education, 62*(4), 361–376.
https://doi.org/10.1080/00220973.1994.9944140

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150
