---
title: "Clustered Designs: Validating Cluster-Level Randomization Inference"
description: >
  Simulation study of perm_solomon() for cluster-randomized Solomon designs,
  with raw and studentized cluster-level statistics, compared with CR2 tests
  from fit_solomon_glm().
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("cluster-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("cluster-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

method_label <- c(
  perm_studentized = "perm_solomon(), studentized",
  perm_difference = "perm_solomon(), difference",
  cr2 = "CR2, Satterthwaite t"
)
design_label <- c(A = "Clusters assigned to the four conditions",
                  B = "Treatment by cluster, pretesting within clusters")
allocation_label <- c(
  balanced_small = "Balanced, small", balanced = "Balanced",
  unbalanced_small = "Unbalanced, small", unbalanced = "Unbalanced, Kvalem-like"
)
clusters_label <- list(
  A = c(balanced_small = "4, 4, 4, 4", balanced = "10, 10, 10, 10",
        unbalanced_small = "4, 4, 8, 8", unbalanced = "15, 15, 47, 47"),
  B = c(balanced_small = "4, 4", balanced = "10, 10",
        unbalanced_small = "4, 8", unbalanced = "15, 47")
)

p$balanced <- p$allocation %in% c("balanced_small", "balanced")
# Tolerances declared on issue #19 before any run.
p$anticonservative <- p$rejection > 0.060
p$conservative <- p$rejection < 0.040
p$within <- !p$anticonservative & !p$conservative
nulls <- p[p$null_contrast, ]

rng <- function(x, d = 3) if (length(x)) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x)) else "--"
```

This article reports the simulation validation of `perm_solomon()` for
cluster-randomized Solomon designs. The aims, data-generating mechanisms,
estimands, methods, performance measures, tolerances, and three decision
rules were posted on
[issue #19](https://github.com/JUhalt/solomonR/issues/19) before the method
was implemented and before any run, following the ADEMP structure of Morris
et al. (2019). An amendment posted before the run corrected which contrasts
are enumerated exactly. The results come from `r reps` replications per
scenario.

## The method

When whole clusters were randomized, `perm_solomon()` permutes the
treatment labels of clusters, never of participants. It supports two
assignment mechanisms:

- **Design A:** whole clusters assigned to the four Solomon conditions, as in
  Kvalem et al. (1996). Treatment labels are permuted among clusters within
  each pretest condition.
- **Design B:** treatment assigned to clusters and pretesting to participants
  within clusters. Treatment labels are permuted among all clusters.

The test statistic comes from cluster-level summaries (Gail et al., 1996;
Hayes & Moulton, 2017, ch. 10):

1. A Stage 1 model with every term of the fit except treatment gives each
   cluster a covariate-adjusted difference residual (Hayes & Moulton, 2017,
   pp. 221–224, following Bennett et al., 2002).
2. The contrast compares unweighted means of these residuals, so each
   cluster counts once.
3. The **difference** statistic is the contrast itself, the statistic of
   Gail et al. (1996, p. 1076). The **studentized** statistic divides it by
   its separate-variances standard error (Hayes & Moulton, 2017, p. 212),
   the studentization Wu and Ding (2021) use for weak null hypotheses,
   applied here with clusters as the units.

When the number of possible allocations is at most the number of
permutations requested, every allocation is used and the p-value is exact.

## Design

```{r allocations}
alloc <- unique(p[, c("design", "allocation")])
alloc <- alloc[order(alloc$design, match(alloc$allocation, names(allocation_label))), ]
knitr::kable(data.frame(
  Design = design_label[alloc$design],
  Allocation = allocation_label[alloc$allocation],
  Clusters = mapply(function(d, a) clusters_label[[d]][[a]], alloc$design, alloc$allocation),
  row.names = NULL
), caption = paste(
  "Clusters per cell: treated-pretested, treated-unpretested, control-pretested,",
  "control-unpretested (Design A); treated, control (Design B)."
))
```

**Data.** Cluster sizes are uniform from 10 to 30. In Design B, half of each
cluster's participants (rounded down) are pretested.

- *Continuous outcomes:* Y = 0.5X + effect + u + w + ε, where X ~ N(0, 1) is
  the baseline, observed as the pretest score in the pretested groups;
  ε ~ N(0, 0.75); u ~ N(0, 0.10 r) is a cluster effect; and w ~ N(0, 0.05 r)
  is a cluster-by-pretest effect. Analyses adjust for the pretest score.
- *Binary outcomes:* each cluster's risk is drawn from a beta distribution
  with mean 0.30 and variance 0.01 r, with no pretest score.

The variance multiplier r is 1 for control clusters and ψ for treated
clusters, with ψ = 1/4, 1, or 4, following the variance ratios of Gail et
al. (1996, Table I). When ψ = 1 and there is no effect, the strong null
hypothesis holds. When ψ ≠ 1, treatment changes the between-cluster variance
but not the average, so only the weak null holds.

**Effect patterns.** *Null:* no average treatment effect. *Treatment:* an
average effect of 0.3 (continuous) or 0.10 in risk (binary) in both pretest
conditions, with no Pretest x Treatment interaction. The full factorial has
`r length(unique(p$scenario))` scenarios.

**Methods.** `perm_solomon()` with each statistic, using 999 permutations or
every allocation when there are at most 999, and the CR2 tests with
Satterthwaite degrees of freedom from `fit_solomon_glm(robust = "CR2")`.

**Tolerances.** Type I error, the rejection rate at α = .05 for a contrast
whose true value is zero, should lie from 0.040 to 0.060: above 0.060 is
anticonservative and below 0.040 conservative. The Monte Carlo standard error
of a rejection rate near 0.05 is about 0.005.

## Type I error

```{r type1-summary}
tab <- aggregate(cbind(cells = 1, within = within, anti = anticonservative, cons = conservative) ~
                   method + balanced, nulls, sum)
tab <- tab[order(tab$balanced, match(tab$method, names(method_label))), ]
knitr::kable(data.frame(
  Allocation = ifelse(tab$balanced, "balanced", "unbalanced"),
  Method = method_label[tab$method],
  `Within 0.040 to 0.060` = sprintf("%d of %d", tab$within, tab$cells),
  `Above 0.060` = tab$anti,
  `Below 0.040` = tab$cons,
  check.names = FALSE, row.names = NULL
), caption = "Null scenario-contrasts, across both designs, both outcomes, and all variance ratios.")
```

```{r type1-by-design}
tab <- aggregate(rejection ~ method + design + allocation, nulls, function(x) rng(x))
tab <- tab[order(tab$design, match(tab$allocation, names(allocation_label)),
                 match(tab$method, names(method_label))), ]
knitr::kable(data.frame(
  Design = tab$design,
  Allocation = allocation_label[tab$allocation],
  Method = method_label[tab$method],
  `Type I error (range)` = tab$rejection,
  check.names = FALSE, row.names = NULL
))
```

### The sharp null hypothesis

```{r sharp}
sharp <- nulls[nulls$psi == 1 & nulls$pattern == "null", ]
perm_sharp <- sharp[sharp$method != "cr2", ]
tab <- aggregate(rejection ~ method + allocation, sharp, max)
wide <- stats::reshape(tab, idvar = "allocation", timevar = "method", direction = "wide")
wide <- wide[match(names(allocation_label), wide$allocation), ]
knitr::kable(data.frame(
  Allocation = allocation_label[wide$allocation],
  Studentized = sprintf("%.3f", wide$rejection.perm_studentized),
  Difference = sprintf("%.3f", wide$rejection.perm_difference),
  CR2 = sprintf("%.3f", wide$rejection.cr2),
  check.names = FALSE, row.names = NULL
), caption = "Largest Type I error with no effect in any cluster (variance ratio 1), across designs, outcomes, and contrasts.")
```

With no effect in any cluster, a randomization test is exact by
construction. The largest Type I error of either permutation statistic was
`r sprintf("%.3f", max(perm_sharp$rejection))` across
`r nrow(perm_sharp)` scenario-contrasts, which shows how far Monte Carlo
error alone can carry a rejection rate above 0.05 in a study of this size.
With four clusters per arm, only 70 allocations exist in a pretest
condition, so the attainable two-sided level at .05 is 2/70, about 0.03:
the permutation tests are conservative there by design. CR2 tests, by
contrast, reached `r sprintf("%.3f", max(sharp$rejection[sharp$method == "cr2"]))`
with four clusters per arm.

### Unequal variances with unequal numbers of clusters

When treated clusters are more variable than control clusters, only the
weak null hypothesis holds, and Gail et al. (1996, p. 1079) showed that a
permutation test of the raw difference can then reject too often if the arm
with fewer clusters is the more variable one.

```{r gail-case}
g <- nulls[!nulls$balanced & nulls$psi != 1, ]
tab <- aggregate(rejection ~ method + psi + allocation, g, max)
tab <- tab[order(tab$allocation, tab$psi, match(tab$method, names(method_label))), ]
knitr::kable(data.frame(
  Allocation = allocation_label[tab$allocation],
  `Treated/control variance ratio` = tab$psi,
  Method = method_label[tab$method],
  `Largest Type I error` = sprintf("%.3f", tab$rejection),
  check.names = FALSE, row.names = NULL
), caption = "Unbalanced allocations with treatment changing the between-cluster variance.")
```

```{r gail-numbers}
st4 <- nulls[nulls$method == "perm_studentized" & nulls$psi == 4, ]
df4 <- nulls[nulls$method == "perm_difference" & nulls$psi == 4 & !nulls$balanced, ]
st_small <- max(st4$rejection[st4$allocation == "unbalanced_small"])
st_kvalem <- max(st4$rejection[st4$allocation == "unbalanced"])
st_bal <- max(st4$rejection[st4$balanced])
```

This is what the study found. With treated clusters four times as variable
as control clusters and fewer treated clusters, the difference statistic
reached a Type I error of `r sprintf("%.3f", max(df4$rejection))`. The
studentized statistic reduced the excess but did not remove it: its largest
Type I error was `r sprintf("%.4f", st_small)` with 4 treated and 8 control
clusters, and `r sprintf("%.4f", st_kvalem)` with 15 and 47. With equal
numbers of treated and control clusters its largest was
`r sprintf("%.3f", st_bal)`, close to the Monte Carlo spread seen under the
sharp null. When treated clusters were less variable (ratio 1/4), both
permutation statistics stayed at or below
`r sprintf("%.3f", max(nulls$rejection[nulls$psi == 0.25 & nulls$method != "cr2"]))`.

The Pretest x Treatment test under the treatment pattern, where treatment
has an effect but no interaction, is the case the existing help page called
approximate. There the studentized statistic's Type I error ranged from
`r rng(nulls$rejection[nulls$pattern == "treatment" & nulls$method == "perm_studentized"])`.

## Power

```{r power}
alt <- p[!p$null_contrast, ]
tab <- aggregate(rejection ~ method + outcome + design + allocation, alt, mean)
tab <- tab[order(tab$outcome, tab$design, match(tab$allocation, names(allocation_label)),
                 match(tab$method, names(method_label))), ]
wide <- stats::reshape(tab, idvar = c("outcome", "design", "allocation"), timevar = "method",
                       direction = "wide")
knitr::kable(data.frame(
  Outcome = wide$outcome,
  Design = wide$design,
  Allocation = allocation_label[wide$allocation],
  Studentized = sprintf("%.2f", wide$rejection.perm_studentized),
  Difference = sprintf("%.2f", wide$rejection.perm_difference),
  CR2 = sprintf("%.2f", wide$rejection.cr2),
  check.names = FALSE, row.names = NULL
), caption = "Mean rejection rate for the nonnull contrasts under the treatment pattern.")
```

## Decision rules

```{r rules}
stud <- nulls[nulls$method == "perm_studentized", ]
rule1 <- all(stud$rejection <= 0.060)
diff_gail <- nulls[nulls$method == "perm_difference" & !nulls$balanced & nulls$psi != 1, ]
rule2 <- any(diff_gail$rejection > 0.060)
key <- c("scenario", "contrast")
paired <- merge(nulls[nulls$method == "cr2", c(key, "rejection", "design", "allocation")],
                stud[, c(key, "rejection")], by = key, suffixes = c(".cr2", ".stud"))
rule3 <- paired[paired$rejection.cr2 > 0.060 & paired$rejection.stud <= 0.060, ]
```

The protocol fixed three rules before the run.

1. **Default statistic.** The studentized statistic becomes the default if
   its Type I error is at most 0.060 for every null contrast in every
   scenario; otherwise the allocations where it exceeds 0.060 are named in
   the help page and `perm_solomon()` warns for such designs. Its largest
   Type I error was `r sprintf("%.4f", max(stud$rejection))`, so the rule is
   `r if (rule1) "met" else "not met"`. It exceeded 0.060 in
   `r sum(stud$rejection > 0.060)` of `r nrow(stud)` scenario-contrasts, all
   with treated clusters four times as variable as control clusters. The
   studentized statistic remains the default, because the difference
   statistic did worse in every such case, and the help page names the
   allocations. `perm_solomon()` warns whenever treated and control clusters
   differ in number, where the excess was systematic. With equal numbers the
   excess was at most `r sprintf("%.3f", st_bal)`, close to the
   `r sprintf("%.3f", max(perm_sharp$rejection))` that Monte Carlo error
   produced under the sharp null, where the test is exact; those allocations
   are named in the help page but not warned about. This reading of the rule
   was reported on issue #19 with the results.
2. **Difference statistic.** If it exceeds 0.060 in an unbalanced scenario
   with unequal variances, `perm_solomon()` warns when it is used with unequal
   numbers of treated and control clusters. Its largest Type I error there
   was `r sprintf("%.3f", max(diff_gail$rejection))`, so
   `r if (rule2) "the warning is added" else "no warning is needed"`.
3. **CR2 comparison.** Where CR2 exceeds 0.060 while the studentized
   permutation test stays within tolerance, the guidance recommends
   `perm_solomon()`. This happened in `r nrow(rule3)` scenario-contrasts,
   `r sum(rule3$allocation == "balanced_small")` of them with four clusters
   per arm, so the help pages of `fit_solomon_glm()` and `perm_solomon()`
   recommend the permutation test for designs with few clusters.

## Summary

- `perm_solomon()` gives exact tests of the sharp null hypothesis for
  cluster-randomized Solomon designs of both kinds studied, for continuous
  and binary outcomes. With four clusters per arm it is conservative,
  because so few allocations exist (Hayes & Moulton, 2017, p. 239).
- With equal numbers of treated and control clusters, the studentized
  statistic also held its level approximately under the weak null
  hypothesis: its excess was within the range Monte Carlo error produced
  under the sharp null.
- With unequal numbers and more variable treated clusters, no method held
  its level: the raw difference failed badly, and the studentized statistic
  and CR2 exceeded 0.060. `perm_solomon()` warns in such designs.
- CR2 tests exceeded the nominal level with four clusters per arm even
  when treatment had no effect on any cluster.
- Counts were not simulated. The test is exact for them under the sharp null
  hypothesis, but its behavior under the weak null has not been checked.

## Reproducibility

The script, `cluster-validation/cluster-simulation.R`, and its results,
`performance.csv` and `run-information.csv`, are in the package repository.
The run took from `r run$started` to `r run$finished` on `r run$workers`
workers (`r run$R_version`), at package commit
`r substr(run$commit, 1, 7)`. It was restarted once to schedule one
scenario per task. Two scenarios that had finished before the restart, at
commit 48c926d, were kept; the package code is the same at both commits,
and each scenario sets its own seed, so the results do not depend on the
scheduling.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References
Bennett, S., Parpia, T., Hayes, R., & Cousens, S. (2002). Methods for the
analysis of incidence rates in cluster randomized trials. *International
Journal of Epidemiology, 31*(4), 839–846. https://doi.org/10.1093/ije/31.4.839

Gail, M. H., Mark, S. D., Carroll, R. J., Green, S. B., & Pee, D. (1996). On
design considerations and randomization-based inference for community
intervention trials. *Statistics in Medicine, 15*(11), 1069–1092.
<https://doi.org/10.1002/(SICI)1097-0258(19960615)15:11%3C1069::AID-SIM220%3E3.0.CO;2-Q>

Hayes, R. J., & Moulton, L. H. (2017). *Cluster randomised trials* (2nd ed.).
Chapman and Hall/CRC. https://doi.org/10.4324/9781315370286

Kvalem, I. L., Sundet, J. M., Rivø, K. I., Eilertsen, D. E., & Bakketeig, L.
S. (1996). The effect of sex education on adolescents' use of condoms:
Applying the Solomon four-group design. *Health Education Quarterly, 23*(1),
34–47. https://doi.org/10.1177/109019819602300103

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086

Wu, J., & Ding, P. (2021). Randomization tests for weak null hypotheses in
randomized experiments. *Journal of the American Statistical Association,
116*(536), 1898–1913. https://doi.org/10.1080/01621459.2020.1750415
