---
title: "Count Outcomes: Validating Poisson Fits and marginal_solomon()"
description: >
  Simulation study of the Solomon contrasts for count outcomes, with and
  without overdispersion, from fit_solomon_glm(family = poisson()) and
  marginal_solomon().
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("count-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("count-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

method_label <- c(
  hc3 = "Poisson fit, HC3",
  model_based = "Poisson fit, model-based",
  marginal_delta = "marginal_solomon(), delta method"
)
scale_label <- c(
  log_rate_ratio_link = "Log rate ratio (model contrast)",
  rate_difference = "Rate difference",
  log_rate_ratio = "Log rate ratio (marginal)"
)

p$supported <- p$fit_failures <= 0.05 * reps
p$null <- abs(p$true_value) < 1e-8
# Tolerances declared on issue #44 before any run.
p$bias_ok <- abs(p$bias) <= 2 * p$bias_mcse
p$cov_ok <- p$coverage >= 0.94 & p$coverage <= 0.96
p$se_ok <- abs(p$relerr) <= 10
p$t1_ok <- ifelse(p$null, p$rejection >= 0.04 & p$rejection <= 0.06, TRUE)
p$all_ok <- p$bias_ok & p$cov_ok & p$se_ok & p$t1_ok
s <- p[p$supported, ]

rng <- function(x, d = 3) if (length(x)) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x)) else "--"
```

This article reports the simulation validation of the Solomon contrasts for
count outcomes. The aims, data-generating mechanisms, estimands, methods,
performance measures, tolerances, and a decision rule about a
negative-binomial option were posted on
[issue #44](https://github.com/JUhalt/solomonR/issues/44) before any run,
following the ADEMP structure of Morris et al. (2019). The results come from
package commit `r substr(run$commit, 1, 7)` and `r reps` replications per
scenario.

## Design

**Data.** Each participant has a latent baseline X ~ N(0, 1), observed only
in the pretested groups, an exposure E ~ Uniform(0.5, 1.5), and a count with

log E[Y] = log E + b0 + bX X + bT T + bP P + bTP T P.

Counts are Poisson or overdispersed: negative binomial with
Var(Y) = μ + α μ², generated as Poisson counts with a gamma multiplier. The
96 scenarios cross participants per cell (20, 50, 100), the control rate
(0.5, 3), overdispersion (α = 0 or 1), the association of the pretest with
the outcome (bX = 0 or 0.5), and four effect patterns: no effects, a
treatment effect only (bT = log 1.5), a treatment and a pretest effect
(bP = log 1.25), and sensitization only (bTP = log 1.5).

**Estimands.** The model's contrasts are the conditional log rate ratios.
The marginal rates per unit of exposure have a closed form, from which come
the Solomon contrasts as rate differences and as marginal log rate ratios.
Log-link rate ratios are collapsible (Daniel et al., 2021), so the model's
treatment and sensitization contrasts estimate the marginal ones.

**Methods.** `fit_solomon_glm(family = poisson(), exposure = )` with the
default HC3 covariance and with model-based covariance, and
`marginal_solomon()` with delta-method intervals.

**Tolerances.** Bias within 2 Monte Carlo standard errors (MCSE); coverage of
95% intervals from 0.940 to 0.960; model standard error within 10% of the
empirical standard error; and, where the true value is zero, a rejection rate
from 0.040 to 0.060.

## Failures

```{r failures}
f <- unique(p[, c("scenario", "n", "control_rate", "fit_failures")])
tab <- aggregate(fit_failures ~ n + control_rate, f, max)
knitr::kable(data.frame(`Per cell` = tab$n, `Control rate` = tab$control_rate,
                        `Failed fits (max over scenarios)` = tab$fit_failures,
                        check.names = FALSE), row.names = FALSE)
```

A fit fails when it does not converge or when a Solomon cell has no counts at
all. `r sum(!unique(p[, c("scenario", "supported")])$supported)` of 96
scenarios had more than 5% failed fits.

## Results by method

```{r summary}
tab <- aggregate(cbind(cells = 1, pass = all_ok, cov = cov_ok, t1 = t1_ok) ~ method + scale + alpha, s, sum)
tab <- tab[order(tab$alpha, tab$method, tab$scale), ]
knitr::kable(data.frame(
  Overdispersion = ifelse(tab$alpha == 0, "none (Poisson)", "alpha = 1"),
  Method = method_label[tab$method], Scale = scale_label[tab$scale],
  `Meeting every tolerance` = sprintf("%d of %d", tab$pass, tab$cells),
  `Coverage within tolerance` = sprintf("%d of %d", tab$cov, tab$cells),
  check.names = FALSE
), row.names = FALSE)
```

```{r coverage}
tab <- aggregate(coverage ~ method + scale + alpha + n, s, function(x) rng(x))
tab <- tab[order(tab$alpha, tab$method, tab$scale, tab$n), ]
knitr::kable(data.frame(
  Overdispersion = ifelse(tab$alpha == 0, "none", "alpha = 1"),
  Method = method_label[tab$method], Scale = scale_label[tab$scale],
  `Per cell` = tab$n, `Coverage (range)` = tab$coverage, check.names = FALSE
), row.names = FALSE)
```

## Overdispersion and the decision rule

```{r decision}
d <- s[s$method == "hc3" & s$alpha == 1 & s$n >= 50, ]
share <- mean(d$cov_ok & d$t1_ok)
```

The protocol fixed a rule before the run: if HC3 inference from the Poisson
fit met the coverage and Type I tolerances in at least 90% of the supported
overdispersed scenario-contrasts with 50 or 100 participants per cell, no
negative-binomial option would be added. The share was
`r sprintf("%.0f%%", 100 * share)` (`r sum(d$cov_ok & d$t1_ok)` of
`r nrow(d)`), so `r if (share >= 0.9) "no negative-binomial option is added" else "a negative-binomial option is proposed, with its own protocol"`.

Where HC3 fell short, it was with strong overdispersion and small cells.
With a control rate of 3 (mean Pearson dispersion of about 4 to 6), its
coverage ranged from
`r rng(s$coverage[s$method == "hc3" & s$alpha == 1 & s$control_rate == 3 & s$n == 20])`
with 20 participants per cell and
`r rng(s$coverage[s$method == "hc3" & s$alpha == 1 & s$control_rate == 3 & s$n == 50])`
with 50, and reached
`r rng(s$coverage[s$method == "hc3" & s$alpha == 1 & s$control_rate == 3 & s$n == 100])`
with 100. With a control rate of 0.5, overdispersion was milder and HC3 met
the tolerances in most contrasts at every cell size.

```{r hc3-by-design}
h <- s[s$method == "hc3" & s$alpha == 1, ]
tab <- aggregate(cbind(cells = 1, ok = cov_ok & t1_ok) ~ n + control_rate, h, sum)
tab <- tab[order(tab$control_rate, tab$n), ]
knitr::kable(data.frame(`Control rate` = tab$control_rate, `Per cell` = tab$n,
                        `HC3 contrasts within the coverage and Type I tolerances` =
                          sprintf("%d of %d", tab$ok, tab$cells), check.names = FALSE),
             row.names = FALSE, caption = "Overdispersed counts (alpha = 1).")
```

Model-based standard errors understate uncertainty whenever the counts are
overdispersed: their coverage ranged from
`r rng(s$coverage[s$method == "model_based" & s$alpha == 1])` with
α = 1. This is the failure Cameron and Trivedi (2013, p. 72) describe. They
also failed with Poisson counts when the pretest predicts the outcome
(coverage from
`r rng(s$coverage[s$method == "model_based" & s$alpha == 0 & s$bX > 0])`),
which the protocol did not anticipate. The unpretested participants have no
pretest in the model, so their counts mix over the unobserved baseline and
are overdispersed even when each person's count is Poisson. With no pretest
effect, model-based coverage ranged from
`r rng(s$coverage[s$method == "model_based" & s$alpha == 0 & s$bX == 0])`.
This is the reason the package's default is the robust covariance.

## Summary

- With the default HC3 covariance, Poisson fits give valid inference for the
  Solomon contrasts on the log rate-ratio scale when the counts are Poisson
  or mildly overdispersed.
- With strong overdispersion, HC3 intervals are somewhat too narrow with 20
  to 50 participants per cell and close to nominal with 100. By the rule
  fixed in the protocol, a negative-binomial option is proposed, with its own
  protocol.
- `marginal_solomon()` adds rate differences, with delta-method intervals.
- Model-based standard errors should not be used for counts: they fail under
  overdispersion, and in the unpretested groups whenever the pretest predicts
  the outcome.

## Reproducibility

The script, `count-validation/count-simulation.R`, and its results,
`performance.csv` and `run-information.csv`, are in the package repository.
The run took from `r run$started` to `r run$finished` on `r run$workers`
workers (`r run$R_version`).

## References

Cameron, A. C., & Trivedi, P. K. (2013). *Regression analysis of count data*
(2nd ed.). Cambridge University Press.
https://doi.org/10.1017/CBO9781139013567

Daniel, R., Zhang, J., & Farewell, D. (2021). Making apples from oranges:
Comparing noncollapsible effect estimators and their standard errors after
adjustment for different covariate sets. *Biometrical Journal, 63*(3),
528–557. https://doi.org/10.1002/bimj.201900297

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086
