---
title: "A History of the Solomon Design and Its Analysis"
description: >
  From Solomon (1949) to the MERIT recommendations: how the four-group design
  was proposed, how its analysis was debated, and which historical procedures
  solomonR reproduces.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 4.2)
library(solomonR)
```

Solomon studies published over seven decades used different analyses, and
several of those analyses were later criticized. Reading an older study, or
replicating one, requires knowing which procedure it followed and what was
later learned about it. This article traces that history from the primary
sources. solomonR labels each historical procedure by its author and year,
reproduces it, and keeps it separate from the analyses it recommends today.

```{r timeline, echo = FALSE, fig.alt = "Timeline of the Solomon design's methodological history from 1949 to 2025, in three strands: the design, its analysis, and evidence and recommendations."}
events <- data.frame(
  year = c(1949, 1957, 1959, 1963, 1968, 1973, 1982, 1988, 1990, 1994, 1996, 2011, 2021, 2025),
  label = c("Solomon: three- and\nfour-group designs",
            "Campbell: 2 x 2\nposttest analysis",
            "Lana: first\nattitude experiment",
            "Campbell & Stanley:\nDesign 5",
            "Bracht & Glass:\nexternal validity",
            "Huck & Sandler:\nANCOVA advice",
            "Willson & Putnam:\nmeta-analysis",
            "Walton Braver & Braver:\nTests A-I",
            "Exchange; 1990\namendment",
            "Sawilowsky et al.:\nType I error",
            "Sawilowsky: 1995\nrevision, allocations",
            "McCambridge et al.:\nreview",
            "French et al.:\nMERIT",
            "El Karkri et al."),
  strand = c("Design", "Analysis", "Evidence", "Design", "Evidence", "Analysis", "Evidence",
             "Analysis", "Analysis", "Evidence", "Evidence", "Evidence", "Evidence", "Analysis")
)
events$strand <- factor(events$strand, c("Evidence", "Analysis", "Design"))
# Alternate labels above and below the line within each strand.
events$nudge <- stats::ave(seq_len(nrow(events)), events$strand,
                           FUN = function(i) rep(c(0.28, -0.28), length.out = length(i)))
ggplot2::ggplot(events, ggplot2::aes(year, strand)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = as.numeric(strand)), colour = "grey85") +
  ggplot2::geom_point(size = 2.6, colour = "#2c6e9b") +
  ggplot2::geom_text(ggplot2::aes(y = as.numeric(strand) + nudge, label = label),
                     size = 2.7, lineheight = 0.9) +
  ggplot2::scale_x_continuous(breaks = seq(1950, 2025, 10), limits = c(1943, 2031)) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(),
                 panel.grid.minor = ggplot2::element_blank())
```

## The design (1949)

Solomon (1949) wrote for researchers studying transfer of training, attitude
change, and the effect of experience on existing skills (p. 137). His
concern was that a pretest does more than measure. It can interact with the
training, changing how participants attend to it or receive it (p. 141).

He first proposed a **three-group design**: an experimental group and a
control group, both pretested, plus a second control group that received
the training without a pretest (p. 141). Since that group had no pretest,
he inferred its pretest mean from the pooled pretested groups and computed
an interaction term from the groups' improvement scores (pp. 141–143).

- **The demonstration.** It used spelling lessons in two classes. The groups
  were "roughly equated" by teachers' judgments rather than randomized
  (p. 144).
- **The result.** The pretest reduced the effect of the lesson, so the usual
  two-group design would have underrated it (p. 145).

`fit_solomon_1949()` computes Solomon's analysis, and `solomon1949` holds
his published means. For the fifth grade (Table II, p. 144), the inferred
pretest is 3.0, and the interaction is I = d1 − (d2 + d3) = −2.2:

```{r solomon-1949}
g5 <- solomon1949[solomon1949$grade == 5, ]
fit_solomon_1949(post_mean = g5$mean, pre_mean = g5$pre_mean[1:2], n = g5$n)
```

The sixth grade gives I = −3.1 (Table III, p. 145). Solomon reported no
standard error for I: he printed the error of Control Group II's
improvement as "?" and judged the interaction from the variability of the
observed means (p. 144).

The **fourth group**, which received neither pretest nor training, came
from field studies. There, events between pretest and posttest affect
everyone, and a group with no pretest and no treatment shows their effect.
In a footnote, Solomon credits Stouffer with pointing out that this group
was necessary (p. 146). That group's change was attributed to the outside
event (p. 147), and Solomon added that time and maturation can change scores
in the same way (p. 148).

With the fourth group, the interaction becomes I = d1 − (d2 + d3 − d4)
(Table V, p. 147). Both unpretested groups get the same inferred pretest,
so it cancels. What remains is the two-by-two interaction contrast of the
posttests, less the pretest difference between the two pretested groups
(`?fit_solomon_1949`).

## The analysis recast (1957)

Campbell (1957) brought the design into the discussion of validity. He
described the interaction of testing and treatment as a sensitization of
participants to the treatment, which leaves a pretest-posttest experiment
no basis for generalizing to people who were never pretested (p. 302). The
Solomon design, which controls and measures that interaction, had become
in his words "the new ideal design for social scientists" (p. 303).

He also set the terms of the analysis that later texts repeated (p. 303):

- **Against the inferred pretest.** Giving the unpretested groups the
  pretested groups' mean pretest "restricts the effective degrees of
  freedom, violates assumptions of independence," and leaves no legitimate
  test of the effects.
- **For the posttests.** The four posttests form a two-by-two analysis of
  variance: pretesting, treatment, and their interaction.
- **For history and maturation.** A t test of the unpretested control
  posttest against the pretests.

## The first experiments (1959–1961)

The first studies built on the design asked whether pretesting mattered at
all.

**Lana (1959)** gave a recorded talk on vivisection to introductory
psychology classes, in a four-group design with a pretest 12 days before
the talk. The talk changed attitudes, but the pretest neither changed them
nor interacted with the talk (pp. 297–298). His analysis of variance of the
posttest means can be reproduced from the published cell statistics,
bundled as `lana1959`:

```{r lana}
with(lana1959, solomon_from_summary(n, mean, sd))
```

The treatment F is 5.36, against the published 5.35 (Table 3, p. 297).
Lana's sums of squares are on the scale of the cell means; divided by the
harmonic mean of the cell sizes, the package's sums of squares match them.

**Entwisle (1961)** paired fourth-graders on sex and IQ and pretested one
member of each pair at random. There was no overall pretest effect, but the
pretest helped some combinations of sex and ability and hindered others
(pp. 612–613). She suggested that Solomon's result was such an effect,
observed in a homogeneous group (p. 614), and that pretest effects may fade
as the interval grows (p. 610).

## The design in the canon (1963)

Campbell and Stanley's survey, reprinted as a book in 1966, made the
four-group design **Design 5** of its three true experimental designs. They
wrote that it "deservedly has higher prestige" and was the first design to
consider external validity explicitly (Campbell & Stanley, 1963/1966,
p. 24). Page numbers here refer to the 1966 book.

- **What the design adds** (p. 25):
  - it determines both the main effect of testing and the interaction of
    testing with the treatment;
  - it replicates the treatment effect in four comparisons;
  - comparing the unpretested control posttest with the pretests estimates
    the combined effect of maturation and history.
- **The analysis problem** (p. 25):
  - no single statistical procedure uses all six sets of observations at
    once;
  - the design's asymmetry rules out an analysis of variance of gain
    scores, and they judged Solomon's gain-score suggestions unacceptable.
- **Their recommended analysis** (p. 25):
  - a two-by-two analysis of variance of the posttest scores, as Campbell
    (1957, p. 303) had proposed;
  - then, if the effects of pretesting are negligible, an analysis of
    covariance of the pretested groups with the pretest as covariate.

They also noted that the last two groups of Design 5 form **Design 6**, the
posttest-only control group design. Design 6 relies on randomization rather
than a pretest for the equivalence of its groups (p. 25). For many purposes
they thought the extra gains of Design 5 might not be worth more than double
the effort (p. 26).

Huck and Sandler (1973) refined this advice:

- **ANCOVA with a pretesting main effect.** The analysis of covariance of
  the pretested groups remains valid even when pretesting has a main effect,
  because random assignment makes that effect the same in both groups
  (pp. 54–55).
- **When the interaction is significant.** The t test on the unpretested
  groups still answers whether the treatment works (p. 55).

`fit_solomon_classic()` computes these tests as Tests A–H. Its `history`
component compares the unpretested control posttest with the pretests, as
Campbell and Stanley describe.

Two papers of 1968 carried the design further:

- **Bracht and Glass (1968)** listed pretest sensitization, posttest
  sensitization, and the timing of measurement among the threats to
  external validity (p. 439). Reviewing the evidence, they found the pretest
  effect most likely when the outcome is a self-report of personality,
  attitude, or opinion, and the evidence for achievement inconclusive
  (p. 463). They also questioned Solomon's spelling result: the pretested
  groups were sent out of the room, and the groups were matched only by
  teachers' judgments (p. 462).
- **Solomon and Lessac (1968)** applied the design to studies of
  development, where a pretest could itself enrich an isolated animal's
  experience. Solomon now presented the design as a two-by-two factorial,
  with the interaction as a difference of differences, and still used the
  pretested groups' combined mean as the best estimate of the unpretested
  groups' starting point (pp. 146–147).

## How often does pretesting matter? (1982)

Willson and Putnam (1982) pooled 32 studies in a meta-analysis of pretest
effects. Among randomized studies, pretesting raised posttest scores by an
average of about a fifth of a standard deviation. There was little evidence
of a pretest-by-treatment interaction.

The article "Planning a Solomon Study" uses their estimates as planning
values. They also frame a question that recurs below: how much should a
researcher pay to detect an effect that is often small?

## A single test for the whole design (1988)

Walton Braver and Braver (1988) answered Campbell and Stanley's analysis
problem with meta-analysis:

- **Test I.** Treat the pretested and unpretested comparisons as two
  studies of the same effect, and combine their one-tailed p-values with
  Stouffer's method. Because it uses all the data, they called Test I "the
  most powerful single test" of the treatment effect (p. 153).
- **The sequence.** A flowchart embedded Test I in Tests A–I. In the
  sequence as they drew it:
  - a significant Test A (the interaction) leads to the simple effects,
    Tests B and C;
  - otherwise testing moves through Test D (the main effect), then a
    pretested-groups test (E, F, or G), then Test H (the unpretested
    groups), stopping at the first significant test;
  - Test I is reached only when all of these are nonsignificant
    (pp. 151–153).
- **Error rates.** They argued that the significance level at each step is
  best read as a probability conditional on having reached that step
  (p. 153, footnote 3).

Their worked example reports Test I as z = 2.05, p = .040 (p. 153). That p
is two-tailed.

## The exchange (1990)

The procedure drew an immediate exchange in *Perceptual and Motor Skills*.

- **Sawilowsky and Markman (1990a).** They constructed an example in which
  Test I missed an effect that the standard analyses detected, as Braver
  and Walton Braver (1990, p. 321) summarize it.
- **Braver and Walton Braver (1990).**
  - They replied that the example did not meet the condition under which
    they recommended Test I: a nonsignificant Test A, meaning the treatment
    effects in the pretested and unpretested groups are homogeneous
    (pp. 321–322).
  - They agreed, however, that using meta-analysis selectively, as their
    flowchart did, introduced selection bias.
  - They therefore amended the flowchart: once Tests A and D are
    nonsignificant, the analyst should complete all tests through Test I,
    which they regarded as the most definitive because only it uses all the
    data (p. 322).
- **Sawilowsky and Markman's (1990b) rejoinder.**
  - They set out four possible outcomes: a treatment effect with or without
    pretest sensitization, and no effect with or without it.
  - When no sensitization is expected, they argued, the Solomon design has
    no advantage over a two-group design (p. 424).
  - They called for a systematic study of the method's properties before
    its use (p. 425).

## The Monte Carlo evidence (1994, 1996)

Sawilowsky et al. (1994) supplied that study. They objected that treating
each step's significance level as conditional was "untenable" (p. 366),
since the experiment-wise error rate can be estimated directly. With 30
participants per group and normal data, the 1988 sequence falsely declared
an effect about 14% of the time, nearly three times the nominal 5% (p. 368,
Table 2). When the tests were not run conditionally, the classical tests
were more powerful than Test I in most of the treatment models studied
(pp. 369–371).

Sawilowsky (1996) extended the evidence:

- **The 1995 revision.** He reported that Walton Braver and Braver (1995, as
  cited in Sawilowsky, 1996, p. 2) had revised the procedure to remove
  Test D; that revision was never published.
- **Error rates.** For most of the ten distributions he studied, the error
  rate was about .12–.14 with or without Test D (Table 1). It stayed near
  .12 across pretest–posttest correlations from 0 to .95 and group sizes
  from 3 to 30 (Table 3).
- **Two remedies.** He proposed two Bonferroni-type allocations of alpha
  across Tests A, E, H, and I (Table 4).
- **A preferred alternative.** Where theory and evidence suggest no
  sensitization, he preferred a two-group randomized design analyzed at the
  full alpha to the Solomon sequence at the reduced alpha its control
  requires (p. 5).

solomonR's replication of these published rates, including both readings of
Test I's criterion, is reported in "Historical Tests: Replicating the
Published Error Rates".

## After the debate (2011–2025)

Later work turned from the test sequence to the design's purpose:

- **McCambridge et al. (2011).** They reviewed Solomon studies of behavior
  change and found sparse, inconsistent evidence of pretest effects.
- **French et al. (2021a, 2021b).** The MERIT recommendations treat
  measurement reactivity as a source of bias in trials. They place the
  Solomon design among several ways to detect or avoid it.
- **El Karkri et al. (2025a).** A classroom Solomon study that reports full
  cell statistics and a sensitization effect.
- **El Karkri et al. (2025b).** They set out an analysis path for
  categorical outcomes.

## What solomonR implements

| Procedure | Source | In solomonR | Status |
|---|---|---|---|
| Improvement scores and the interaction term I | Solomon (1949) | `fit_solomon_1949()`, data `solomon1949` | historical |
| Tests A–H | Campbell (1957); Campbell & Stanley (1963/1966); Huck & Sandler (1973) | `fit_solomon_classic()` | historical |
| History/maturation check | Solomon (1949); Campbell (1957); Campbell & Stanley (1963/1966) | `fit_solomon_classic()$history` | historical |
| Lana's analysis of variance of posttest means | Lana (1959) | `solomon_from_summary()`, data `lana1959` | historical |
| Test I and the 1988 sequence | Walton Braver & Braver (1988) | `fit_solomon_classic(flow = "1988")` | historical |
| 1990 amendment | Braver & Walton Braver (1990) | `fit_solomon_classic(flow = "1990")` | historical |
| 1995 revision without Test D | Walton Braver & Braver (1995, as cited in Sawilowsky, 1996) | `fit_solomon_classic(flow = "1995")` | historical |
| Alpha allocations | Sawilowsky (1996) | `fit_solomon_classic(alpha_allocation = ...)` | historical |
| Categorical rule | El Karkri et al. (2025b) | `fisher_solomon()` | historical |
| One model for all four groups, with robust standard errors | contemporary practice; see the methods guide | `fit_solomon_glm()` | recommended |

`plot_classic_flow()` draws each version of the sequence:

```{r flows, fig.height = 5, fig.alt = "The historical Tests A-I decision sequence in its 1995 version, without Test D."}
plot_classic_flow(flow = "1995")
```

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References
Bracht, G. H., & Glass, G. V. (1968). The external validity of experiments.
*American Educational Research Journal, 5*(4), 437–474.
https://doi.org/10.3102/00028312005004437

Braver, S. L., & Walton Braver, M. C. (1990). Meta-analysis for Solomon
four-group designs reconsidered: A reply to Sawilowsky and Markman.
*Perceptual and Motor Skills, 71*(1), 321–322.
https://doi.org/10.2466/pms.1990.71.1.321

Campbell, D. T. (1957). Factors relevant to the validity of experiments in
social settings. *Psychological Bulletin, 54*(4), 297–312.
https://doi.org/10.1037/h0040950

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025a). The dual impact of
pretest sensitisation and the cognitive acceleration through science
education programme in the Solomon four-group design. *Brain Sciences,
16*(1), Article 64. https://doi.org/10.3390/brainsci16010064

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025b). Methodological
aspects of the Solomon four-group design: Detecting pre-test sensitisation
and analysing qualitative and quantitative variables in education research.
*Review of Education, 13*(1), Article e70050.
https://doi.org/10.1002/rev3.70050

Entwisle, D. R. (1961). Interactive effects of pretesting. *Educational and
Psychological Measurement, 21*(3), 607–620.
https://doi.org/10.1177/001316446102100307

French, D. P., Miles, L. M., Elbourne, D., Farmer, A., Gulliford, M.,
Locock, L., Sutton, S., McCambridge, J., & MERIT Collaborative Group.
(2021a). Reducing bias in trials due to reactions to measurement: Experts
produced recommendations informed by evidence. *Journal of Clinical
Epidemiology, 139*, 130–139. https://doi.org/10.1016/j.jclinepi.2021.06.028

French, D. P., Miles, L. M., Elbourne, D., Farmer, A., Gulliford, M.,
Locock, L., Sutton, S., McCambridge, J., & MERIT Collaborative Group.
(2021b). Reducing bias in trials from reactions to measurement: The MERIT
study including developmental work and expert workshop. *Health Technology
Assessment, 25*(55), 1–72. https://doi.org/10.3310/hta25550

Huck, S. W., & Sandler, H. M. (1973). A note on the Solomon 4-group design:
Appropriate statistical analyses. *The Journal of Experimental Education,
42*(2), 54–55. https://doi.org/10.1080/00220973.1973.11011460

Lana, R. E. (1959). Pretest-treatment interaction effects in attitudinal
studies. *Psychological Bulletin, 56*(4), 293–300.
https://doi.org/10.1037/h0044646

McCambridge, J., Butor-Bhavsar, K., Witton, J., & Elbourne, D. (2011). Can
research assessments themselves cause bias in behaviour change trials? A
systematic review of evidence from Solomon 4-group studies. *PLoS ONE,
6*(10), Article e25223. https://doi.org/10.1371/journal.pone.0025223

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
https://digitalcommons.wayne.edu/coe_tbf/29/

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S. (1994).
Meta-analysis and the Solomon four-group design. *The Journal of
Experimental Education, 62*(4), 361–376.
https://doi.org/10.1080/00220973.1994.9944140

Sawilowsky, S. S., & Markman, B. S. (1990a). Another look at the power of
meta-analysis in the Solomon four-group design. *Perceptual and Motor
Skills, 71*(1), 177–178. https://doi.org/10.2466/pms.1990.71.1.177

Sawilowsky, S. S., & Markman, B. S. (1990b). Rejoinder to Braver and Walton
Braver. *Perceptual and Motor Skills, 71*(2), 424–426.
https://doi.org/10.2466/pms.1990.71.2.424

Solomon, R. L. (1949). An extension of control group design. *Psychological
Bulletin, 46*(2), 137–150. https://doi.org/10.1037/h0062958

Solomon, R. L., & Lessac, M. S. (1968). A control group design for
experimental studies of developmental processes. *Psychological Bulletin,
70*(3, Pt. 1), 145–150. https://doi.org/10.1037/h0026147

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150

Willson, V. L., & Putnam, R. R. (1982). A meta-analysis of pretest
sensitization effects in experimental design. *American Educational Research
Journal, 19*(2), 249–258. https://doi.org/10.3102/00028312019002249
