---
title: "Latent Contrasts: Validating the Invariance Check"
description: >
  Simulation study of which measurement-invariance criterion should govern
  latent Solomon mean contrasts, how often each criterion falsely rejects or
  detects noninvariance in Solomon-sized groups, and how noninvariance
  biases the sensitization contrast.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("invariance-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("invariance-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

crit <- c("chisq", "chen", "both")
crit_label <- c(chisq = "C1: scaled chi-square difference",
                chen = "C2: change in fit (Chen, 2007)",
                both = "C3: both")
pattern_label <- c("1: none", "2: intercept +0.3, pretested groups",
                   "3: intercept +0.6, pretested groups", "4: intercept +0.6, U0 only",
                   "5: loading -0.3, pretested groups")

rej <- p[p$criterion %in% crit, ]
w <- stats::reshape(rej[, c("scenario", "k", "n", "pattern", "criterion", "measure", "value")],
                    idvar = c("scenario", "k", "n", "pattern", "criterion"),
                    timevar = "measure", direction = "wide")
names(w) <- sub("^value[.]", "", names(w))

null <- w[w$pattern == 1, ]
null$larger <- pmax(null$reject_metric, null$reject_scalar)
alt <- w[w$pattern >= 2, ]
alt$detect <- ifelse(alt$pattern == 5, alt$reject_metric, alt$reject_scalar)

eligible <- vapply(crit, function(cr) {
  d <- null[null$criterion == cr, ]
  over <- d$larger > 0.060
  !(sum(over) > 1 || any(over & d$n >= 60))
}, logical(1))
detection <- tapply(alt$detect, alt$criterion, mean)[crit]
low <- stats::aggregate(detect ~ scenario + k + n + pattern, alt, max)
low <- low[low$detect < 0.5, ]

a3 <- p[p$criterion %in% c("scalar_model", "partial_model"), ]
a3w <- stats::reshape(a3[, c("scenario", "k", "n", "pattern", "criterion", "measure", "value", "mcse")],
                      idvar = c("scenario", "k", "n", "pattern", "criterion"),
                      timevar = "measure", direction = "wide")

pct <- function(x) sprintf("%.1f%%", 100 * x)
rng <- function(x, d = 3) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x))
fails <- max(p$failures)
```

`fit_solomon_sem_latent()` compares latent means across the four Solomon
groups. That comparison assumes scalar measurement invariance: equal
loadings and intercepts in every group (Meredith, 1993). This study asked
which published criterion, if any, should decide whether the package
refuses the latent contrasts when invariance fails.

The aims, data-generating mechanisms, estimands, methods, performance
measures, and decision rules were posted on
[issue #55](https://github.com/JUhalt/solomonR/issues/55) before any run,
following the ADEMP structure of Morris et al. (2019). The results come
from package commit `r substr(run$commit, 1, 7)`, with `r reps`
replications per scenario.

## Design

**Data.** One latent factor with k = 3, 4, or 6 normally distributed
indicators:

- loadings of 0.8, 0.9, 0.7, 0.8, 0.9, and 0.7 (the first k);
- residual SD 0.6;
- latent variance 1 in each group;
- latent means of 0.4, 0, 0.4, and 0 in the pretested treatment (P1),
  pretested control (P0), unpretested treatment (U1), and unpretested
  control (U0) groups, so there is a treatment effect and no
  sensitization.

There were 30, 60, or 120 participants per group.

**Noninvariance patterns,** all on the last indicator:

```{r patterns}
knitr::kable(data.frame(Pattern = pattern_label,
                        `Step that should fail` = c("none", "scalar", "scalar", "scalar", "metric"),
                        check.names = FALSE))
```

Patterns 2 and 3 are the response shift a pretest could cause in both
pretested groups. The full factorial has `r length(unique(p$scenario))`
scenarios.

**Criteria.** `invariance_solomon()` with MLR estimation, under three
decision criteria:

- **C1:** the scaled chi-square difference test at alpha = .05 (Satorra &
  Bentler, 2001), which Vandenberg and Lance (2000, p. 46) recommend as the
  primary criterion;
- **C2:** the change in CFI, RMSEA, and SRMR, judged against Chen's (2007,
  pp. 501–502) cutoffs;
- **C3:** both C1 and C2 indicate noninvariance.

**Decision rules** (fixed before the run):

1. **Eligibility.** A criterion is eligible to govern refusals if it
   falsely rejects metric or scalar invariance at a rate of .060 or less in
   at least 8 of the 9 pattern-1 scenarios, and in every pattern-1 scenario
   with 60 or more per group.
2. **Choice.** The eligible criterion that detects noninvariance most often
   becomes the refusal rule.
3. **No eligible criterion.** If none is eligible, the fit runs the check,
   reports it, and warns, without refusing.
4. **Small samples.** Group sizes at which every criterion detects
   noninvariance less than half the time are named as too small for the
   check to protect the contrasts.

**Amendment.** The simulation script recorded the rejection rate of each
step separately, not the rate of rejecting either step. The rate of
rejecting either step is at least the larger of the two, so eligibility
below uses the larger rate. A criterion that fails with the larger rate
fails with the combined rate too. Here every criterion failed on the
larger rate alone, so the missing combined rate does not change any
decision.

## Failures

`r if (fails == 0) "No model failed to converge in any replication, so every scenario is in the supported range." else sprintf("Up to %d replications per scenario failed to converge.", fails)`

## False rejection when invariance holds

The table gives, for each pattern-1 scenario, the larger of the two
steps' false-rejection rates. Values above .060 are outside the tolerance.

```{r null}
tab <- stats::reshape(null[, c("k", "n", "criterion", "larger")], idvar = c("k", "n"),
                      timevar = "criterion", direction = "wide")
tab <- tab[order(tab$n, tab$k), ]
names(tab) <- c("Indicators", "n per group", crit_label[sub("^larger[.]", "", names(tab)[-(1:2)])])
knitr::kable(tab, digits = 3, row.names = FALSE)
```

```{r eligibility}
elig_text <- if (any(eligible)) {
  sprintf("Eligible: %s.", paste(crit_label[eligible], collapse = "; "))
} else {
  "No criterion was eligible."
}
```

`r elig_text`

- **The chi-square difference test** falsely rejected at rates of
  `r rng(null$larger[null$criterion == "chisq"])`.
- **Chen's cutoffs** ranged from
  `r rng(null$larger[null$criterion == "chen"])`. They rejected far too
  often with 30 or 60 per group (up to
  `r sprintf("%.3f", max(null$larger[null$criterion == "chen" & null$n <= 60]))`),
  as Chen (2007, p. 502) anticipated for RMSEA and SRMR in small samples.
  With 120 per group they stayed within the tolerance except with three
  indicators (`r sprintf("%.3f", null$larger[null$criterion == "chen" & null$n == 120 & null$k == 3])`).
- **Requiring both** ranged from
  `r rng(null$larger[null$criterion == "both"])`.

## Detection of noninvariance

Rates at which each criterion rejected the step that should fail
(patterns 2–5):

```{r detection}
dt <- stats::reshape(alt[, c("pattern", "k", "n", "criterion", "detect")],
                     idvar = c("pattern", "k", "n"), timevar = "criterion", direction = "wide")
dt <- dt[order(dt$pattern, dt$n, dt$k), ]
dt$pattern <- pattern_label[dt$pattern]
names(dt) <- c("Pattern", "Indicators", "n per group",
               crit_label[sub("^detect[.]", "", names(dt)[-(1:3)])])
knitr::kable(dt, digits = 3, row.names = FALSE)
```

Mean detection across patterns 2–5 was `r sprintf("%.3f", detection[["chisq"]])`
for C1, `r sprintf("%.3f", detection[["chen"]])` for C2, and
`r sprintf("%.3f", detection[["both"]])` for C3.
`r if (nrow(low)) sprintf("Every criterion detected noninvariance less than half the time in %d scenarios, all with %s per group (rule 4).", nrow(low), paste(sort(unique(low$n)), collapse = " or ")) else "In every scenario at least one criterion detected noninvariance at least half the time (rule 4)."`

```{r small}
s30 <- alt[alt$n == 30 & alt$pattern %in% c(2, 5), ]
```

The small shifts were hard to detect with 30 per group. For the intercept
shift of 0.3 (pattern 2) and the loading shift (pattern 5), the
chi-square test detected noninvariance in
`r rng(s30$detect[s30$criterion == "chisq"], 2)` of replications. Chen's
cutoffs detected it in `r rng(s30$detect[s30$criterion == "chen"], 2)`, at
the cost of the false rejections above.

## Bias in the sensitization contrast

The sensitization contrast is 0 in every pattern. The table gives its bias
and the coverage of its 95% interval under the scalar model, which assumes
full invariance, and under the partial model, which frees the last
indicator's intercept.

```{r aim3}
b <- a3w[order(a3w$pattern, a3w$n, a3w$k, a3w$criterion), ]
out <- data.frame(Pattern = pattern_label[b$pattern], Indicators = b$k, `n per group` = b$n,
                  Model = sub("_model$", "", b$criterion),
                  Bias = b$value.sens_bias, MCSE = b$mcse.sens_bias,
                  Coverage = b$value.sens_coverage, check.names = FALSE)
knitr::kable(out, digits = 3, row.names = FALSE)
p4 <- b[b$pattern == 4 & b$criterion == "scalar_model", ]
p4p <- b[b$pattern == 4 & b$criterion == "partial_model", ]
p23 <- b[b$pattern %in% 2:3 & b$criterion == "scalar_model", ]
z23 <- max(abs(p23$value.sens_bias / p23$mcse.sens_bias))
z4p <- max(abs(p4p$value.sens_bias / p4p$mcse.sens_bias))
# 95th percentile of the largest of m absolute standard normal deviates:
# the largest deviation chance alone produces across m unbiased scenarios.
max_null <- function(m) stats::qnorm((1 + 0.95^(1 / m)) / 2)
```

Two patterns of results stand out:

- **A shift common to both pretested groups** (patterns 2 and 3) left the
  sensitization contrast unbiased under the scalar model. The bias ranged
  from `r rng(p23$value.sens_bias)`. The largest was
  `r sprintf("%.1f", z23)` Monte Carlo standard errors from zero, within
  what chance alone produces across `r nrow(p23)` unbiased scenarios (95th
  percentile of the largest deviation: `r sprintf("%.1f", max_null(nrow(p23)))`). The
  contrast is a difference of differences within each pretest condition,
  so a shift shared by P1 and P0 cancels.
- **A shift in one group** (pattern 4, U0 only) biased the contrast under
  the scalar model, by `r rng(p4$value.sens_bias)` latent standard
  deviations, with coverage from `r rng(p4$value.sens_coverage)`. Freeing
  the shifted intercept removed the bias: under the partial model the
  largest bias was `r sprintf("%.1f", z4p)` Monte Carlo standard errors
  from zero.

## Decision

```{r decision}
decision_text <- if (any(eligible)) {
  "At least one criterion was eligible; see the rules above."
} else {
  paste("Under rule 3, `fit_solomon_sem_latent()` does not refuse latent contrasts.",
        "It runs `invariance_solomon()` on the POST indicators, stores the result in",
        "the fit's `invariance` element, prints a one-line summary, and warns (class",
        "`solomonR_invariance_warning`) when a criterion does not support scalar (or",
        "partial scalar) invariance. Researchers who established invariance elsewhere",
        "can set `check_invariance = FALSE`.")
}
```

`r decision_text`

The warning is a prompt to look at the invariance results, not a verdict.
In groups of 30 to 60, the criteria often flag invariance that holds.
Freeing parameters for a partial-invariance model must rest on
substantive grounds, not on the data at hand (Byrne et al., 1989, p. 465;
Vandenberg & Lance, 2000, p. 38).

## Limitations

- **The data-generating mechanisms.** Normal data, one factor, equal group
  sizes, and noninvariance in one indicator only.
- **The latent contrasts.** Bias and coverage were studied for the
  sensitization contrast only, under patterns 1–4.
- **The criteria.** Other criteria, such as modification indices or
  Bayesian approaches, were not studied.

The script is `vignettes/articles/invariance-validation/invariance-simulation.R`
in the package repository.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Byrne, B. M., Shavelson, R. J., & Muthén, B. (1989). Testing for the
equivalence of factor covariance and mean structures: The issue of partial
measurement invariance. *Psychological Bulletin, 105*(3), 456–466.
https://doi.org/10.1037/0033-2909.105.3.456

Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of
measurement invariance. *Structural Equation Modeling: A Multidisciplinary
Journal, 14*(3), 464–504. https://doi.org/10.1080/10705510701301834

Meredith, W. (1993). Measurement invariance, factor analysis and factorial
invariance. *Psychometrika, 58*(4), 525–543.
https://doi.org/10.1007/BF02294825

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086

Satorra, A., & Bentler, P. M. (2001). A scaled difference chi-square test
statistic for moment structure analysis. *Psychometrika, 66*(4), 507–514.
https://doi.org/10.1007/BF02296192

Vandenberg, R. J., & Lance, C. E. (2000). A review and synthesis of the
measurement invariance literature: Suggestions, practices, and
recommendations for organizational research. *Organizational Research
Methods, 3*(1), 4–70. https://doi.org/10.1177/109442810031002
