---
title: "Missing Posttests: Validating the Sensitivity Analysis"
description: >
  Simulation study of fit_solomon_mi(): multiple imputation of missing
  posttests under missing at random and under known departures from it.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("mi-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("mi-validation/run-information.csv", stringsAsFactors = FALSE)
agree <- utils::read.csv("mi-validation/agreement.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

# Tolerances declared on issue #82 before any run.
p$null <- abs(p$true_value) < 1e-8
p$bias_ok <- abs(p$bias) <= 2 * p$bias_mcse
p$cov_ok <- p$coverage >= 0.94 & p$coverage <= 0.96
p$se_ok <- abs(p$relerr) <= 10
p$t1_ok <- ifelse(p$null, p$rejection >= 0.04 & p$rejection <= 0.06, TRUE)
p$all_ok <- p$bias_ok & p$cov_ok & p$se_ok & p$t1_ok

# M3 is the analysis with the true offsets; under pattern A it is M2.
m3 <- p[p$method == "mi_true" | (p$offsets == "A" & p$method == "mi_mar"), ]
big <- m3[m3$n >= 60, ]
short <- m3[!m3$all_ok, ]
agree$within <- abs(agree$mean_difference) <= 2 * agree$mcse
odd <- agree[!agree$within, ]

pct <- function(x) sprintf("%.1f%%", 100 * x)
f3 <- function(x) sprintf("%.3f", x)
rng <- function(x, d = 3) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x))
offset_label <- c(A = "A: none (missing at random)", B = "B: all four groups",
                  C = "C: the treatment groups", D = "D: the pretested treatment group only")
method_label <- c(complete_case = "M1: complete case", mi_mar = "M2: MI, delta = 0",
                  mi_true = "M3: MI, true offsets")
```

This article reports the simulation validation of `fit_solomon_mi()`, the
multiple-imputation analysis of missing posttests that
`tipping_point_solomon()` repeats over a range of offsets. The aims,
data-generating mechanisms, estimands, methods, performance measures,
tolerances, and decision rules were posted on
[issue #82](https://github.com/JUhalt/solomonR/issues/82) before any code or
run, following the ADEMP structure of Morris et al. (2019). The results come
from package commit `r substr(run$commit, 1, 7)`, with `r reps` replications
per scenario and `r run$imputations` imputations per analysis.

## Design

**Complete data.** The model of `simulate_solomon()` on a standardized
scale: a treatment effect of 0.5, no sensitization, a pretest effect of 0.2,
and a pretest-posttest correlation of 0.5.

**Missing posttests.** In the pretested groups, whether a posttest is
missing depends on the observed pretest (missing at random), with the odds
doubling per standard deviation. The unpretested groups have no pretest, so
their posttests are missing completely at random within group. Each unit
with a missing posttest has its true outcome shifted by the offset of its
group: the pattern-mixture model of Carpenter et al. (2023, Eq. 10.10).

**Scenarios.** `r length(unique(p$scenario))` in all:

- **Participants per group:** 30, 60, and 120.
- **Missing proportions:** 20% in every group, or 30% in the treatment
  groups and 10% in the control groups.
- **Offsets of −0.5 SD:** (A) none, so the posttests are missing at
  random; (B) in all four groups; (C) in the two treatment groups; and (D)
  in the pretested treatment group only.

**Estimands.** The four Solomon contrasts, defined on the means of all
randomized participants, including the shifted outcomes of those whose
posttest is missing.

**Methods.**

- **M1:** complete-case `fit_solomon_glm()` with HC3 standard errors.
- **M2:** `fit_solomon_mi(delta = 0)`, the analysis under missing at random.
- **M3:** `fit_solomon_mi()` with the true offsets. Under pattern A it is M2.

**Tolerances.** Bias within 2 Monte Carlo standard errors (MCSE); coverage
of 95% intervals from 0.940 to 0.960; model standard error within 10% of the
empirical standard error; and, where the true value is zero, a rejection
rate from 0.040 to 0.060.

## Failures

`r if (sum(p$failures) == 0 && all(p$errors == 0)) "No analysis failed or returned an error in any replication." else sprintf("%d analyses failed.", sum(p$failures))`

## The analysis with the true offsets

**Decision rule 1** required M3 to meet every tolerance in at least 90% of
its `r nrow(m3)` contrast-by-scenario cells and in every cell with 60 or more
participants per group. It met them in `r sum(m3$all_ok)` cells
(`r pct(mean(m3$all_ok))`) and in `r sum(big$all_ok)` of the `r nrow(big)`
cells with 60 or more per group, so **the rule was not met**, and the
function is not validated as a whole.

```{r rule1}
tab <- aggregate(cbind(cells = 1, all = all_ok, coverage = cov_ok, type1 = t1_ok, bias = bias_ok) ~ n,
                 m3, sum)
knitr::kable(data.frame(
  `Per group` = tab$n,
  `Every tolerance` = sprintf("%d of %d", tab$all, tab$cells),
  `Coverage` = sprintf("%d of %d", tab$coverage, tab$cells),
  `Type I error` = sprintf("%d of %d", tab$type1, tab$cells),
  `Bias` = sprintf("%d of %d", tab$bias, tab$cells),
  check.names = FALSE
), align = "lrrrr", caption = "Cells of M3 within each tolerance, by group size (all offsets and missing proportions).")
```

Every shortfall was a conservative interval or a small bias flag; no
interval covered too rarely:

```{r shortfalls}
knitr::kable(data.frame(
  `Per group` = short$n,
  Missing = short$missing,
  Offsets = short$offsets,
  Contrast = short$contrast,
  `Bias (SD)` = sprintf("%.4f", short$bias),
  `Bias / MCSE` = sprintf("%.1f", short$bias / short$bias_mcse),
  Coverage = sprintf("%.3f", short$coverage),
  `Type I` = ifelse(short$null, sprintf("%.3f", short$rejection), ""),
  check.names = FALSE
), align = "rllllrrr", caption = "M3 cells outside a tolerance.")
```

- **With 60 or more per group,** coverage and Type I error were within
  tolerance in every cell: coverage ranged from `r rng(big$coverage, 4)`.
  Two contrasts in one scenario (a common offset with differential
  missingness, 60 per group) had bias just beyond 2 MCSE:
  `r paste(sprintf("%s SD (%.1f MCSE)", sprintf("%.3f", big$bias[!big$bias_ok]), big$bias[!big$bias_ok] / big$bias_mcse[!big$bias_ok]), collapse = " and ")`.
- **With 30 per group,** intervals were conservative: mean coverage
  `r f3(mean(m3$coverage[m3$n == 30]))` (up to `r f3(max(m3$coverage[m3$n == 30]))`),
  with the model standard error exceeding the empirical standard error by
  `r sprintf("%.1f%%", mean(m3$relerr[m3$n == 30]))` on average. The
  complete-case HC3 analysis under missing at random averaged coverage of
  `r f3(mean(p$coverage[p$method == "complete_case" & p$offsets == "A" & p$n == 30]))`
  at this size, with its model standard error
  `r sprintf("%.1f%%", mean(p$relerr[p$method == "complete_case" & p$offsets == "A" & p$n == 30]))`
  above the empirical one, so the imputation adds a little conservatism to
  that of the complete-data analysis. (The package's earlier study found HC3
  intervals close to nominal, 0.951 to 0.954, with 20 or more per cell;
  issues #10 and #22.)

A method with no bias would exceed 2 MCSE in about 5 of 96 cells by chance;
`r sum(!m3$bias_ok)` did. That is context for reading the table, not a
change to the rule.

## Agreement with the complete-case analysis

Under missing at random, multiple imputation should agree with the
complete-case analysis (Carpenter et al., 2023, p. 256). **Decision rule
2** required the paired mean difference between M2 and M1 to lie within 2
MCSE of zero. It did in `r sum(agree$within)` of `r nrow(agree)`
comparisons, and no difference exceeded
`r sprintf("%.4f", max(abs(agree$mean_difference)))` SD in absolute value.

The exception was `r odd$contrast` with `r odd$n` per group and
`r odd$missing` missingness: `r sprintf("%.5f", odd$mean_difference)` SD,
`r sprintf("%.1f", abs(odd$mean_difference) / odd$mcse)` MCSE. In the
unpretested groups the imputation model is the group mean, and under its
posterior draw the expected imputed value equals the observed group mean,
so the imputation cannot move this contrast systematically. The difference
is attributed to chance among `r nrow(agree)` comparisons and to the Monte
Carlo error of a finite number of imputations.

## When posttests are not missing at random

**Decision rule 3** was descriptive: the bias of the two analyses that
assume missing at random (M1 and M2) when the missing posttests are shifted.
The two had nearly the same bias throughout.

```{r mar-bias}
mar <- p[p$offsets != "A" & p$method %in% c("complete_case", "mi_mar"), ]
agg <- aggregate(cbind(bias, coverage) ~ offsets + missing + method + contrast, mar, mean)
agg <- agg[agg$method == "mi_mar", ]
wide <- reshape(agg[, c("offsets", "missing", "contrast", "bias")], idvar = c("offsets", "missing"),
                timevar = "contrast", direction = "wide")
names(wide) <- sub("^bias\\.", "", names(wide))
wide <- wide[order(wide$offsets, wide$missing), ]
knitr::kable(data.frame(
  Offsets = offset_label[wide$offsets], Missing = wide$missing,
  ATE = sprintf("%.3f", wide$`ATE (avg over pretest)`),
  `Pretest x Treatment` = sprintf("%.3f", wide$`Pretest x Treatment`),
  `Treatment | pretested` = sprintf("%.3f", wide$`Treatment | pretested`),
  `Treatment | unpretested` = sprintf("%.3f", wide$`Treatment | unpretested`),
  check.names = FALSE
), align = "llrrrr", row.names = FALSE,
caption = "Mean bias (SD units) of M2, the MI analysis under missing at random, averaged over group sizes. M1 was within 0.001 of these values.")
```

- **A common offset** biased nothing when the missing proportions were
  equal, because the shifts cancel in every contrast. With differential
  missingness it biased the treatment contrasts.
- **Offsets in the treatment groups** biased the treatment contrasts but
  hardly the sensitization contrast, because the offset and the missing
  proportion were the same in both treatment groups.
- **An offset in the pretested treatment group alone** biased the
  sensitization contrast by
  `r sprintf("%.2f", wide[wide$offsets == "D" & wide$missing == "equal", "Pretest x Treatment"])`
  SD with equal missingness and
  `r sprintf("%.2f", wide[wide$offsets == "D" & wide$missing == "differential", "Pretest x Treatment"])`
  SD with differential missingness. This is the Solomon-specific case: a departure from missing at random
  in one pretested group can hide a true difference in sensitization, or
  create a spurious one, while the analysis under missing at random
  reports none.

This is why `tipping_point_solomon()` lets the offsets differ between the
pretested and unpretested groups: `groups = 1` shifts the pretested treatment
group alone, the pattern that bears on sensitization.

## Number of imputations

**Decision rule 4** required the Monte Carlo error from `m = 100`
imputations to stay below 10% of the pooled standard error. The largest
ratio was `r sprintf("%.3f", max(p$max_mc_ratio, na.rm = TRUE))`, so the
default stays at 100. The mean fraction of missing information was
`r rng(p$mean_fmi[p$method != "complete_case"], 2)`.

## What this means for users

- With 60 or more participants per group, the intervals of
  `fit_solomon_mi()` had nominal coverage and Type I error, both under
  missing at random and under the pattern-mixture departures studied when
  the offsets were right.
- With 30 per group, they were conservative: slightly too wide, so tests
  lose a little power but do not reject too often.
- An analysis under missing at random can be biased in the sensitization
  contrast when the departure from it differs between the pretested groups.
  A tipping-point analysis with `groups = 1` shows how large such a
  departure would need to be to change the conclusion.
- The functions remain experimental, because decision rule 1 was not met.

## Deviation from the protocol

A first run, from commit 633a69e, was stopped after 11 of its 96 task
parts, and its results were discarded unread. `fit_solomon_mi()` had been
changed to compute each completed-data analysis from matrix products rather
than by calling `fit_solomon_glm()`; the package's tests check that the
results agree to 1e-10. The reported run is entirely from commit
`r substr(run$commit, 1, 7)`, and the design, methods, and rules are those
of the protocol.

## Reproducing the study

From the package root, `Rscript vignettes/articles/mi-validation/mi-simulation.R`
writes `performance.csv` and `run-information.csv`, and
`Rscript vignettes/articles/mi-validation/mi-agreement.R` then writes
`agreement.csv` from the saved replications. The run used R
`r sub("R version ([0-9.]+).*", "\\1", run$R_version)` and
`r run$workers` workers and took from `r run$started` to `r run$finished`.

## References

Carpenter, J. R., Bartlett, J. W., Morris, T. P., Wood, A. M., Quartagno,
M., & Kenward, M. G. (2023). *Multiple imputation and its application* (2nd
ed.). Wiley. https://doi.org/10.1002/9781119756118

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086
