---
title: "Missing and Repeated Posttests"
description: >
  How to analyze a Solomon study when some posttests are missing, and when
  the posttest is measured on several occasions and participants drop out.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
has_mmrm <- requireNamespace("mmrm", quietly = TRUE)
```

Two situations call for more than one model fitted to complete data:

- **Some posttests are missing.** The main analysis rests on an assumption
  about why they are missing, and a sensitivity analysis asks how far that
  assumption could be wrong before the conclusion changes (White et al.,
  2011).
- **The posttest is measured on several occasions,** and some participants
  drop out along the way. A model for all occasions at once uses every
  observed score and stays valid when dropout depends on earlier scores
  (Fitzmaurice et al., 2011).

Both are shown here on simulated data whose true effects are known. The
methods are validated in the articles
[Missing Posttests: Validating the Sensitivity Analysis](mi-validation.html)
and
[Longitudinal Designs: Validating the Repeated-Measures Analysis](mmrm-validation.html).

## Missing posttests

### The data

A study with 50 participants per group, a treatment effect of 5 points, no
sensitization, and a pretest-posttest correlation of 0.6. In the pretested
groups, participants with lower pretests are more likely to miss the
posttest; in the unpretested groups, one in five posttests is missing at
random.

```{r mi-data}
d <- simulate_solomon(n = 50, delta = 5, sens = 0, pretest_effect = 2, rho = 0.6,
                      sigma = 10, mean = 50, digits = 0, seed = 2082)
set.seed(2083)
p_missing <- ifelse(d$pretested == 1, stats::plogis(-1.4 - 0.08 * (d$y_pre - 50)), 0.2)
d$y_post[stats::runif(nrow(d)) < p_missing] <- NA
check_solomon_missing(y_post, treat, pretested, y_pre, data = d)
```

### The main analysis

Here the posttests are missing at random given group and, in the pretested
groups, the pretest: whether a posttest is missing depends only on things
the model includes. The model fitted to the observed posttests is then a
valid main analysis:

```{r mi-glm}
fit <- fit_solomon_glm(y_post, treat, pretested, y_pre, data = d)
fit$effects[, c("contrast", "estimate", "std.error", "conf.low", "conf.high", "p.value")]
```

Multiple imputation under the same assumption agrees with it, as it should
(Carpenter et al., 2023, p. 256):

```{r mi-mar}
mar <- fit_solomon_mi(y_post, treat, pretested, y_pre, seed = 1, data = d)
mar$effects[, c("contrast", "estimate", "std.error", "fmi")]
```

The `fmi` column is the fraction of missing information for each contrast.

### How far could the assumption be wrong?

Missing at random cannot be checked from the data. The sensitivity analysis
asks: if the missing posttests were systematically lower (or higher) than
the observed ones predict, by how much would they have to differ to change
the conclusion? `fit_solomon_mi()` answers for one set of offsets, and
`tipping_point_solomon()` for a range of them. By default the range is one
standard deviation either way; here the treatment effect is large enough
that its conclusion holds throughout that range, so the range is widened
to show where it would change. The offsets are applied to the treatment
groups, the departure that bears on the treatment effect:

```{r tipping, fig.width = 7, fig.height = 4.5, fig.alt = "Pooled average treatment effect with its confidence band as the offset added to imputed treatment-group posttests goes from minus to plus one standard deviation."}
tp <- tipping_point_solomon(y_post, treat, pretested, y_pre, groups = "treatment",
                            deltas = seq(-25, 10, by = 2.5), seed = 1, data = d)
tp$tipping
tp$tipping_sd
plot_tipping_point(tp)
```

`tp$tipping` gives the smallest offsets, in points, at which the
conclusion about the average treatment effect changes, and `tp$tipping_sd`
gives them in standard deviations of the observed posttests (`NA` means
the conclusion did not change within the range tried). Here the effect
would lose its statistical significance only if the missing treatment-group
posttests were about `r abs(round(tp$tipping_sd[["negative"]], 1))`
standard deviations lower than the imputation model predicts.
Whether such an offset is plausible is a substantive judgment. White et al.
(2011) advise specifying sensitivity analyses in detail before the
unblinded data are seen, and `analysis_plan_solomon()` includes this
analysis in the plan.

### Sensitization and the pretested groups

A departure from missing at random that differs between the pretested and
unpretested groups bears on the sensitization contrast. In the validation
study, a departure confined to the pretested treatment group biased the
analyses that assume missing at random by 0.10 to 0.16 standard deviations.
`groups = 1` shifts that group alone:

```{r tipping-sens}
tp_sens <- tipping_point_solomon(y_post, treat, pretested, y_pre,
                                 contrast = "Pretest x Treatment", groups = 1,
                                 deltas = seq(-30, 30, by = 2.5), seed = 1, data = d)
tp_sens$tipping_sd
```

Here a significant Pretest x Treatment contrast would appear only if the
missing posttests of the pretested treatment group were about
`r round(tp_sens$tipping_sd[["positive"]], 1)` standard deviations higher
than predicted.

### Reporting

`report_solomon()` writes the method, the offsets explored, and the tipping
points, with their references:

```{r report-tp}
report_solomon(tp)
```

## Repeated posttests

### The data

The same design with posttests right after the treatment, at 6 months, and
at 12 months. The treatment effect fades from 5 to 3 points. After each
posttest, participants with lower scores are more likely to drop out, so
the later posttests are missing at random given the earlier ones.

```{r long-data}
set.seed(57)
n <- 60
g <- rep(1:4, each = n)
treat <- as.integer(g %in% c(1, 3))
pretested <- as.integer(g <= 2)
sds <- c(10, 10, 12, 14)  # the baseline, then the three posttests
cov <- diag(sds) %*% outer(0:3, 0:3, function(i, j) 0.6^abs(i - j)) %*% diag(sds)
z <- MASS::mvrnorm(4 * n, rep(0, 4), cov)
y <- sapply(1:3, function(t) 50 + 2 * pretested + c(5, 4, 3)[t] * treat) + z[, 2:4]
last <- rep(3L, 4 * n)
for (t in 1:2) {
  drop <- last == 3L & stats::runif(4 * n) < stats::plogis(-2 - 0.07 * (y[, t] - 50))
  last[drop] <- t
}
for (t in 2:3) y[last < t, t] <- NA
occasions <- c("post", "6 months", "12 months")
long <- data.frame(
  id = rep(seq_len(4 * n), 3),
  occasion = factor(rep(occasions, each = 4 * n), levels = occasions),
  y_post = as.vector(y),
  treat = rep(treat, 3),
  pretested = rep(pretested, 3),
  y_pre = rep(ifelse(pretested == 1, 50 + z[, 1], NA), 3)
)
head(long)
```

The data are in long format: one row per participant and occasion, with the
participant's identifier.

### The model

`fit_solomon_mmrm()` fits a mixed model for repeated measures (Mallinckrodt
et al., 2008) and reports the four Solomon contrasts and the pretest
effects at each occasion, and the change in sensitization from the first
occasion to the last:

```{r mmrm, eval = has_mmrm}
mm <- fit_solomon_mmrm(y_post, treat, pretested, id, occasion, y_pre, data = long)
mm
```

The model needs the mmrm package. It estimates the covariance of the
posttests separately for pretested and unpretested participants, because
the pretest adjustment makes their residual covariance differ, and it uses
Kenward-Roger degrees of freedom. The table of observed posttests shows the
dropout.

### Why not analyze each occasion separately?

An analysis of the participants still observed at 12 months uses only them,
and it is valid only if dropout is unrelated to the outcome. Here dropout
followed low scores:

```{r per-occasion, eval = has_mmrm}
late <- long[long$occasion == "12 months" & !is.na(long$y_post), ]
cc <- fit_solomon_glm(y_post, treat, pretested, y_pre, data = late)
# The four treatment contrasts, selected from each effects table by label.
labels <- c("ATE (avg over pretest)", "Pretest x Treatment",
            "Treatment | pretested", "Treatment | unpretested")
pick <- function(effects) effects$estimate[match(labels, effects$contrast)]
data.frame(
  contrast = labels,
  complete_case = round(pick(cc$effects), 2),
  mmrm = round(pick(mm$effects[mm$effects$occasion == "12 months", ]), 2),
  truth = c(3, 0, 3, 3)
)
```

In this data set the two analyses differ by about a point at 12 months,
well within their standard errors of about 2 points, and here the
complete-case estimate happens to land closer to the truth. In a single
data set that is a matter of chance. The difference is systematic only on
average: across 2,000 data sets per scenario in the validation study,
per-occasion complete-case analyses were biased by up to 0.09 standard
deviations at the last occasion under heavy dropout, while the
repeated-measures model was not.

### Planning and reporting

`analysis_plan_solomon(occasions = ...)` writes a plan with this model as
the primary analysis and the contrasts at the primary occasion as
confirmatory, and `report_solomon()` reports the fit:

```{r mmrm-report, eval = has_mmrm}
report_solomon(mm, digits = 1)$method
```

For a published study with three posttest occasions, reproduced from its
group statistics, see
[Worked Example: Repeated Posttests](repeated-posttests-example.html).

## References

Carpenter, J. R., Bartlett, J. W., Morris, T. P., Wood, A. M., Quartagno,
M., & Kenward, M. G. (2023). *Multiple imputation and its application* (2nd
ed.). Wiley. https://doi.org/10.1002/9781119756118

Fitzmaurice, G. M., Laird, N. M., & Ware, J. H. (2011). *Applied longitudinal
analysis* (2nd ed.). Wiley. https://doi.org/10.1002/9781119513469

Mallinckrodt, C. H., Lane, P. W., Schnell, D., Peng, Y., & Mancuso, J. P.
(2008). Recommendations for the primary analysis of continuous endpoints in
longitudinal clinical trials. *Drug Information Journal, 42*(4), 303–319.
https://doi.org/10.1177/009286150804200402

White, I. R., Horton, N. J., Carpenter, J., & Pocock, S. J. (2011). Strategy
for intention to treat analysis in randomised trials with missing outcome
data. *BMJ, 342*, Article d40. https://doi.org/10.1136/bmj.d40
