---
title: "Validating fit_solomon_ml(): A Simulation Study"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      fig.width = 8, fig.height = 5)
library(ggplot2)

performance <- utils::read.csv("ml-validation/performance.csv")
run <- utils::read.csv("ml-validation/run-information.csv")

contrast_levels <- c(
  "ATE (avg over pretest)",
  "Pretest x Treatment",
  "Treatment | pretested",
  "Treatment | unpretested"
)
method_levels <- c(
  "ML, Wald (default)",
  "ML, Satterthwaite (small-sample option)",
  "GLM HC3 (t)"
)

performance$contrast <- factor(performance$contrast, contrast_levels)
performance$method <- factor(performance$method, method_levels)
cell_sizes <- sort(unique(performance$n_cell))

# Tolerances declared in the protocol before any results were examined.
performance$bias_ok <- abs(performance$bias) <= 2 * performance$mcse_bias
performance$coverage_ok <- performance$coverage >= 0.94 & performance$coverage <= 0.96
performance$se_ok <- abs(performance$relative_se_error) <= 5
performance$type1_ok <- with(
  performance,
  ifelse(contrast == "Pretest x Treatment" & sensitization == 0,
         rejection >= 0.04 & rejection <= 0.06, TRUE)
)
performance$failures_ok <- performance$n_failed == 0
performance$all_ok <- with(
  performance,
  bias_ok & coverage_ok & se_ok & type1_ok & failures_ok
)

wide_by_cell_size <- function(data, value, fmt) {
  data <- data[, c("method", "n_cell", value)]
  wide <- reshape(data, idvar = "method", timevar = "n_cell", direction = "wide")
  wide <- wide[order(wide$method), ]
  cells <- sort(unique(data$n_cell))
  out <- data.frame(Method = as.character(wide$method), check.names = FALSE)
  for (n in cells) {
    out[[paste("n =", n)]] <- fmt(wide[[paste(value, n, sep = ".")]])
  }
  out
}
```

This article reports the simulation validation of `fit_solomon_ml()`. The
aims, data-generating mechanisms, estimands, methods, performance measures,
and tolerances were posted to
[issue #10](https://github.com/JUhalt/solomonR/issues/10) before any results
were examined, following the ADEMP structure of Morris, White and Crowther
(2019).

The first run found that the default Wald intervals were too narrow in small
samples, which led to the small-sample option in
[issue #22](https://github.com/JUhalt/solomonR/issues/22). Before the results
reported here were examined, the protocol was amended on that issue in three
ways:

- 30 and 40 participants per cell were added;
- the exploratory small-sample method was replaced by the implemented
  `inference = "satterthwaite"` option;
- a rule for the small-sample warning threshold was added.

This run supersedes the first; the first run's results remain in the
package's version history.

## Design

**Aims.** Check whether `fit_solomon_ml()` recovers the four Solomon
contrasts, whether its standard errors are calibrated, and whether its 95%
intervals reach nominal coverage. Both inference options are evaluated:

- `inference = "wald"`, the default, which follows van Engelenburg (1999);
- `inference = "satterthwaite"`, which uses unbiased residual variances
  within each pretest condition and Welch-Satterthwaite degrees of freedom
  for contrasts that combine them (Satterthwaite, 1946; Welch, 1947).

Both are compared with the unified GLM using HC3 standard errors and t
reference distributions. A further aim is to locate the cell size below
which `fit_solomon_ml()` warns that Wald inference may be unreliable.

**Data-generating mechanisms.** A full factorial of `r nrow(unique(performance[, c("n_cell", "sensitization", "rho", "sd_ratio")]))`
scenarios crosses four factors:

- participants per cell: `r paste(cell_sizes, collapse = ", ")`;
- sensitization: 0 or 0.3;
- pretest-posttest correlation: 0, 0.5, or 0.8;
- residual standard deviation of treated relative to control participants:
  1 or 1.5.

Every participant has a latent baseline that only pretested participants
observe, so structural pretest absence is present in every scenario. The
treatment effect is 0.5 among unpretested participants and 0.5 plus the
sensitization among pretested participants.

**Performance measures.** Each of the following comes with its Monte Carlo
standard error (MCSE), from `r format(run$nsim, big.mark = ",")`
replications per scenario:

- bias;
- empirical and model-based standard errors;
- coverage of 95% intervals;
- the rejection rate at alpha = .05;
- fit failures.

**Tolerances.**

- Bias within 2 MCSE of zero.
- Coverage between 0.940 and 0.960.
- Model standard errors within 5% of the empirical standard error.
- Type I error for the sensitization test between 0.040 and 0.060 when there
  is no sensitization.
- No fit failures.

**Warning threshold rule.** The threshold is the smallest cell size at
which Wald inference meets both of the following criteria within each
residual-SD ratio, at that size and at every larger size studied:

- mean coverage of at least 0.940;
- mean sensitization Type I error of at most 0.060.

These are the liberal edges of the tolerance bands. The warning concerns
intervals that are too narrow, so over-coverage does not count against Wald
inference.

## Fit failures

```{r failures}
failures <- aggregate(
  n_failed ~ method,
  data = performance[performance$contrast == contrast_levels[1], ],
  FUN = sum
)
names(failures) <- c("Method", "Failed fits across all scenarios")
knitr::kable(failures)
```

## Coverage

```{r coverage-plot, fig.height = 6}
coverage <- aggregate(
  coverage ~ method + contrast + n_cell + sd_ratio,
  data = performance,
  FUN = mean
)
coverage$sd_ratio <- factor(paste("SD ratio", coverage$sd_ratio))

ggplot(coverage, aes(n_cell, coverage, colour = method)) +
  annotate("rect", xmin = 0, xmax = Inf, ymin = 0.94, ymax = 0.96, alpha = 0.15) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = cell_sizes) +
  facet_grid(sd_ratio ~ contrast, labeller = label_wrap_gen(width = 18)) +
  labs(x = "Participants per cell (log scale)",
       y = "Coverage of 95% intervals", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

Coverage is averaged over the correlation and sensitization conditions; the
shaded band is the pre-declared tolerance. Scenario-level values and their
MCSEs are in `ml-validation/performance.csv`.

```{r coverage-tables}
mean_coverage <- aggregate(coverage ~ method + n_cell, data = performance, FUN = mean)
knitr::kable(
  wide_by_cell_size(mean_coverage, "coverage", function(x) sprintf("%.3f", x)),
  caption = "Mean coverage across scenarios and contrasts"
)

lowest_coverage <- aggregate(coverage ~ method + n_cell, data = performance, FUN = min)
knitr::kable(
  wide_by_cell_size(lowest_coverage, "coverage", function(x) sprintf("%.3f", x)),
  caption = "Lowest coverage across scenarios and contrasts"
)
```

## Bias

```{r bias}
bias_summary <- data.frame(
  Method = method_levels,
  `Largest absolute bias` = sprintf(
    "%.3f", tapply(abs(performance$bias), performance$method, max)[method_levels]
  ),
  `Estimates beyond 2 MCSE` = sprintf(
    "%d of %d",
    as.integer(tapply(!performance$bias_ok, performance$method, sum)[method_levels]),
    as.integer(table(performance$method)[method_levels])
  ),
  check.names = FALSE
)
knitr::kable(bias_summary)
```

The three methods produce identical point estimates, so their bias is
identical. With 2 MCSE as the tolerance, roughly 5% of scenario-by-contrast
estimates are expected to fall outside it by chance even for an unbiased
estimator.

## Standard error calibration

```{r se}
relative_se <- aggregate(relative_se_error ~ method + n_cell, data = performance, FUN = mean)
knitr::kable(
  wide_by_cell_size(relative_se, "relative_se_error", function(x) sprintf("%+.1f%%", x)),
  caption = "Mean relative error of model-based standard errors"
)
```

## Type I error for sensitization

```{r type1}
null <- performance[performance$contrast == "Pretest x Treatment" & performance$sensitization == 0, ]
type1 <- aggregate(rejection ~ method + n_cell, data = null, FUN = mean)
knitr::kable(
  wide_by_cell_size(type1, "rejection", function(x) sprintf("%.3f", x)),
  caption = "Rejection rate of the Pretest x Treatment test with no sensitization (nominal .05)"
)
```

## Scenarios meeting every tolerance

```{r tolerance}
scenario_ok <- aggregate(all_ok ~ method + scenario + n_cell, data = performance, FUN = all)
meeting <- aggregate(all_ok ~ method + n_cell, data = scenario_ok, FUN = sum)
totals <- aggregate(all_ok ~ method + n_cell, data = scenario_ok, FUN = length)
meeting$all_ok <- sprintf("%d of %d", meeting$all_ok, totals$all_ok)
knitr::kable(
  wide_by_cell_size(meeting, "all_ok", identity),
  caption = "Scenarios in which all four contrasts meet every tolerance"
)
```

## Warning threshold

```{r threshold}
wald <- performance[performance$method == method_levels[1], ]
wald_null <- null[null$method == method_levels[1], ]
rule <- merge(
  aggregate(coverage ~ n_cell + sd_ratio, data = wald, FUN = mean),
  aggregate(rejection ~ n_cell + sd_ratio, data = wald_null, FUN = mean),
  by = c("n_cell", "sd_ratio")
)
rule$meets <- rule$coverage >= 0.940 & rule$rejection <= 0.060
rule <- rule[order(rule$n_cell, rule$sd_ratio), ]

meets_by_size <- tapply(rule$meets, rule$n_cell, all)
sizes <- as.integer(names(meets_by_size))
qualifies <- vapply(
  seq_along(sizes),
  function(i) all(meets_by_size[i:length(sizes)]),
  logical(1)
)
threshold <- if (any(qualifies)) sizes[which(qualifies)[1]] else max(sizes)

# The package constant must match the rule applied to these results.
stopifnot(threshold == solomonR:::.solomon_ml_small_cell)

knitr::kable(
  data.frame(
    `Participants per cell` = rule$n_cell,
    `SD ratio` = rule$sd_ratio,
    `Mean coverage` = sprintf("%.3f", rule$coverage),
    `Type I error` = sprintf("%.3f", rule$rejection),
    `Meets both criteria` = ifelse(rule$meets, "yes", "no"),
    check.names = FALSE
  ),
  caption = "Wald inference from fit_solomon_ml() against the threshold criteria"
)
```

By this rule, the threshold is `r threshold` participants per cell.
`fit_solomon_ml()` warns when the smallest cell has fewer participants and
`inference` was not supplied.

## Conclusions

**Estimand recovery is validated.** All three methods produce the same point
estimates. The largest absolute bias was 0.028 standard-deviation units, at 6
participants per cell. 17 of 336 scenario-by-contrast estimates (5.1%) fell
outside 2 MCSE, close to the 15 expected by chance. No fit failed.

**Default Wald inference from `fit_solomon_ml()` is too liberal in small
samples.**

- **Coverage.** Mean coverage of its 95% intervals was 0.893 with 6
  participants per cell, 0.920 with 10, 0.936 with 20, and 0.941 with 30,
  rising to 0.948 at 100. The treatment effect among pretested participants
  was affected most, with mean coverage of 0.871 at 6 per cell.
- **Standard errors.** They were 10.8% too small at 6 per cell and 6.5% too
  small at 10.
- **Type I error.** The sensitization test rejected a true null 9.9% of the
  time at 6 per cell, 7.7% at 10, and 6.4% at 20.
- **Tolerances.** Wald inference met every tolerance in none of the scenarios
  with 20 or fewer participants per cell, in 2 of 12 at 30, 3 of 12 at 40,
  5 of 12 at 50, and 8 of 12 at 100.

**The warning threshold is 40 participants per cell.**

- **At 30 per cell,** Wald inference narrowly failed the rule with unequal
  residual variances: mean coverage was 0.9399 and Type I error 0.063.
- **From 40 per cell onward,** it met both criteria in both variance
  conditions.
- **The threshold marks where Wald inference becomes approximately adequate,
  not exact.** At 40 and 50 per cell its mean coverage was still 0.943 and
  0.945, and about a quarter of scenario-by-contrast coverage values fell
  below 0.940.

**The small-sample option was calibrated at every cell size.**

- **Calibration.** Mean coverage was between 0.949 and 0.950, and Type I
  error between 0.050 and 0.053. Model standard errors were within 0.2% of
  the empirical standard errors on average and within 5% in every scenario.
- **Misses match Monte Carlo error.** 14 of 336 coverage values fell outside
  0.940 to 0.960, against 13.5 expected by chance. 3 of 42 sensitization
  Type I error rates fell outside 0.040 to 0.060, the most extreme at 0.0625.
- **Lowest coverage.** 0.931, for the ATE with 6 participants per cell and
  unequal residual variances.

**The unified GLM with HC3 is conservative in small samples.**

- **Over-coverage.** Its intervals over-covered at 10 or fewer participants
  per cell, with mean coverage of 0.961 at 6 and 0.958 at 10.
- **Standard errors.** They overestimated variability by 14.0% at 6 per cell
  and 7.1% at 10.
- **Type I error.** The sensitization test's Type I error was 0.034 at 6 per
  cell.
- **Near nominal from 20 per cell.** Mean coverage was 0.951 to 0.954, and
  its lowest coverage in any scenario was 0.938.

Unequal residual standard deviations between arms (ratio 1.5) changed mean
coverage by less than 0.01 for every method at every cell size.

**Practical guidance from this validation:**

- With fewer than 40 participants in the smallest cell, use
  `fit_solomon_ml(inference = "satterthwaite")` or `fit_solomon_glm()`
  rather than the default Wald intervals.
- The small-sample option can be used at any cell size; with 100
  participants per cell its results and those of Wald inference are close.
- `fit_solomon_glm()` with HC3 errs on the conservative side with 10 or
  fewer participants per cell.

**Not covered:** non-normal errors, clustered designs, missing posttest
scores, and incidental pretest missingness.

## Reproducing these results

The simulation script is `vignettes/articles/ml-validation/ml-simulation.R`
in the package repository. It was run with R `r run$r_version`, base seed
`r run$base_seed`, and `r format(run$nsim, big.mark = ",")` replications per
scenario.

The run used the development version of the package that implemented the
small-sample option, based on commit `4e1d820`; `run-information.csv` records
this. In that version, the estimation and inference code of
`fit_solomon_ml()` and `fit_solomon_glm()` is identical to the code committed
with this article. Later changes affected only documentation and the warning
threshold, and the simulation suppresses warnings.

## References

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*,
2074-2102.

Satterthwaite, F. E. (1946). An approximate distribution of estimates of
variance components. *Biometrics Bulletin, 2*, 110-114.

van Engelenburg, G. (1999). *Statistical analysis for the Solomon
four-group design* (Research Report 99-06). University of Twente.

Welch, B. L. (1947). The generalization of "Student's" problem when several
different population variances are involved. *Biometrika, 34*, 28-35.
