---
title: "Validating fit_solomon_ml(): A Simulation Study"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      fig.width = 8, fig.height = 5)
library(ggplot2)

performance <- utils::read.csv("ml-validation/performance.csv")
run <- utils::read.csv("ml-validation/run-information.csv")

contrast_levels <- c(
  "ATE (avg over pretest)",
  "Pretest x Treatment",
  "Treatment | pretested",
  "Treatment | unpretested"
)
method_levels <- c(
  "ML (Wald z)",
  "GLM HC3 (t)",
  "Separate OLS, Welch-Satterthwaite (exploratory)"
)

performance$contrast <- factor(performance$contrast, contrast_levels)
performance$method <- factor(performance$method, method_levels)

# Tolerances declared in the protocol before any results were examined.
performance$bias_ok <- abs(performance$bias) <= 2 * performance$mcse_bias
performance$coverage_ok <- performance$coverage >= 0.94 & performance$coverage <= 0.96
performance$se_ok <- abs(performance$relative_se_error) <= 5
performance$type1_ok <- with(
  performance,
  ifelse(contrast == "Pretest x Treatment" & sensitization == 0,
         rejection >= 0.04 & rejection <= 0.06, TRUE)
)
performance$failures_ok <- performance$n_failed == 0
performance$all_ok <- with(
  performance,
  bias_ok & coverage_ok & se_ok & type1_ok & failures_ok
)

wide_by_cell_size <- function(data, value, fmt) {
  data <- data[, c("method", "n_cell", value)]
  wide <- reshape(data, idvar = "method", timevar = "n_cell", direction = "wide")
  wide <- wide[order(wide$method), ]
  cells <- sort(unique(data$n_cell))
  out <- data.frame(Method = as.character(wide$method), check.names = FALSE)
  for (n in cells) {
    out[[paste("n =", n)]] <- fmt(wide[[paste(value, n, sep = ".")]])
  }
  out
}
```

This article reports the simulation validation of `fit_solomon_ml()`
requested in [issue #10](https://github.com/JUhalt/solomonR/issues/10). The
aims, data-generating mechanisms, estimands, methods, performance measures,
and tolerances were posted to that issue before any results were examined,
following the ADEMP structure of Morris, White and Crowther (2019).

## Design

**Aims.** Check whether `fit_solomon_ml()` recovers the four Solomon
contrasts, whether its standard errors are calibrated, and whether its 95%
Wald intervals reach nominal coverage. It is compared with the unified GLM
using HC3 standard errors and t reference distributions, and with an
exploratory small-sample counterpart: separate least-squares fits in each
pretest condition, with Welch-Satterthwaite degrees of freedom for contrasts
that combine them (Satterthwaite, 1946; Welch, 1947).

**Data-generating mechanisms.** A full factorial of 60 scenarios crossing
participants per cell (6, 10, 20, 50, 100), sensitization (0, 0.3),
pretest-posttest correlation (0, 0.5, 0.8), and the residual standard
deviation of treated relative to control participants (1, 1.5). Every
participant has a latent baseline that only pretested participants observe,
so structural pretest absence is present in every scenario. The treatment
effect is 0.5 among unpretested participants and 0.5 plus the sensitization
among pretested participants.

**Performance measures.** Bias, empirical and model-based standard errors,
coverage of 95% intervals, the rejection rate at alpha = .05, and fit
failures, each with its Monte Carlo standard error (MCSE), from
`r format(run$nsim, big.mark = ",")` replications per scenario.

**Tolerances.** Bias within 2 MCSE of zero; coverage between 0.940 and
0.960; model standard errors within 5% of the empirical standard error; Type
I error for the sensitization test between 0.040 and 0.060 when there is no
sensitization; and no fit failures.

## Fit failures

```{r failures}
failures <- aggregate(
  n_failed ~ method,
  data = performance[performance$contrast == contrast_levels[1], ],
  FUN = sum
)
names(failures) <- c("Method", "Failed fits across all scenarios")
knitr::kable(failures)
```

## Coverage

```{r coverage-plot, fig.height = 6}
coverage <- aggregate(
  coverage ~ method + contrast + n_cell + sd_ratio,
  data = performance,
  FUN = mean
)
coverage$sd_ratio <- factor(paste("SD ratio", coverage$sd_ratio))

ggplot(coverage, aes(n_cell, coverage, colour = method)) +
  annotate("rect", xmin = 0, xmax = Inf, ymin = 0.94, ymax = 0.96, alpha = 0.15) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = c(6, 10, 20, 50, 100)) +
  facet_grid(sd_ratio ~ contrast, labeller = label_wrap_gen(width = 18)) +
  labs(x = "Participants per cell (log scale)",
       y = "Coverage of 95% intervals", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

Coverage is averaged over the correlation and sensitization conditions; the
shaded band is the pre-declared tolerance. Scenario-level values and their
MCSEs are in `ml-validation/performance.csv`.

```{r coverage-tables}
mean_coverage <- aggregate(coverage ~ method + n_cell, data = performance, FUN = mean)
knitr::kable(
  wide_by_cell_size(mean_coverage, "coverage", function(x) sprintf("%.3f", x)),
  caption = "Mean coverage across scenarios and contrasts"
)

lowest_coverage <- aggregate(coverage ~ method + n_cell, data = performance, FUN = min)
knitr::kable(
  wide_by_cell_size(lowest_coverage, "coverage", function(x) sprintf("%.3f", x)),
  caption = "Lowest coverage across scenarios and contrasts"
)
```

## Bias

```{r bias}
bias_summary <- data.frame(
  Method = method_levels,
  `Largest absolute bias` = sprintf(
    "%.3f", tapply(abs(performance$bias), performance$method, max)[method_levels]
  ),
  `Estimates beyond 2 MCSE` = sprintf(
    "%d of %d",
    as.integer(tapply(!performance$bias_ok, performance$method, sum)[method_levels]),
    as.integer(table(performance$method)[method_levels])
  ),
  check.names = FALSE
)
knitr::kable(bias_summary)
```

The three methods produce identical point estimates, so their bias is
identical. With 2 MCSE as the tolerance, roughly 5% of scenario-by-contrast
estimates are expected to fall outside it by chance even for an unbiased
estimator.

## Standard error calibration

```{r se}
relative_se <- aggregate(relative_se_error ~ method + n_cell, data = performance, FUN = mean)
knitr::kable(
  wide_by_cell_size(relative_se, "relative_se_error", function(x) sprintf("%+.1f%%", x)),
  caption = "Mean relative error of model-based standard errors"
)
```

## Type I error for sensitization

```{r type1}
null <- performance[performance$contrast == "Pretest x Treatment" & performance$sensitization == 0, ]
type1 <- aggregate(rejection ~ method + n_cell, data = null, FUN = mean)
knitr::kable(
  wide_by_cell_size(type1, "rejection", function(x) sprintf("%.3f", x)),
  caption = "Rejection rate of the Pretest x Treatment test with no sensitization (nominal .05)"
)
```

## Scenarios meeting every tolerance

```{r tolerance}
scenario_ok <- aggregate(all_ok ~ method + scenario + n_cell, data = performance, FUN = all)
meeting <- aggregate(all_ok ~ method + n_cell, data = scenario_ok, FUN = sum)
totals <- aggregate(all_ok ~ method + n_cell, data = scenario_ok, FUN = length)
meeting$all_ok <- sprintf("%d of %d", meeting$all_ok, totals$all_ok)
knitr::kable(
  wide_by_cell_size(meeting, "all_ok", identity),
  caption = "Scenarios in which all four contrasts meet every tolerance"
)
```

## Conclusions

**Estimand recovery is validated.** All three methods produce the same point
estimates. The largest absolute bias was 0.060 standard-deviation units, at 6
participants per cell, and 18 of 240 scenario-by-contrast estimates (7.5%)
fell outside 2 MCSE, close to the chance expectation. No fit failed.

**Inference from `fit_solomon_ml()` is not validated below about 50
participants per cell.** Its 95% Wald intervals had mean coverage of 0.893
with 6 participants per cell, 0.922 with 10, and 0.936 with 20, reaching
0.945 to 0.947 only at 50 to 100. Its standard errors were 10.8% too small at
6 per cell and 5.8% too small at 10, and the sensitization test rejected a
true null 9.8% of the time at 6 per cell and 7.6% at 10. It met every
tolerance in none of the scenarios with 20 or fewer participants per cell,
in 8 of 12 at 50, and in 10 of 12 at 100.

**The unified GLM with HC3 is conservative in small samples.** Its intervals
over-covered at 10 or fewer participants per cell (mean coverage 0.962 at 6
and 0.959 at 10), its standard errors overestimated variability, and the
sensitization test's Type I error was 0.032 at 6 per cell. It was close to
nominal from 20 participants per cell onward and did not under-cover by more
than about one percentage point in any scenario.

**The exploratory Welch-Satterthwaite counterpart was calibrated at every
cell size**, with mean coverage between 0.949 and 0.951, model standard
errors within 1% of the empirical standard errors on average, and Type I
error between 0.048 and 0.051. Because it shares the maximum-likelihood point
estimates, a small-sample inference option for `fit_solomon_ml()` is proposed
in [issue #22](https://github.com/JUhalt/solomonR/issues/22).

Unequal residual standard deviations between arms (ratio 1.5) changed mean
coverage by less than 0.01 for every method.

**Practical guidance from this validation:** with fewer than about 50
participants per cell, do not rely on the intervals or p-values from
`fit_solomon_ml()`; use `fit_solomon_glm()`, whose intervals err on the
conservative side.

**Not covered:** non-normal errors, clustered designs, missing posttest
scores, and incidental pretest missingness.

## Reproducing these results

The simulation script is `vignettes/articles/ml-validation/ml-simulation.R`
in the package repository. It was run at commit `r run$git_commit` with R
`r run$r_version`, base seed `r run$base_seed`, and
`r format(run$nsim, big.mark = ",")` replications per scenario.

## References

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*,
2074-2102.

Satterthwaite, F. E. (1946). An approximate distribution of estimates of
variance components. *Biometrics Bulletin, 2*, 110-114.

van Engelenburg, G. (1999). *Statistical analysis for the Solomon
four-group design* (Research Report 99-06). University of Twente.

Welch, B. L. (1947). The generalization of "Student's" problem when several
different population variances are involved. *Biometrika, 34*, 28-35.
