---
title: "Longitudinal Designs: Validating the Repeated-Measures Analysis"
description: >
  Simulation study of fit_solomon_mmrm(): the Solomon contrasts at each of
  three posttest occasions under dropout that is missing at random.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

p <- utils::read.csv("mmrm-validation/performance.csv", stringsAsFactors = FALSE)
info <- utils::read.csv("mmrm-validation/run-information.csv", stringsAsFactors = FALSE)
run <- stats::setNames(as.list(info$value), info$item)
reps <- as.integer(run$replications)

# Tolerances declared on issue #57 before any run.
p$null <- abs(p$true_value) < 1e-8
p$bias_ok <- abs(p$bias) <= 2 * p$bias_mcse
p$cov_ok <- p$coverage >= 0.94 & p$coverage <= 0.96
p$se_ok <- abs(p$relerr) <= 10
p$t1_ok <- ifelse(p$null, p$rejection >= 0.04 & p$rejection <= 0.06, TRUE)
p$all_ok <- p$bias_ok & p$cov_ok & p$se_ok & p$t1_ok
p$type <- sub("^[0-9]+: |^3 vs 1: ", "", p$estimand)
p$occasion <- sub(":.*$", "", p$estimand)

kr <- p[p$method == "mmrm_kr", ]
sat <- p[p$method == "mmrm_satterthwaite", ]
nor <- p[p$method == "mmrm_normal", ]
cc <- p[p$method == "complete_case", ]
sh <- p[p$method == "mmrm_shared_kr", ]

pct <- function(x) sprintf("%.1f%%", 100 * x)
f3 <- function(x) sprintf("%.3f", x)
rng <- function(x, d = 3) sprintf(paste0("%.", d, "f to %.", d, "f"), min(x), max(x))
count <- function(ok) sprintf("%d of %d", sum(ok), length(ok))
method_label <- c(mmrm_kr = "M1: MMRM, Kenward-Roger", mmrm_satterthwaite = "M2: MMRM, Satterthwaite",
                  mmrm_normal = "M3: MMRM, normal reference", complete_case = "M4: complete case per occasion",
                  mmrm_shared_kr = "M5: MMRM, one shared covariance")
```

This article reports the simulation validation of `fit_solomon_mmrm()`, the
analysis of Solomon designs with several posttest occasions. The aims,
data-generating mechanisms, estimands, methods, performance measures,
tolerances, and decision rules were posted on
[issue #57](https://github.com/JUhalt/solomonR/issues/57) before any code,
following the ADEMP structure of Morris et al. (2019). Two amendments were
posted before any run: the covariance is estimated separately for pretested
and unpretested participants, with a shared-covariance model added as a
comparator; and the Kenward-Roger adjustment uses the element-wise
parameterization that reproduces SAS. The results come from package commit
`r substr(run$commit, 1, 7)`, with `r reps` replications per scenario, mmrm
`r run$mmrm_version`.

## Design

**Outcomes.** A latent baseline and three posttests, multivariate normal,
with posttest standard deviations 1.0, 1.2, and 1.4 and correlation
0.6^|lag|. The pretest is the baseline, observed in the pretested groups.
The pretest effect is 0.2 at every occasion, and the treatment effect among
unpretested participants is 0.5, 0.4, and 0.3.

**Dropout.** Monotone and missing at random. Before the first posttest, 5%
of the control groups and 10% of the treatment groups are missing
completely at random. After each posttest, a participant drops out with a
probability that doubles in odds per standard deviation of that posttest,
at a rate per occasion of 8% (control) and 12% (treatment) for moderate
dropout, or 20% and 30% for heavy dropout.

**Scenarios.** `r length(unique(p$scenario))` in all: 30, 60, or 120 per
group; moderate or heavy dropout; and no sensitization, or sensitization of
0.3, 0.15, and 0 at the three occasions ("fading").

**Estimands.** The four Solomon contrasts at each occasion and the change in
the Pretest x Treatment contrast from the first occasion to the last: 13 in
all, defined on the means of all randomized participants.

**Methods.**

- **M1:** `fit_solomon_mmrm()`, Kenward-Roger degrees of freedom.
- **M2:** `fit_solomon_mmrm(df = "satterthwaite")`.
- **M3:** M2's estimates and standard errors with a normal reference
  distribution (Fitzmaurice et al., 2011, p. 101).
- **M4:** `fit_solomon_glm()` at each occasion on the participants still
  observed, whose point estimates equal those of unweighted generalized
  estimating equations with an independence working correlation.
- **M5:** the same MMRM with one unstructured covariance shared by all four
  groups, with Kenward-Roger degrees of freedom.

**Tolerances.** Bias within 2 Monte Carlo standard errors (MCSE); coverage
of 95% intervals from 0.940 to 0.960; model standard error within 10% of the
empirical standard error; and, where the true value is zero, a rejection
rate from 0.040 to 0.060.

## Failures and convergence

`r if (sum(p$failures) == 0 && all(p$errors == 0)) "No analysis failed or returned an error in any replication." else sprintf("%d analyses failed.", sum(p$failures))`
The unstructured covariance converged in every replication, so the fallback
structures were never needed (**decision rule 4**).

## Choosing the degrees of freedom

**Decision rule 1** made the default the approximation, Kenward-Roger or
Satterthwaite, that met every tolerance in more cells. It had to meet them
in at least 90% of the `r nrow(kr)` estimand-by-scenario cells and in every
cell with 60 or more per group.

```{r rule1}
rows <- lapply(c("mmrm_kr", "mmrm_satterthwaite", "mmrm_normal", "mmrm_shared_kr"), function(m) {
  d <- p[p$method == m, ]
  data.frame(Method = method_label[m],
             `Every tolerance` = sprintf("%d of %d (%s)", sum(d$all_ok), nrow(d), pct(mean(d$all_ok))),
             `60 or more per group` = count(d$all_ok[d$n >= 60]),
             Coverage = rng(d$coverage),
             `Type I error` = rng(d$rejection[d$null], 4),
             check.names = FALSE)
})
knitr::kable(do.call(rbind, rows), row.names = FALSE, align = "lrrrr",
             caption = "Cells meeting every tolerance, and the range of coverage and of Type I error where the true value is zero.")
```

Both approximations met every tolerance in `r pct(mean(kr$all_ok))` of cells,
but neither did so in every cell with 60 or more per group, so **neither
qualified under the rule**. The default stays Kenward-Roger, the
approximation Mallinckrodt et al. (2008, p. 312) specify and the one the
protocol's tie-break prefers when the two are close; here they differed by
`r abs(sum(kr$all_ok[kr$n >= 60]) - sum(sat$all_ok[sat$n >= 60]))` cell. The
Kenward-Roger shortfalls were:

```{r shortfalls}
s <- kr[!kr$all_ok, ]
knitr::kable(data.frame(
  `Per group` = s$n, Dropout = s$dropout, Sensitization = s$sensitization, Estimand = s$estimand,
  `Bias (SD)` = sprintf("%.4f", s$bias), `Bias / MCSE` = sprintf("%.1f", s$bias / s$bias_mcse),
  Coverage = sprintf("%.4f", s$coverage),
  `Type I` = ifelse(s$null, sprintf("%.4f", s$rejection), ""),
  check.names = FALSE
), row.names = FALSE, align = "rlllrrrr", caption = "Kenward-Roger cells outside a tolerance.")
```

They are scattered across group sizes, dropout levels, occasions, and
contrasts. The mean coverage was
`r paste(sprintf("%.4f", tapply(kr$coverage, kr$n, mean)), collapse = ", ")`
at 30, 60, and 120 per group.

```{r calibrated}
R <- reps
# Applied exactly as the coverage tolerance is: to the observed proportion.
x <- 0:R
p_cov <- sum(stats::dbinom(x, R, 0.95)[x / R >= 0.94 & x / R <= 0.96])
p_bias <- 2 * stats::pt(2, R - 1) - 1
p_both <- p_cov * p_bias
```

**A note on the rule.** Monte Carlo error alone makes the rule stricter than
any method can reliably meet. With `r R` replications, a method whose
intervals cover exactly 95% and whose estimates are exactly unbiased falls
inside the coverage tolerance with probability `r f3(p_cov)` in each cell and
inside the bias tolerance with probability `r f3(p_bias)`, so it meets both
in about `r pct(p_both)` of cells. The chance that it meets them in all
`r sum(kr$n >= 60)` cells with 60 or more per group is about
`r sprintf("%.4f", p_both^sum(kr$n >= 60))`. Both approximations' pass rates
are what a correctly calibrated method would show. The verdict stands as the
pre-specified rule gave it; future protocols in the package will state
tolerances that allow for Monte Carlo error across many cells.

**The normal reference distribution** (M3), which ignores the uncertainty in
the estimated covariance, had mean coverage
`r f3(mean(nor$coverage[nor$n == 30]))` at 30 per group (as low as
`r f3(min(nor$coverage[nor$n == 30]))`) and Type I error up to
`r sprintf("%.4f", max(nor$rejection[nor$null]))`. The t reference matters
mainly in small groups, as Fitzmaurice et al. (2011, p. 101) caution.

## Complete-case analysis under dropout

```{r complete-case}
last <- p[p$occasion == "3" & p$type != "Pretest x Treatment" &
            p$method %in% c("complete_case", "mmrm_kr"), ]
tab <- aggregate(cbind(bias, coverage) ~ method + dropout, last, function(x) c(mean = mean(x), min = min(x)))
knitr::kable(data.frame(
  Method = method_label[tab$method], Dropout = tab$dropout,
  `Mean bias (SD)` = sprintf("%.3f", tab$bias[, "mean"]),
  `Mean coverage` = sprintf("%.3f", tab$coverage[, "mean"]),
  `Lowest coverage` = sprintf("%.3f", tab$coverage[, "min"]),
  check.names = FALSE
), row.names = FALSE, align = "llrrr",
caption = "The three treatment contrasts at the last occasion, over group sizes and sensitization patterns.")
```

**Decision rule 3** was descriptive. The per-occasion complete-case analysis
(M4) was unbiased at the first occasion, where missingness was completely at
random. After dropout that depended on the last posttest, it underestimated
the treatment contrasts at the last occasion, and its coverage fell as the
groups grew, because the bias stays while the standard error shrinks
(`r f3(min(cc$coverage[cc$n == 120]))` at 120 per group). The MMRM, which uses
every observed posttest, was unbiased in the same cells. The sensitization
contrast was nearly unbiased even under M4, because dropout followed the
same rule in the pretested and unpretested groups; a dropout process that
differed between them would bias it too.

## One covariance for all four groups

The MMRM with one covariance shared by all four groups (M5) was, like M1,
nearly unbiased (bias within 2 MCSE in `r count(sh$bias_ok)` cells), but its
standard errors were wrong:

```{r shared}
sh$contrast <- sh$type
tab <- aggregate(cbind(relerr, coverage) ~ contrast, sh, function(x) c(min = min(x), max = max(x)))
knitr::kable(data.frame(
  Contrast = tab$contrast,
  `Model SE vs empirical SE` = sprintf("%+.1f%% to %+.1f%%", tab$relerr[, "min"], tab$relerr[, "max"]),
  Coverage = sprintf("%.3f to %.3f", tab$coverage[, "min"], tab$coverage[, "max"]),
  check.names = FALSE
), row.names = FALSE, align = "lrr", caption = "M5 over all scenarios and occasions.")
```

A shared matrix averages the conditional covariance of the pretested groups
with the larger marginal covariance of the unpretested groups, so it
understates the standard errors of the unpretested contrasts and overstates
those of the pretested ones. Estimating the covariance separately by
pretest condition, as `fit_solomon_mmrm()` does, is necessary.

## What this means for users

- Under dropout that is missing at random, `fit_solomon_mmrm()` gave
  unbiased contrasts at every occasion, with coverage from
  `r rng(kr$coverage)` and Type I error from `r rng(kr$rejection[kr$null], 4)`,
  from 30 to 120 per group. Kenward-Roger and Satterthwaite degrees of
  freedom performed alike.
- Per-occasion analyses of the participants still observed can be biased
  when dropout depends on earlier scores.
- The function remains experimental, because decision rule 1 was not met.

## Reproducing the study

From the package root,
`Rscript vignettes/articles/mmrm-validation/mmrm-simulation.R` writes
`performance.csv` and `run-information.csv`. The run used R
`r sub("R version ([0-9.]+).*", "\\1", run$R_version)`, mmrm
`r run$mmrm_version`, and `r run$workers` workers, from `r run$started` to
`r run$finished`.

## References

Fitzmaurice, G. M., Laird, N. M., & Ware, J. H. (2011). *Applied longitudinal
analysis* (2nd ed.). Wiley. https://doi.org/10.1002/9781119513469

Mallinckrodt, C. H., Lane, P. W., Schnell, D., Peng, Y., & Mancuso, J. P.
(2008). Recommendations for the primary analysis of continuous endpoints in
longitudinal clinical trials. *Drug Information Journal, 42*(4), 303–319.
https://doi.org/10.1177/009286150804200402

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086
