---
title: "Planning a Solomon Study"
description: >
  Choosing sample sizes for a Solomon four-group study with plan_solomon(),
  power_solomon(), and plot_power_solomon(), using planning values from the
  published literature.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

A Solomon four-group study answers three questions: does the treatment
work, does pretesting change the outcome, and does pretesting change the
treatment's effect (sensitization)? This article shows how to choose a
sample size for each, and where the planning values come from.

## What you have to specify

`plan_solomon()` finds the smallest design whose power reaches a target for
each Solomon estimand. It needs five planning values:

| Argument | Meaning |
|---|---|
| `delta` | The treatment effect among unpretested participants |
| `sens` | Sensitization: the additional treatment effect among pretested participants |
| `rho` | The pretest–posttest correlation among pretested participants |
| `sigma` | The posttest standard deviation; with `sigma = 1`, effects are in SD units |
| `allocation` | The relative sizes of the four groups |

The treatment effect among pretested participants is then `delta + sens`,
and the average treatment effect is `delta + sens / 2`. A pretest main
effect, which shifts both pretested groups equally, does not enter the power
of any treatment contrast: the analysis model absorbs it with the pretest
indicator.

## Planning values from the literature

**Pretest effects.** The meta-analysis of Willson and Putnam (1982)
combined 134 outcomes from randomized studies of pretest effects. Pretesting
raised posttest scores by .22 standard deviations on average, by .43 for
cognitive outcomes, .29 for attitude outcomes, and .48 for personality
outcomes, with other outcomes about zero (p. 249). Pretesting matters, then,
even though a main effect of pretesting does not change the sample size
needed for the treatment contrasts.

**Sensitization.** In the same meta-analysis, the pretest effect averaged
.13 among treated groups and .39 among untreated groups, a difference that
was not statistically significant (p. 252). Read as a Solomon interaction,
that is a sensitization effect of about −.26 standard deviations: the
treatment worked somewhat less well among pretested participants. Two
cautions apply. The figure averages over very different studies and
outcomes, and the effect distributions were skewed (p. 252). And a later
systematic review of Solomon studies with behavioral outcomes found only
sparse evidence of interactions, with too few high-quality studies to settle
the question (McCambridge et al., 2011, p. 1). A sensitization of about a
quarter of a standard deviation is therefore a reasonable scenario to plan
for, not an established value.

**The pretest–posttest correlation.** These sources do not supply `rho`;
it depends on the measure and the interval between pretest and posttest.
Use the test–retest reliability of your measure over a similar interval, and
check how much the plan depends on it (below).

## A plan for all three questions

With a treatment effect of 0.30 standard deviations, sensitization of 0.26,
and a pretest–posttest correlation of .5 (power depends on the size of
sensitization, not its sign, so the magnitude from Willson and Putnam is
used):

```{r plan}
plan <- plan_solomon(power = 0.80, delta = 0.30, sens = 0.26, rho = 0.5)
plan[, c("estimand", "true_effect", "n1", "n2", "n3", "n4", "total_n", "power")]
```

The average treatment effect needs `r plan$total_n[plan$estimand == "ATE (avg over pretest)"]`
participants; sensitization needs
`r plan$total_n[plan$estimand == "Pretest x Treatment"]`. This is not a
quirk of these values. The sensitization contrast is the difference between
the two simple treatment effects and the average treatment effect is their
average, so sensitization has exactly four times the sampling variance of
the average treatment effect, whatever the correlation and allocation
(`?plan_solomon`). A study powered only for the treatment effect is usually
underpowered for the question the Solomon design exists to answer.

`plot_power_solomon()` shows the power curves behind the plan:

```{r power-plot, fig.width = 7, fig.height = 4.5, fig.alt = "Power curves for each Solomon estimand across sample sizes per group, for a treatment effect of 0.30 and sensitization of 0.26 standard deviations."}
plot_power_solomon(delta = 0.30, sens = 0.26, rho = 0.5, n = seq(20, 420, by = 40))
```

## How much the plan depends on the correlation

```{r rho}
rhos <- c(0.3, 0.5, 0.7)
totals <- sapply(rhos, function(r) {
  p <- plan_solomon(power = 0.80, delta = 0.30, sens = 0.26, rho = r)
  stats::setNames(p$total_n, p$estimand)
})
colnames(totals) <- paste("rho =", rhos)
totals
```

A higher correlation helps the contrasts that use the pretest (the
pretested effect, the average, and sensitization) and leaves the unpretested
effect unchanged, because the unpretested groups have no pretest.

## Allocation

When pretesting is the expensive part of a study, fewer participants can be
pretested. `allocation` gives the relative group sizes, in the order
pretested treated, pretested control, unpretested treated, unpretested
control:

```{r allocation}
plan_2to1 <- plan_solomon(power = 0.80, delta = 0.30, sens = 0.26, rho = 0.5,
                          allocation = c(1, 1, 2, 2))
plan_2to1[, c("estimand", "n1", "n2", "n3", "n4", "total_n")]
```

```{r allocation-cost, include = FALSE}
pretests <- function(p, est) sum(unlist(p[p$estimand == est, c("n1", "n2")]))
ate <- "ATE (avg over pretest)"
sens_label <- "Pretest x Treatment"
```

If pretests are the costly part, count them. With twice as many unpretested
as pretested participants, the average treatment effect needs
`r pretests(plan_2to1, ate)` pretested participants instead of
`r pretests(plan, ate)`, and sensitization `r pretests(plan_2to1, sens_label)`
instead of `r pretests(plan, sens_label)`, at the price of more participants
in total (`r plan_2to1$total_n[plan_2to1$estimand == ate]` instead of
`r plan$total_n[plan$estimand == ate]`, and
`r plan_2to1$total_n[plan_2to1$estimand == sens_label]` instead of
`r plan$total_n[plan$estimand == sens_label]`).

## Small designs

The analytic plans use normal-theory power. The package's own analysis
uses HC3 standard errors, which are conservative with 20 or fewer
participants per group, so for small designs `method = "simulation"` plans
for the test the package actually runs:

```{r simulation}
simulated <- plan_solomon(power = 0.80, delta = 0.8, rho = 0.5, estimand = "unpretested",
                          method = "simulation", sims = 500, seed = 49)
analytic <- plan_solomon(power = 0.80, delta = 0.8, rho = 0.5, estimand = "unpretested")
rbind(analytic, simulated)[, c("basis", "n1", "total_n", "power", "mcse")]
```

For a large effect in small groups, the simulated plan needs
`r simulated$n1` participants per group against `r analytic$n1` from the
normal-theory calculation. The
re-simulation study behind this advice is reported in
[Designs from plan_solomon()](plan-validation.html), and the rejection rates
of `power_solomon()` in [Validating power_solomon()](power-validation.html);
both are summarized on the [Validation Evidence](validation-evidence.html)
page.

## The historical test sequence

These plans are for the package's recommended analysis. Planning for the
historical sequence of Tests A–I, and the power cost of the alpha
allocations Sawilowsky (1996) proposed to control its experiment-wise error,
will be added with
[issue #51](https://github.com/JUhalt/solomonR/issues/51).

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References
McCambridge, J., Butor-Bhavsar, K., Witton, J., & Elbourne, D. (2011). Can
research assessments themselves cause bias in behaviour change trials? A
systematic review of evidence from Solomon 4-group studies. *PLoS ONE,
6*(10), Article e25223. https://doi.org/10.1371/journal.pone.0025223

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
http://digitalcommons.wayne.edu/coe_tbf/29

Willson, V. L., & Putnam, R. R. (1982). A meta-analysis of pretest
sensitization effects in experimental design. *American Educational Research
Journal, 19*(2), 249–258. https://doi.org/10.3102/00028312019002249
