---
title: "Checking plan_solomon(): Re-simulating Planned Designs"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

performance <- utils::read.csv("plan-validation/performance.csv", stringsAsFactors = FALSE)
run <- utils::read.csv("plan-validation/run-information.csv", stringsAsFactors = FALSE)

num <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
performance$size <- cut(performance$smallest_cell, c(0, 29, 99, Inf),
                        labels = c("under 30", "30 to 99", "100 or more"))
performance$within <- abs(performance$difference) <= 0.02
```

`plan_solomon()` returns the smallest design whose normal-theory power
reaches a target. This check asks whether the package's own test, run on
data from the returned design, actually reaches that target. The criterion
was posted to [issue #24](https://github.com/JUhalt/solomonR/issues/24)
before the check was run: from 30 participants per cell, re-simulated power
should be within 0.02 of the target.

## Design

**Plans.** Each design came from `plan_solomon()` with a target of 80%
power, for every estimand with a nonzero effect. That gave
`r run$plans` plans from `r run$designs` designs, crossing these settings:

- **Treatment effect among unpretested participants:** 0.3 or 0.6.
- **Sensitization:** 0 or 0.3.
- **Pretest-posttest correlation:** 0 or 0.5.
- **Allocation:** equal, or 1:1:2:2 (twice as many unpretested
  participants).

**Re-simulation.** Each returned design was simulated with `power_solomon()`
at `r format(run$sims, big.mark = ",")` replications, a Monte Carlo standard
error (MCSE) of about 0.009 at power 0.80. The rejection rate of the unified
GLM test with HC3 standard errors, the package default, was compared with
the target. Fit failures were counted separately.

## Results

```{r by-size}
by_size <- do.call(rbind, lapply(split(performance, performance$size), function(g) {
  if (nrow(g) == 0) return(NULL)
  data.frame(
    `Smallest cell` = as.character(g$size[1]),
    Plans = nrow(g),
    `Mean difference from target` = sprintf("%+.3f", mean(g$difference)),
    `Largest shortfall` = sprintf("%+.3f", min(g$difference)),
    `Within 0.02` = sum(g$within),
    check.names = FALSE
  )
}))
knitr::kable(by_size, row.names = FALSE)
```

```{r misses}
large <- performance[performance$smallest_cell >= 30, ]
misses <- large[!large$within, ]
```

- **Criterion met.** From 30 participants per cell, `r sum(large$within)` of
  `r nrow(large)` plans reached the target within 0.02.
  `r if (nrow(misses)) sprintf("The %s outside were above the target, at %s, which is consistent with simulation error.", if (nrow(misses) == 1) "one plan" else paste(nrow(misses), "plans"), paste(num(misses$simulated_power), collapse = " and ")) else ""`
- **Fit failures.** `r sum(performance$failures)` across all re-simulations.
- **Small designs.** Below 30 participants per cell, analytic plans run
  slightly short, because the HC3 standard errors the package uses are
  conservative there. This matches the validation of `power_solomon()`. For
  such designs, `plan_solomon(method = "simulation")` plans for the package's
  own test directly.

```{r sensitization}
ate <- performance[performance$estimand == "ATE (avg over pretest)" & performance$sens == 0, ]
sens <- performance[performance$estimand == "Pretest x Treatment", ]
pairs <- merge(
  sens[sens$delta == 0.3, c("rho", "allocation", "total_n")],
  ate[ate$delta == 0.3, c("rho", "allocation", "total_n")],
  by = c("rho", "allocation"), suffixes = c("_sensitization", "_ate")
)
pairs$ratio <- pairs$total_n_sensitization / pairs$total_n_ate
```

**The four-times-variance result, in practice.** Detecting sensitization of
0.3 standard deviations needed between `r num(min(pairs$ratio), 2)` and
`r num(max(pairs$ratio), 2)` times as many participants as detecting an
average treatment effect of the same size, in line with the exact
four-to-one variance ratio stated on `?plan_solomon`.

## All plans

```{r all}
knitr::kable(
  data.frame(
    Estimand = performance$estimand,
    Design = sprintf("delta %s, sens %s, rho %s, %s", performance$delta,
                     performance$sens, performance$rho, performance$allocation),
    `Cell sizes` = sprintf("%d/%d/%d/%d", performance$n1, performance$n2,
                           performance$n3, performance$n4),
    `Analytic power` = num(performance$analytic_power),
    `Simulated power` = num(performance$simulated_power),
    MCSE = num(performance$mcse, 4),
    check.names = FALSE
  )[order(performance$smallest_cell), ],
  row.names = FALSE
)
```

## Reproducing this check

The script is `vignettes/articles/plan-validation/plan-resimulation.R` in
the package repository. It was run at `r run$git_commit` with R
`r run$r_version` and base seed `r run$base_seed`. The results also feed the
shared tables on the [Validation Evidence](validation-evidence.html) page.
