---
title: "Validating power_solomon(): A Simulation Study"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      fig.width = 8, fig.height = 5)
library(ggplot2)

performance <- utils::read.csv("power-validation/performance.csv")
run <- utils::read.csv("power-validation/run-information.csv")
invariance <- utils::read.csv("power-validation/invariance.csv")

test_levels <- c("GLM (HC3, t)", "2x2 ANOVA interaction",
                 "Test I (Braver & Braver, 1988)")
performance$test <- factor(performance$test, test_levels)
allocation_levels <- unique(performance$allocation[order(performance$n1 + performance$n2 +
                                                         performance$n3 + performance$n4)])
performance$allocation <- factor(performance$allocation, allocation_levels)

# Tolerances declared on the issue before any results were examined.
performance$null_case <- !is.na(performance$true_effect) & performance$true_effect == 0
performance$tolerance <- pmax(0.02, 2 * performance$mcse)
performance$agrees <- !is.na(performance$analytic_power) &
  abs(performance$difference) <= performance$tolerance
performance$type1_ok <- ifelse(performance$null_case,
                               performance$power >= 0.04 & performance$power <= 0.06,
                               NA)

wide_by_allocation <- function(data, value, fmt) {
  data <- data[, c("test", "allocation", value)]
  wide <- reshape(data, idvar = "test", timevar = "allocation", direction = "wide")
  out <- data.frame(Test = as.character(wide$test), check.names = FALSE)
  for (a in levels(performance$allocation)) {
    column <- paste(value, a, sep = ".")
    if (column %in% names(wide)) out[[a]] <- fmt(wide[[column]])
  }
  out
}
```

This article reports the simulation validation of `power_solomon()` requested
in [issue #18](https://github.com/JUhalt/solomonR/issues/18). The aims,
data-generating mechanisms, estimands, methods, performance measures, and
tolerances were posted to that issue before any results were examined,
following the ADEMP structure of Morris, White and Crowther (2019). An
amendment reducing the grid to a practical runtime was also posted before the
run, and changed no tolerance.

The study drives `power_solomon()` itself rather than a parallel
implementation, so what is validated is the function users call.

## Design

**Aims.** Establish that each reported test holds its nominal Type I error
when the corresponding effect is zero, that simulated power agrees with
analytic power where an analytic result exists, and that every reported
number carries its Monte Carlo uncertainty.

**Data-generating mechanisms.** Every participant has a latent baseline drawn
from a standard normal distribution, which only pretested participants
observe, so structural pretest absence is present in every scenario. The
posttest residual standard deviation is 1 in every cell, the
pretest-posttest correlation among pretested participants is `rho`, the
treatment effect is `delta` among unpretested participants, and sensitization
adds `sens` among pretested participants.

The grid crosses:

- **allocation:** 10, 20, 30, 50, and 100 participants per cell, plus the
  unequal allocations 60/60/20/20 and 20/20/60/60;
- **`delta`:** 0, 0.3, 0.6;
- **`sens`:** 0, 0.3;
- **`rho`:** 0, 0.5, 0.8.

**Methods.** The three procedures `power_solomon()` reports: the unified GLM
with HC3 standard errors and t reference distributions, the 2x2 ANOVA
interaction, and the historical one-tailed Test I (Braver & Braver, 1988).

**Analytic benchmarks.** Normal-theory rejection probabilities computed from
the design: a two-sample t test for the unpretested contrast, an ANCOVA
comparison with residual variance reduced by the squared pretest-posttest
correlation for the pretested contrast, and Welch-Satterthwaite degrees of
freedom for the contrasts that combine pretest conditions. The 2x2 ANOVA
interaction is benchmarked with pooled variance and no pretest adjustment.
Test I combines two p-values directionally and has no closed form, so it is
reported without a benchmark.

**Performance measures.** The rejection rate of each test with its Monte
Carlo standard error, the difference from the analytic benchmark, and the
number of fit failures. Replications: `r format(run$sims_null, big.mark = ",")`
in complete-null scenarios and `r format(run$sims_alternative, big.mark = ",")`
elsewhere.

**Tolerances.** Type I error between 0.040 and 0.060 wherever the true effect
is zero; power within 0.02 of the analytic benchmark or within 2 Monte Carlo
standard errors, whichever is larger; no fit failures; and scale invariance
at a residual standard deviation of 2 with proportionally scaled effects.

## Fit failures

```{r failures}
failures <- aggregate(failures ~ test, data = performance, FUN = sum)
names(failures) <- c("Test", "Failed fits across all scenarios")
knitr::kable(failures)
```

## Type I error

```{r type1}
null_rows <- performance[performance$null_case, ]
type1 <- aggregate(power ~ test + allocation, data = null_rows, FUN = mean)
knitr::kable(
  wide_by_allocation(type1, "power", function(x) sprintf("%.3f", x)),
  caption = "Mean rejection rate where the true effect is zero (nominal .05)"
)
```

```{r type1-plot, fig.height = 4}
ggplot(null_rows, aes(allocation, power, colour = test)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.04, ymax = 0.06, alpha = 0.15) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.7) +
  labs(x = NULL, y = "Rejection rate under a true null", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

The shaded band is the pre-declared tolerance. Each point is one scenario.

## Agreement with analytic power

```{r agreement-plot, fig.height = 5}
bench <- performance[!is.na(performance$analytic_power) & !performance$null_case, ]

ggplot(bench, aes(analytic_power, power, colour = allocation)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.7) +
  facet_wrap(~ test) +
  labs(x = "Analytic power", y = "Simulated power", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r agreement-tables}
differences <- aggregate(difference ~ test + allocation, data = bench, FUN = mean)
knitr::kable(
  wide_by_allocation(differences, "difference", function(x) sprintf("%+.3f", x)),
  caption = "Mean difference between simulated and analytic power"
)

within <- aggregate(agrees ~ test + allocation, data = bench, FUN = mean)
knitr::kable(
  wide_by_allocation(within, "agrees", function(x) sprintf("%.0f%%", 100 * x)),
  caption = "Share of scenarios within the pre-declared tolerance"
)
```

## Historical Test I

```{r test-i}
test_i <- performance[performance$test == "Test I (Braver & Braver, 1988)", ]
test_i_null <- test_i[test_i$delta == 0 & test_i$sens == 0, ]
summary_i <- aggregate(power ~ allocation, data = test_i_null, FUN = mean)
names(summary_i) <- c("Allocation", "Rejection rate under the complete null")
knitr::kable(summary_i, digits = 3)
```

Test I combines two one-tailed p-values, so it is reported as a historical
procedure rather than a recommended one, and its rejection rate is not
compared with a two-sided benchmark.

## Scale invariance

```{r invariance}
knitr::kable(
  data.frame(
    Estimand = invariance$estimand,
    Test = invariance$test,
    `SD 1` = sprintf("%.3f", invariance$power_sigma_1),
    `SD 2` = sprintf("%.3f", invariance$power_sigma_2),
    Difference = sprintf("%+.3f", invariance$difference),
    check.names = FALSE
  ),
  caption = "Rejection rates with proportionally scaled effects and residual SD"
)
```

## Conclusions

**The simulator is validated.** Across 126 scenarios and 315,000
replications, no fit failed, and three independent checks agreed:

- **Analytic agreement.** The 2x2 ANOVA interaction matched its normal-theory
  benchmark in every scenario at every allocation, with mean differences of
  0.002 or less.
- **Nominal size for Test I.** Under the complete null, the historical
  one-tailed combination rejected between 4.4% and 5.6% of the time at
  alpha = .05.
- **Scale invariance.** Doubling the residual standard deviation with
  proportionally scaled effects reproduced every rejection rate exactly.

**Rejection rates from the unified GLM are conservative in small samples.**
Type I error averaged 0.041 with 10 participants per cell, rising through
0.044 at 20 and 0.047 at 30 to 0.049 at 100. Simulated power fell below the
normal-theory benchmark by 0.030 on average with 10 per cell, 0.017 with 20,
0.009 with 30, 0.005 with 50, and 0.003 with 100; the largest single
difference was 0.083, for the ATE with 10 per cell. Agreement within the
pre-declared tolerance rose from 41% of scenarios at 10 per cell to 98% at
100.

**This traces to HC3, not to the simulator.** The ANOVA arm is computed on
the same simulated data sets and matched its benchmark exactly, so the
deviation belongs to the covariance estimator rather than the data-generating
mechanism. It is the same conservatism the `fit_solomon_ml()` validation
found, where HC3 intervals over-covered at small cell sizes (mean coverage
0.961 with 6 participants per cell). The protocol registered this deviation
in advance.

**Unequal allocation behaved like equal allocation of similar size.** The
60/60/20/20 and 20/20/60/60 designs agreed with their benchmarks in 94% and
74% of scenarios, between the 20- and 50-per-cell equal allocations.

**Of 210 null scenarios, 17 fell outside the 0.040 to 0.060 band**, 15 of
them below it. The low side is the HC3 conservatism described above; the two
above the band, at 0.0605 and 0.0620, are within about two Monte Carlo
standard errors of 0.05 and are consistent with simulation noise.

**Practical guidance from this validation:**

- Power reported for the GLM tests with 20 or fewer participants per cell is
  a conservative figure; the normal-theory benchmark in this article is the
  optimistic end of the range.
- The 2x2 ANOVA interaction is the arm to compare against analytic
  expectations, but it ignores the pretest, so it is less powerful than the
  GLM whenever the pretest is informative.
- Test I remains a historical procedure. Its size is nominal under the
  complete null, which does not make the conditional sequence it belongs to
  safe; see the method guide.

**Not covered:** non-normal errors, clustered designs, binary and count
outcomes, missing posttest scores, and incidental pretest missingness.

## Reproducing these results

The simulation script is
`vignettes/articles/power-validation/power-simulation.R` in the package
repository. It was run at `r run$git_commit` with R `r run$r_version`, base
seed `r run$base_seed`, and the replication counts above.

## References

Braver, M. W., & Braver, S. L. (1988). Statistical treatment of the Solomon
four-group design: A meta-analytic approach. *Psychological Bulletin, 104*,
150-154.

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*,
2074-2102.

Solomon, R. L. (1949). An extension of control group design.
*Psychological Bulletin, 46*, 137-150.
