---
title: "Worked Example: Repeated Posttests"
description: >
  A published Solomon study with three posttest occasions (Jordaan, 2014):
  its analyses reproduced from the published group statistics, the same
  data analyzed with one model for all occasions, and what the published
  statistics can and cannot say about change in sensitization.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
has_mmrm <- requireNamespace("mmrm", quietly = TRUE)
```

When a Solomon study measures the outcome on several occasions, a new
question arises: does pretest sensitization persist, grow, or fade over
time? Jordaan (2014) evaluated a six-month life skills program for young
adult male offenders with a Solomon four-group design, and tested every
group three times: at the end of the program, and 3 and 6 months later.

The thesis reports the size, mean, and standard deviation of every group on
every occasion, and solomonR bundles them as `jordaan2014` (see
`?jordaan2014` for the design and the measures). This example:

- reproduces the published analyses from those statistics;
- fits one model for all occasions with `fit_solomon_mmrm()`;
- asks what the published statistics can say about change in
  sensitization, and what they cannot.

It reanalyzes published group statistics to teach the package. For the
program, the measures, and the author's conclusions, read the thesis
(Jordaan, 2014).

## The design and the data

- **Assignment.** 120 offenders aged 21 to 25 with long sentences were
  assigned at random to the program or to the control condition, and half
  of each condition was assigned at random to be pretested (Jordaan, 2014,
  pp. 86–87).
- **Occasions.** All four groups were tested after the program and again
  3 and 6 months later (p. 98).
- **Measures.** The three subscales of the Coping Strategy Indicator, each
  11 items scored from 1 to 3. Higher scores on problem solving and on
  seeking social support, and lower scores on avoidance, indicate better
  coping (p. 92).
- **Attrition.** Transfers removed 17 offenders from the program groups
  and 7 from the control groups (p. 109). The 96 who remained completed
  every posttest.

The problem-solving scores, as means (standard deviations):

```{r data}
ps <- subset(jordaan2014, subscale == "Problem solving")
levels(ps$occasion) <- c("Pretest", "Posttest", "3 months", "6 months")
ps$label <- factor(ps$group, labels = c("Program, pretested (n = 22)",
                                        "Program, not pretested (n = 21)",
                                        "Control, pretested (n = 22)",
                                        "Control, not pretested (n = 31)"))
tab <- with(ps, tapply(sprintf("%.2f (%.2f)", mean, sd), list(label, occasion), c))
tab[is.na(tab)] <- ""
knitr::kable(tab)
```

Randomization makes the groups comparable as assigned. More offenders were
lost from the program groups than from the control groups, so the 96 who
remained are comparable only if the transfers were unrelated to the
outcome. No analysis below can check or correct this: the offenders who
were transferred have no posttest at all.

## Reproduce the published analyses

The thesis analyzed each occasion separately by the decision sequence of
Walton Braver and Braver (1988). On each occasion it began with a 2 x 2
analysis of variance of the posttests, Pretest x Treatment.
`solomon_from_summary()` computes that analysis from group statistics. For
problem solving 6 months after the program:

```{r six-months}
six <- subset(ps, occasion == "6 months")
with(six, solomon_from_summary(n, mean, sd, treat = treat, pretested = pretested))
```

The interaction test, F(1, 92) = 5.54, p = .021, is the published result:
F = 5.556, p = .021 (Table 7.13, p. 121). The small difference comes from
the published means and standard deviations, which are rounded to two
decimals.

The thesis then compared the program with the control among pretested
and among unpretested offenders (Table 7.14, p. 122). Those are Tests B and
C above, but the thesis used two-group t tests, each with only its own two
groups' variance. Among unpretested offenders it found t(50) = −2.04,
p = .046; with the variance pooled over all four groups, Test C gives
t(92) = −1.79, p = .076.

The interaction tests of all three subscales on all three occasions,
against the published F values (Tables 7.4–7.19, pp. 113–127):

```{r all-occasions}
published <- data.frame(
  subscale = rep(levels(jordaan2014$subscale), each = 3),
  occasion = rep(c("Posttest", "Follow-up 1", "Follow-up 2"), 3),
  published_F = c(9.678, 0.266, 2.306, 0.819, 0.563, 5.556, 0.373, 0.688, 0.327)
)
interaction <- function(s, o) {
  x <- subset(jordaan2014, subscale == s & occasion == o)
  a <- with(x, solomon_from_summary(n, mean, sd, treat = treat, pretested = pretested))$anova
  a[a$source == "Treatment x Pretest", c("F", "p.value")]
}
nine <- cbind(published, do.call(rbind, Map(interaction, published$subscale, published$occasion)))
nine$p_holm <- p.adjust(nine$p.value, method = "holm")
rownames(nine) <- NULL
knitr::kable(nine, digits = 3)
```

Every published value is reproduced within rounding. In the thesis,
"Follow-up 1" and "Follow-up 2" are the tests 3 and 6 months after the
program. Two of the nine interactions are significant at .05: social
support after the program, and problem solving 6 months later. The
thesis reads both as signs of pretest sensitization (pp. 113, 122). It ran
the nine tests without adjustment. Adjusted together by Holm's (1979)
procedure (`p_holm`), only the social support interaction after the
program stays below .05.

## One model for all occasions

`fit_solomon_mmrm()` analyzes all occasions in one model: a mixed model for
repeated measures, specified as Mallinckrodt et al. (2008) recommend for
longitudinal trials. It needs each participant's scores, which the thesis
does not report. The function `rebuild()` below creates data whose group
means and standard deviations on every occasion are exactly the published
ones (`MASS::mvrnorm()` with `empirical = TRUE`). The correlations between
occasions are not reported either, so one must be assumed; here it is 0.5
for every pair of occasions.

```{r rebuild}
rebuild <- function(stats, r) {
  post <- subset(stats, occasion != "Pretest")
  occasions <- levels(droplevels(post$occasion))
  do.call(rbind, lapply(split(post, post$group), function(g) {
    g <- g[order(g$occasion), ]
    R <- matrix(r, nrow(g), nrow(g))
    diag(R) <- 1
    y <- MASS::mvrnorm(g$n[1], mu = g$mean, Sigma = diag(g$sd) %*% R %*% diag(g$sd),
                       empirical = TRUE)
    data.frame(id = paste(g$group[1], row(y), sep = "-"), treat = g$treat[1],
               pretested = g$pretested[1],
               occasion = factor(as.character(g$occasion)[col(y)], levels = occasions),
               y_post = as.vector(y))
  }))
}
set.seed(2014)
long <- rebuild(ps, r = 0.5)
```

The published analyses used the posttests alone, so the model is fitted
without the pretest:

```{r mmrm, eval = has_mmrm}
fit <- fit_solomon_mmrm(y_post, treat, pretested, id, occasion, data = long)
fit
```

The squared t statistics of the Pretest x Treatment rows are the F values of
the separate analyses:

```{r mmrm-check, eval = has_mmrm}
px <- fit$effects[fit$effects$contrast == "Pretest x Treatment", ]
data.frame(occasion = px$occasion, F = round(px$statistic^2, 3), df = round(px$df, 1),
           published_F = c(0.819, 0.563, 5.556))
```

They agree because no posttest is missing. The model's estimate of each
group's mean on an occasion is then the observed mean. Its estimate of the
variance on that occasion is the variance pooled within the four groups,
which is what the analysis of variance uses. So on each occasion the
assumed correlation has no effect: any value gives the same contrasts,
standard errors, and tests.

## Did sensitization change over time?

For problem solving, the interaction was 1.89 points after the program,
1.44 at 3 months, and 3.32 at 6 months, and only the last was significant.
That pattern does not show that sensitization grew. A significant result
on one occasion and a nonsignificant one on another is not itself evidence
of a difference between them (Gelman & Stern, 2006). The model tests the
change directly, in the row "Change in Pretest x Treatment": the
interaction at 6 months minus the interaction after the program.

The standard error of that change depends on how strongly each offender's
scores at the two occasions are correlated, which the thesis does not
report. So the analysis is repeated over a range of correlations:

```{r change, eval = has_mmrm}
change <- function(stats, r) {
  set.seed(2014)
  e <- fit_solomon_mmrm(y_post, treat, pretested, id, occasion, data = rebuild(stats, r))$effects
  e[e$contrast == "Change in Pretest x Treatment", c("estimate", "std.error", "p.value")]
}
ss <- subset(jordaan2014, subscale == "Social support")
r <- c(0.2, 0.5, 0.8)
knitr::kable(cbind(subscale = rep(c("Problem solving", "Social support"), each = 3),
                   correlation = rep(r, 2),
                   rbind(do.call(rbind, lapply(r, change, stats = ps)),
                         do.call(rbind, lapply(r, change, stats = ss)))),
             digits = 3, row.names = FALSE)
```

The estimate does not depend on the correlation, but its standard error
does.

- **Problem solving.** The interaction rose by 1.43 points. The rise is not
  significant whatever the correlation, so the data do not show that
  sensitization grew.
- **Social support.** The interaction fell by 3.26 points, from 5.79 after
  the program to 2.53 at 6 months. Whether that fall is significant
  depends on the correlation: p = .149 if it is 0.2, and p = .005 if it is
  0.8.

Whether sensitization to the social support items faded cannot be settled
from the published statistics.

## What the example shows

- **Separate occasions need only group statistics.** Every per-occasion
  Solomon analysis can be reproduced from sizes, means, and standard
  deviations. With no posttest missing, one model for all occasions gives
  the same per-occasion results.
- **Change over time needs the correlations.** Whether sensitization
  persists, grows, or fades depends on how each participant's scores on two
  occasions are related. When a study does not report the correlations, a
  reanalysis can only show the conclusion over a range of them, as here.
  When you report a Solomon study with several posttests, give the
  correlations between occasions, or the standard deviations of the change
  scores, so that others can reanalyze it.
- **The model's main advantage does not show here.** It stays valid when
  later posttests are missing at random (see "Missing and Repeated
  Posttests"), but nothing was missing among the 96 offenders analyzed. No
  model can recover the 24 offenders who were transferred before any
  posttest.
- **The joint model is a solomonR extension.** The literature search
  documented on issue
  [#57](https://github.com/JUhalt/solomonR/issues/57) found no published
  method for analyzing all the posttest occasions of a Solomon design in
  one model. Published studies, Jordaan (2014) among them, analyzed each
  occasion separately. `fit_solomon_mmrm()` applies the mixed model for
  repeated measures to the Solomon contrasts; its validation is in the
  article
  [Longitudinal Designs: Validating the Repeated-Measures Analysis](mmrm-validation.html).

## References

Gelman, A., & Stern, H. (2006). The difference between "significant" and
"not significant" is not itself statistically significant. *The American
Statistician, 60*(4), 328–331. https://doi.org/10.1198/000313006X152649

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics, 6*(2), 65–70.
https://www.jstor.org/stable/4615733

Jordaan, J. (2014). *The development and evaluation of a life skills
programme for young adult prisoners* [Doctoral thesis, University of the
Free State]. KovsieScholar. https://hdl.handle.net/11660/832

Mallinckrodt, C. H., Lane, P. W., Schnell, D., Peng, Y., & Mancuso, J. P.
(2008). Recommendations for the primary analysis of continuous endpoints in
longitudinal clinical trials. *Drug Information Journal, 42*(4), 303–319.
https://doi.org/10.1177/009286150804200402

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150
