---
title: "Reporting a Solomon Study"
description: >
  What to report from a Solomon four-group study, and how report_solomon()
  drafts the results, the design statement, and the references.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

A Solomon study report has to let readers judge two things: the treatment
effect and whether the pretest changed it. This article describes what to
report and shows how `report_solomon()` drafts it. The helper writes
sentences. The researcher decides what they mean.

## What to report

- **The design.**
  - How participants were assigned.
  - How many were assigned to and analyzed in each of the four groups. Any
    attrition, by group.
  - How the pretest and posttest were given in each group. MERIT
    Recommendation 11 is to use identical measurement protocols in all arms
    (French et al., 2021b, p. 34).
- **What was decided in advance.** Whether the sensitization analysis was
  pre-specified, and any smallest effect of interest for an equivalence
  test.
- **The estimands and analysis.**
  - Which contrasts were estimated: the average treatment effect, the
    Pretest x Treatment contrast, and the two simple treatment effects.
  - The model and how its uncertainty was computed.
- **The results.** Each estimate with its confidence interval. For binary
  or count outcomes, the scale of each contrast, because sensitization can
  differ between scales.
- **The references** for the methods used.

## Drafting it with `report_solomon()`

`report_solomon()` takes a fitted analysis and returns APA 7 results
sentences, a design statement, and the references for exactly the methods
and options that analysis used. Facts the data cannot show are supplied in
`design`. The helper never infers them. The example data have 30
participants per group; the numbers randomized below are hypothetical.

```{r report}
fit <- with(solomon_example, fit_solomon_glm(y_post, treat, pretested, y_pre))
report_solomon(fit, design = list(
  randomized = c(32, 31, 30, 31),
  prespecified = TRUE,
  measurement = "The same 20-item test, given by the same research assistant in all groups.",
  assignment = "random"
))
```

In the example:
- **Attrition.** The numbers randomized are compared with the numbers
  analyzed, so the statement reports attrition by group.
- **Pre-specification.** It is stated because it was supplied.
- **Other analyses.** The design statement is the same, but the results and
  references change. A clustered fit cites its cluster-robust standard
  errors. A historical analysis names the version of the test sequence. A
  nonrandomized design speaks of differences between groups rather than
  treatment effects.

For a manuscript, `format = "markdown"` italicizes statistical symbols as
APA style requires:

```{r markdown}
report_solomon(fit, format = "markdown")$results
```

## What the helper does not do

The report states results; it does not interpret them. Whether a
sensitization contrast is small enough to ignore is a substantive judgment.
`equivalence_solomon()` supports it when a smallest effect of interest was
fixed in advance. The choice of analysis, the reading of the results, and
the conclusions remain the researcher's.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

French, D. P., Miles, L. M., Elbourne, D., Farmer, A., Gulliford, M.,
Locock, L., Sutton, S., McCambridge, J., & MERIT Collaborative Group.
(2021b). Reducing bias in trials from reactions to measurement: The MERIT
study including developmental work and expert workshop. *Health Technology
Assessment, 25*(55), 1–72. https://doi.org/10.3310/hta25550
