---
title: "Structural Equation Models for Solomon Designs"
description: >
  Observed- and latent-variable SEM for the Solomon four-group design:
  the four-group mean structure, the pretested-groups ANCOVA, latent
  outcomes, and measurement invariance across the Solomon groups.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

When a Solomon study's outcome is measured by several items, a
structural equation model can compare latent means instead of scale
scores. This article shows the package's two SEM functions:

- **The observed-variable model** of the four groups, and its ANCOVA
  version for the pretested groups.
- **The latent-variable model**, and the measurement invariance its
  comparisons require.

The SEM functions use lavaan (Rosseel, 2012), and their tests and intervals
are lavaan's large-sample Wald results. They are a solomonR extension built
on established SEM methods, not a procedure proposed for the Solomon design
in the literature.

## Observed outcomes

`fit_solomon_sem()` estimates the posttest mean of each Solomon group in a
four-group mean-structure model and reports the four Solomon contrasts:

```{r observed}
sem <- with(solomon_example, fit_solomon_sem(y_post, treat, pretested))
sem
```

- **No global fit test.** The four-group mean-structure model is saturated,
  so its global fit indices are not diagnostic, and the function does not
  report them.
- **Unadjusted contrasts.** The contrasts are unadjusted posttest mean
  differences, which is the SEM counterpart of the two-by-two analysis.

With `ancova = TRUE`, the model is restricted to the pretested groups and
adjusts for the pretest. That is the analysis of covariance Campbell and
Stanley (1963/1966, p. 25) and Huck and Sandler (1973) describe for those
groups:

```{r observed-ancova}
with(solomon_example, fit_solomon_sem(y_post, treat, pretested, y_pre, ancova = TRUE))
```

The unpretested groups have no pretest by design, so the adjusted model
uses only the pretested groups. It does not treat the absent pretests as
missing data.

## Latent outcomes

`fit_solomon_sem_latent()` defines a latent posttest from several
indicators and compares its means across the four groups.

- **Identification.** The latent mean of the unpretested control group is
  fixed at 0; the contrasts are differences between latent means, so they
  do not depend on that choice. With the default `std_lv = TRUE`, the
  latent variance is fixed at 1 in the pretested treated group, so the
  contrasts are in standard deviations of the latent posttest there.
- **Optional ANCOVA.** With `ancova = TRUE` and pretest indicators, a latent
  ANCOVA in the pretested groups is added. With `std_lv = TRUE`, its
  adjusted effect is in residual standard deviations of the latent
  posttest given the latent pretest, a different unit from that of the
  four-group contrasts.

The example data below are simulated for this article. Four indicators
measure one construct in every group, and the treatment raises the latent
mean by 0.4 standard deviations:

```{r latent-data}
set.seed(2049)
n <- 80
g <- rep(1:4, each = n)
treat <- c(1, 0, 1, 0)[g]
pretested <- c(1, 1, 0, 0)[g]
f <- rnorm(4 * n, 0.4 * treat)
items <- data.frame(
  y1 = 0.8 * f + rnorm(4 * n, 0, 0.6),
  y2 = 0.9 * f + rnorm(4 * n, 0, 0.6),
  y3 = 0.7 * f + rnorm(4 * n, 0, 0.6),
  y4 = 0.8 * f + rnorm(4 * n, 0, 0.6)
)
latent <- fit_solomon_sem_latent(items, c("y1", "y2", "y3", "y4"), treat, pretested)
latent
```

## Measurement invariance

Latent means can be compared only if the indicators measure the construct
in the same way in every group. The loadings and the intercepts must be
equal across groups, which is scalar invariance (Meredith, 1993).

- **Why this matters for Solomon designs.** Taking the pretest may change
  how participants answer some items at the posttest, for example by making
  them familiar with the items.
- **Where the damage falls.** A shift shared by both pretested groups is
  absorbed into their latent means and looks like an effect of pretesting.
  It cancels from the sensitization contrast, which compares treatment
  effects within each pretest condition. A shift in one group only, such
  as the treated pretested participants reading an item differently, biases
  the sensitization contrast itself. The package's simulation study
  confirmed both patterns, with the one-group shift placed in the
  unpretested control group (see "Latent Contrasts: Validating the
  Invariance Check").

`invariance_solomon()` fits the configural, metric, and scalar models
across the four groups, in the sequence Vandenberg and Lance (2000,
pp. 56–57) recommend:

```{r invariance}
inv <- invariance_solomon(items, c("y1", "y2", "y3", "y4"), treat, pretested)
inv
```

Each step reports two published criteria:

- **The chi-square difference test.** Scaled for the robust MLR estimator
  (Satorra & Bentler, 2001). Vandenberg and Lance (2000, p. 46) recommend
  it as the primary criterion.
- **The change in fit.** The drop in CFI and the rises in RMSEA and SRMR,
  against Chen's (2007, pp. 501–502) cutoffs. Cheung and Rensvold (2002,
  pp. 234–235) favor such changes over the chi-square difference because
  the latter depends on sample size.

Here both criteria support scalar invariance, so the latent contrasts above
can be interpreted.

### When an intercept is not invariant

Suppose the pretest makes item `y4` read higher in the two pretested groups
at the same latent level:

```{r noninvariant}
shifted <- items
shifted$y4[pretested == 1] <- shifted$y4[pretested == 1] + 0.6
invariance_solomon(shifted, c("y1", "y2", "y3", "y4"), treat, pretested)
```

The scalar step fails. If theory identifies the noninvariant parameter,
freeing it gives a partial-invariance model, in which the latent means are
compared through the invariant indicators (Byrne et al., 1989). Byrne et
al. caution against relaxing constraints that do not make sense
substantively (p. 465). Vandenberg and Lance (2000, p. 38) set three
conditions:

- the freed parameters involve only a minority of the indicators;
- the choice rests on as strong a theoretical basis as possible;
- cross-validation evidence supports it.

The package enforces the first condition and never chooses the parameters
itself:

```{r partial}
fit_partial <- fit_solomon_sem_latent(shifted, c("y1", "y2", "y3", "y4"), treat, pretested,
                                      partial_post = "y4 ~ 1")
fit_partial$effects_post[, c("contrast", "estimate", "conf.low", "conf.high")]
```

## What the package does when invariance fails

`fit_solomon_sem_latent()` runs `invariance_solomon()` on the POST
indicators before it fits the latent means. When a criterion does not
support scalar invariance (or partial scalar invariance, with
`partial_post`), it warns. It does not refuse the contrasts:

```{r check-warning}
fit_shifted <- withCallingHandlers(
  fit_solomon_sem_latent(shifted, c("y1", "y2", "y3", "y4"), treat, pretested),
  solomonR_invariance_warning = function(w) {
    message("Warning: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
fit_shifted$invariance_status
fit_shifted$effects_post[, c("contrast", "estimate", "conf.low", "conf.high")]
```

The package's simulation study settled that choice. Its decision rules
were posted on issue #55 before any run (see "Latent Contrasts:
Validating the Invariance Check"). A criterion could govern an automatic
refusal only if it falsely rejected invariance at a rate of .060 or less
in Solomon-sized groups, and none did. The warning is therefore a prompt
to examine the invariance results, not a verdict.

The shift in this example is common to both pretested groups, so the
sensitization contrast is close to the one from the partial model above;
the shift is absorbed into the effect of pretesting. A shift in one group
only would bias the sensitization contrast, and there the partial model
matters. `check_invariance = FALSE` skips the check when invariance was
established elsewhere.

## Reporting latent models

`report_solomon()` drafts the methods and results of a latent fit. It gives
what the reporting standards for structural equation models ask for
(Appelbaum et al., 2018):

- **The estimation.** The software and its version, the estimator, and how
  missing values were handled.
- **The model.** The constraints across groups, any freed parameters, and
  how the latent scale was identified.
- **The fit.** The chi-square test, CFI, RMSEA, and SRMR.

Its statement about invariance follows the fit's own check, so for the
partial model above it names the freed intercept:

```{r report-latent}
report_solomon(fit_partial)$method
```

For `invariance_solomon()` results, it reports the fit of each model, the
comparisons between them, and the decision under each criterion, the
minimal information Putnick and Bornstein (2016) propose for invariance
tests:

```{r report-invariance}
report_solomon(inv)$results
```

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Appelbaum, M., Cooper, H., Kline, R. B., Mayo-Wilson, E., Nezu, A. M., &
Rao, S. M. (2018). Journal article reporting standards for quantitative
research in psychology: The APA Publications and Communications Board task
force report. *American Psychologist, 73*(1), 3–25.
https://doi.org/10.1037/amp0000191

Byrne, B. M., Shavelson, R. J., & Muthén, B. (1989). Testing for the
equivalence of factor covariance and mean structures: The issue of partial
measurement invariance. *Psychological Bulletin, 105*(3), 456–466.
https://doi.org/10.1037/0033-2909.105.3.456

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of
measurement invariance. *Structural Equation Modeling: A Multidisciplinary
Journal, 14*(3), 464–504. https://doi.org/10.1080/10705510701301834

Cheung, G. W., & Rensvold, R. B. (2002). Evaluating goodness-of-fit indexes
for testing measurement invariance. *Structural Equation Modeling: A
Multidisciplinary Journal, 9*(2), 233–255.
https://doi.org/10.1207/S15328007SEM0902_5

Huck, S. W., & Sandler, H. M. (1973). A note on the Solomon 4-group design:
Appropriate statistical analyses. *The Journal of Experimental Education,
42*(2), 54–55. https://doi.org/10.1080/00220973.1973.11011460

Meredith, W. (1993). Measurement invariance, factor analysis and factorial
invariance. *Psychometrika, 58*(4), 525–543.
https://doi.org/10.1007/BF02294825

Putnick, D. L., & Bornstein, M. H. (2016). Measurement invariance
conventions and reporting: The state of the art and future directions for
psychological research. *Developmental Review, 41*, 71–90.
https://doi.org/10.1016/j.dr.2016.06.004

Rosseel, Y. (2012). lavaan: An R package for structural equation modeling.
*Journal of Statistical Software, 48*(2), 1–36.
https://doi.org/10.18637/jss.v048.i02

Satorra, A., & Bentler, P. M. (2001). A scaled difference chi-square test
statistic for moment structure analysis. *Psychometrika, 66*(4), 507–514.
https://doi.org/10.1007/BF02296192

Vandenberg, R. J., & Lance, C. E. (2000). A review and synthesis of the
measurement invariance literature: Suggestions, practices, and
recommendations for organizational research. *Organizational Research
Methods, 3*(1), 4–70. https://doi.org/10.1177/109442810031002
