---
title: "Structural Equation Models for Solomon Designs"
description: >
  Observed- and latent-variable SEM for the Solomon four-group design:
  the four-group mean structure, the pretested-groups ANCOVA, latent
  outcomes, and measurement invariance across the Solomon groups.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

When a Solomon study's outcome is measured by several items, a
structural equation model can compare latent means instead of scale
scores. This article shows the package's two SEM functions:

- **The observed-variable model** of the four groups, and its ANCOVA
  version for the pretested groups.
- **The latent-variable model**, and the measurement invariance its
  comparisons require.

The SEM functions use lavaan (Rosseel, 2012), and their tests and intervals
are lavaan's large-sample Wald results. Latent-variable analysis of a
Solomon design has a published precedent, Dukes et al. (1995), which the
last section compares with the package's approach. The four-group models
of the SEM functions, and the observed-variable model, are solomonR
extensions built on established SEM methods.

## Observed outcomes

`fit_solomon_sem()` estimates the posttest mean of each Solomon group in a
four-group mean-structure model and reports the four Solomon contrasts:

```{r observed}
sem <- with(solomon_example, fit_solomon_sem(y_post, treat, pretested))
sem
```

- **No global fit test.** The four-group mean-structure model is saturated,
  so its global fit indices are not diagnostic, and the function does not
  report them.
- **Unadjusted contrasts.** The contrasts are unadjusted posttest mean
  differences, which is the SEM counterpart of the two-by-two analysis.

With `ancova = TRUE`, the model is restricted to the pretested groups and
adjusts for the pretest. That is the analysis of covariance Campbell and
Stanley (1963/1966, p. 25) and Huck and Sandler (1973) describe for those
groups:

```{r observed-ancova}
with(solomon_example, fit_solomon_sem(y_post, treat, pretested, y_pre, ancova = TRUE))
```

The unpretested groups have no pretest by design, so the adjusted model
uses only the pretested groups. It does not treat the absent pretests as
missing data.

## Latent outcomes

`fit_solomon_sem_latent()` defines a latent posttest from several
indicators and compares its means across the four groups.

- **Identification.** The latent mean of the unpretested control group is
  fixed at 0; the contrasts are differences between latent means, so they
  do not depend on that choice.
- **Optional ANCOVA.** With `ancova = TRUE` and pretest indicators, a latent
  ANCOVA in the pretested groups is added.

The example data below are simulated for this article. Four indicators
measure one construct in every group, and the treatment raises the latent
mean by 0.4 standard deviations:

```{r latent-data}
set.seed(2049)
n <- 80
g <- rep(1:4, each = n)
treat <- c(1, 0, 1, 0)[g]
pretested <- c(1, 1, 0, 0)[g]
f <- rnorm(4 * n, 0.4 * treat)
items <- data.frame(
  y1 = 0.8 * f + rnorm(4 * n, 0, 0.6),
  y2 = 0.9 * f + rnorm(4 * n, 0, 0.6),
  y3 = 0.7 * f + rnorm(4 * n, 0, 0.6),
  y4 = 0.8 * f + rnorm(4 * n, 0, 0.6)
)
latent <- fit_solomon_sem_latent(items, c("y1", "y2", "y3", "y4"), treat, pretested)
latent
```

## Measurement invariance

Latent means can be compared only if the indicators measure the construct
in the same way in every group. The loadings and the intercepts must be
equal across groups, which is scalar invariance (Meredith, 1993).

- **Why this matters for Solomon designs.** Taking the pretest may change
  how participants answer some items at the posttest, for example by making
  them familiar with the items.
- **Where the damage falls.** A shift shared by both pretested groups is
  absorbed into their latent means and looks like an effect of pretesting.
  It cancels from the sensitization contrast, which compares treatment
  effects within each pretest condition. A shift in one group only, such
  as the treated pretested participants reading an item differently, biases
  the sensitization contrast itself. The package's simulation study
  confirmed both patterns, with the one-group shift placed in the
  unpretested control group (see "Latent Contrasts: Validating the
  Invariance Check").

`invariance_solomon()` fits the configural, metric, and scalar models
across the four groups, in the sequence Vandenberg and Lance (2000,
pp. 56–57) recommend:

```{r invariance}
inv <- invariance_solomon(items, c("y1", "y2", "y3", "y4"), treat, pretested)
inv
```

Each step reports two published criteria:

- **The chi-square difference test.** Scaled for the robust MLR estimator
  (Satorra & Bentler, 2001). Vandenberg and Lance (2000, p. 46) recommend
  it as the primary criterion.
- **The change in fit.** The drop in CFI and the rises in RMSEA and SRMR,
  against Chen's (2007, pp. 501–502) cutoffs. Cheung and Rensvold (2002,
  pp. 234–235) favor such changes over the chi-square difference because
  the latter depends on sample size.

Here both criteria support scalar invariance, so the latent contrasts above
can be interpreted.

### When an intercept is not invariant

Suppose the pretest makes item `y4` read higher in the two pretested groups
at the same latent level:

```{r noninvariant}
shifted <- items
shifted$y4[pretested == 1] <- shifted$y4[pretested == 1] + 0.6
invariance_solomon(shifted, c("y1", "y2", "y3", "y4"), treat, pretested)
```

The scalar step fails. If theory identifies the noninvariant parameter,
freeing it gives a partial-invariance model, in which the latent means are
compared through the invariant indicators (Byrne et al., 1989). Byrne et
al. caution against relaxing constraints that do not make sense
substantively (p. 465). Vandenberg and Lance (2000, p. 38) set three
conditions:

- the freed parameters involve only a minority of the indicators;
- the choice rests on as strong a theoretical basis as possible;
- cross-validation evidence supports it.

The package enforces the first condition and never chooses the parameters
itself:

```{r partial}
fit_partial <- fit_solomon_sem_latent(shifted, c("y1", "y2", "y3", "y4"), treat, pretested,
                                      partial_post = "y4 ~ 1")
fit_partial$effects_post[, c("contrast", "estimate", "conf.low", "conf.high")]
```

## What the package does when invariance fails

`fit_solomon_sem_latent()` runs `invariance_solomon()` on the POST
indicators before it fits the latent means. When a criterion does not
support scalar invariance (or partial scalar invariance, with
`partial_post`), it warns. It does not refuse the contrasts:

```{r check-warning}
fit_shifted <- withCallingHandlers(
  fit_solomon_sem_latent(shifted, c("y1", "y2", "y3", "y4"), treat, pretested),
  solomonR_invariance_warning = function(w) {
    message("Warning: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
fit_shifted$invariance_status
fit_shifted$effects_post[, c("contrast", "estimate", "conf.low", "conf.high")]
```

The package's simulation study settled that choice. Its decision rules
were posted on issue #55 before any run (see "Latent Contrasts:
Validating the Invariance Check"). A criterion could govern an automatic
refusal only if it falsely rejected invariance at a rate of .060 or less
in Solomon-sized groups, and none did. The warning is therefore a prompt
to examine the invariance results, not a verdict.

The shift in this example is common to both pretested groups, so the
sensitization contrast is close to the one from the partial model above;
the shift is absorbed into the effect of pretesting. A shift in one group
only would bias the sensitization contrast, and there the partial model
matters. `check_invariance = FALSE` skips the check when invariance was
established elsewhere.

## Dukes et al. (1995) and solomonR

Dukes et al. (1995) evaluated a school drug-prevention program with a
Solomon design and four latent outcomes measured by self-report items
(pp. 412, 414). They compared latent means in multisample structural
equation models.

- **Three two-group models.** They fitted three separate two-group models
  rather than one model of the four groups (p. 420): the pretested
  control's pretest against the unpretested control's posttest, for
  maturation, and the program against the control within the pretested and
  within the unpretested groups.
- **Loadings tested, intercepts imposed.** In each comparison, equal factor
  loadings were tested by chi-square difference tests. The latent means
  were then compared with the indicators' means held equal, a constraint
  that was imposed, not tested (pp. 420–426).
- **No test of the interaction.** Pretest effects were judged by comparing
  which program effects were significant in the two program comparisons
  (p. 426), the comparison Gelman and Stern (2006) caution against. The
  abstract calls the pretest reactive for one outcome, resistance to peer
  pressure (p. 409). The results are more consistent with a pretest main
  effect, seen in one of that outcome's two indicators, than with
  sensitization. The paper reports no test of the interaction, and
  solomonR's approximate re-analyses find none, either in the difference
  between the two latent program effects (Table 4, p. 425) or in two-by-two
  analyses of the two indicators' classroom posttest means (Table 2,
  pp. 415–416).
- **Classroom means.** They analyzed 440 classroom means (pp. 413, 417).
  Such means are more stable and precise (p. 412), and the program was
  delivered to intact classes, although the school, the unit assigned to
  the program, was the appropriate unit (p. 432, note 3).

solomonR fits the four groups in one model. Its latent mean structure is
identified by fixing the unpretested control's mean at 0
([#16](https://github.com/JUhalt/solomonR/issues/16)), so the
sensitization contrast, the average treatment effect, and both simple
effects have their own Wald tests on one scale. `invariance_solomon()`
tests configural, metric, and scalar invariance across all four groups
([#55](https://github.com/JUhalt/solomonR/issues/55)), including the
posttests of pretested against unpretested groups, which Dukes et al. did
not compare. The SEM functions make no adjustment for clusters. When
clusters such as schools were assigned to conditions, analyze at the level
of assignment, or analyze observed scores with
`fit_solomon_glm(robust = "CR2", cluster = )`.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Byrne, B. M., Shavelson, R. J., & Muthén, B. (1989). Testing for the
equivalence of factor covariance and mean structures: The issue of partial
measurement invariance. *Psychological Bulletin, 105*(3), 456–466.
https://doi.org/10.1037/0033-2909.105.3.456

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of
measurement invariance. *Structural Equation Modeling: A Multidisciplinary
Journal, 14*(3), 464–504. https://doi.org/10.1080/10705510701301834

Cheung, G. W., & Rensvold, R. B. (2002). Evaluating goodness-of-fit indexes
for testing measurement invariance. *Structural Equation Modeling: A
Multidisciplinary Journal, 9*(2), 233–255.
https://doi.org/10.1207/S15328007SEM0902_5

Dukes, R. L., Ullman, J. B., & Stein, J. A. (1995). An evaluation of D.A.R.E.
(Drug Abuse Resistance Education), using a Solomon four-group design with
latent variables. *Evaluation Review, 19*(4), 409–435.
https://doi.org/10.1177/0193841X9501900404

Gelman, A., & Stern, H. (2006). The difference between "significant" and
"not significant" is not itself statistically significant. *The American
Statistician, 60*(4), 328–331. https://doi.org/10.1198/000313006X152649

Huck, S. W., & Sandler, H. M. (1973). A note on the Solomon 4-group design:
Appropriate statistical analyses. *The Journal of Experimental Education,
42*(2), 54–55. https://doi.org/10.1080/00220973.1973.11011460

Meredith, W. (1993). Measurement invariance, factor analysis and factorial
invariance. *Psychometrika, 58*(4), 525–543.
https://doi.org/10.1007/BF02294825

Rosseel, Y. (2012). lavaan: An R package for structural equation modeling.
*Journal of Statistical Software, 48*(2), 1–36.
https://doi.org/10.18637/jss.v048.i02

Satorra, A., & Bentler, P. M. (2001). A scaled difference chi-square test
statistic for moment structure analysis. *Psychometrika, 66*(4), 507–514.
https://doi.org/10.1007/BF02296192

Vandenberg, R. J., & Lance, C. E. (2000). A review and synthesis of the
measurement invariance literature: Suggestions, practices, and
recommendations for organizational research. *Organizational Research
Methods, 3*(1), 4–70. https://doi.org/10.1177/109442810031002
