---
title: "Designs With Several Treatments"
description: >
  Solomon six- and eight-group designs: one model for all the groups,
  comparisons of each treatment with the control, planned comparisons for
  factorial designs, and Steyn's (2009) sequence of tests.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

A Solomon design can compare more than one treatment. With k treatments
and one control, each with and without a pretest, it has 2(k + 1) groups:
six for two treatments and eight for three (Steyn, 2009). Edmonds and
Kennedy (2017, pp. 93–101) call these Solomon six-group and eight-group
designs. Two control groups serve every treatment, so the design needs
fewer participants than a separate four-group study of each treatment
(Steyn, 2009).

This article shows how to analyze such a design with solomonR:

- one model for all the groups, in place of a separate analysis for each
  pair of conditions;
- comparisons of each treatment with the control, with p-values adjusted
  for their number;
- planned comparisons, such as the main effects of two treatments crossed
  factorially;
- the sequence of tests that Steyn (2009) proposed.

## The design

```{r design, fig.width = 7, fig.height = 4.2, fig.alt = "Schematic of a Solomon six-group design: two treatments and a control, each with and without a pretest."}
plot_solomon_design(treatments = 2)
```

The example is the study of Mai et al. (2020), bundled as `mai2020`.
Participants were randomly assigned to relapse prevention (`RP`), goal
setting (`GS`), or a control condition, and to be pretested or not.

```{r validate}
validate_solomon(post_behavior, condition, pretested, pre_behavior,
                 control = "Control", data = mai2020)
```

`control` names the control condition. solomonR never guesses it from the
order of the conditions.

## Why one model

Published studies with several treatments analyzed them as overlapping
four-group designs. Mai et al. (2020) ran three: relapse prevention against
control, goal setting against control, and relapse prevention against goal
setting. McCarthy and Tucker (2002) ran one for each of three treatments
against the control. Each analysis tests the pretest-by-treatment
interaction:

```{r overlapping}
pairs <- list(c("RP", "Control"), c("GS", "Control"), c("RP", "GS"))
overlapping <- do.call(rbind, lapply(pairs, function(p) {
  d <- subset(mai2020, condition %in% p & !is.na(post_behavior))
  d$treat <- as.integer(d$condition == p[1])
  a <- anova(lm(post_behavior ~ treat * pretested, data = d))
  data.frame(comparison = paste(p, collapse = " vs "),
             F = round(a["treat:pretested", "F value"], 3),
             p = round(a["treat:pretested", "Pr(>F)"], 3))
}))
overlapping
```

These are the F values of the authors' Table 4. Two things make them hard
to interpret together:

- **The tests are dependent.** Every group appears in two of the three
  analyses.
- **Their number is not allowed for.** Three tests at the .05 level reject
  at least one true null hypothesis more often than 5% of the time.

One model for all six groups asks the question once. Does pretesting change
the effect of any treatment?

```{r joint-conventional}
joint <- fit_solomon_glm(post_behavior, condition, pretested,
                         control = "Control", robust = "none", data = mai2020)
joint$omnibus
```

The Pretest x Condition test has two numerator degrees of freedom, one for
each treatment. This fit uses the posttests alone and conventional standard
errors, to match the analysis of variance in the article.

## The joint model

By default the model also adjusts for the pretest in the pretested groups
and uses HC3 standard errors, as for the four-group design:

```{r joint}
fit <- fit_solomon_glm(post_behavior, condition, pretested, pre_behavior,
                       control = "Control", data = mai2020)
fit
```

The output has three parts:

- **Omnibus tests.** Each asks whether the conditions differ on one
  contrast.
- **Contrasts.** The four Solomon contrasts of the four-group design, for
  each treatment against the control, and then the pretest effect in each
  condition (pretested minus unpretested participants, at the pretested
  participants' mean pretest) and averaged over the conditions.
- **Adjusted p-values.** Each contrast's p-values are adjusted across the
  comparisons by Holm's (1979) procedure, which controls the chance of at
  least one false rejection in that family; the treatments' pretest effects
  are adjusted across the treatments. The confidence intervals are not
  adjusted.

`contrasts = "pairwise"` adds the comparisons between treatments:

```{r pairwise}
pairwise <- fit_solomon_glm(post_behavior, condition, pretested, pre_behavior,
                            control = "Control", contrasts = "pairwise",
                            data = mai2020)
subset(pairwise$effects, contrast == "Pretest x Treatment",
       c(comparison, estimate, std.error, p.value, p.adjusted))
```

Only 133 of the 211 participants gave a posttest, and the loss differed by
group:

```{r missing}
check_solomon_missing(post_behavior, condition, pretested, pre_behavior,
                      control = "Control", data = mai2020)
```

## Plots

`plot_sensitization()` shows the model's posttest means by pretest
condition, one line per condition. Lines that are not parallel to the
control's indicate sensitization.

```{r plot-sens, fig.width = 7, fig.height = 4.5, fig.alt = "Model-based posttest means with and without a pretest for relapse prevention, goal setting, and control."}
plot_sensitization(fit)
```

`plot_solomon_effects()` shows the four contrasts of every comparison with
their confidence intervals:

```{r plot-effects, fig.width = 7, fig.height = 5, fig.alt = "Estimates and confidence intervals of the four Solomon contrasts for each treatment against the control."}
plot_solomon_effects(fit)
```

## Planned comparisons: a factorial design

McCarthy and Tucker's (2002) eight groups crossed two treatments, a lecture
and a service-learning project, with pretesting: students received the
lecture, the project, both, or neither. In such a design the questions of
interest are the main effect of each treatment and their interaction, not
only each condition against the control. Give them as weights over the
conditions. The data here are simulated with known effects: the lecture
adds 2 points, the project 4, and together they add 9.

```{r factorial}
d <- simulate_solomon(n = 40, delta = c(Lecture = 2, Service = 4, Both = 9),
                      rho = 0.5, sigma = 10, mean = 50, seed = 2002)
factorial <- fit_solomon_glm(
  y_post, treat, pretested, y_pre, control = "Control", data = d,
  contrasts = list(
    "Lecture (main effect)" = c(Lecture = 0.5, Both = 0.5, Service = -0.5, Control = -0.5),
    "Service (main effect)" = c(Service = 0.5, Both = 0.5, Lecture = -0.5, Control = -0.5),
    "Lecture x Service" = c(Both = 1, Lecture = -1, Service = -1, Control = 1)
  )
)
shown <- subset(factorial$effects,
                contrast %in% c("ATE (avg over pretest)", "Pretest x Treatment"),
                c(comparison, contrast, estimate, std.error, p.adjusted))
shown$p.adjusted <- round(shown$p.adjusted, 3)
print(shown, digits = 3, row.names = FALSE)
```

The true main effects are 3.5 points for the lecture and 5.5 for the
project, and the true interaction is 3. With 40 participants per group the
estimates carry sampling error: in this data set the interaction is
estimated at more than one standard error above its true value. Each
comparison also has a Pretest x Treatment row: the difference in that
comparison between pretested and unpretested participants. Its true value
is zero here, because the data were simulated without sensitization; the
estimates in those rows differ from zero only by sampling error. The
weights of a comparison must sum to zero.

## A published eight-group study

Steyn (2005) ran the eight-group study that Steyn (2009) describes: three
treatments and a control, each with and without a pretest, with 1,723
police trainees. Each treatment gave participants more information about
their own ability, and the outcome was a questionnaire of self-efficacy
perceptions. The thesis reports each group's size, mean, and standard
deviation (pp. 151–152), bundled as `steyn2005`. `solomon_from_summary()`
analyzes a design from those statistics:

```{r steyn2005}
steyn2005[, c("group", "condition", "pretested", "n", "mean", "sd")]
s2005 <- with(steyn2005, solomon_from_summary(n, mean, sd, treat = condition,
                                              pretested = pretested, control = "Control"))
s2005
```

The thesis analyzed the study as overlapping four-group designs and then
compared the eight posttest means. The joint model gives one test of
sensitization, the Pretest x Condition row, and the comparisons of each
treatment with the control. All three treatments lowered the scores.

Two cautions apply to this reanalysis, as to the original analysis.
Existing classes, not participants, were allocated to the eight groups, and
not at random (Steyn, 2005, pp. 105–106). The analysis treats participants
as the units, so it does not allow for the classes.

`report_solomon()` drafts the method and results of the reanalysis. Because
the classes were not allocated at random, `assignment = "nonrandom"` makes
it describe differences between groups rather than treatment effects. The
omnibus tests and the first comparison:

```{r steyn2005-report}
report_solomon(s2005, design = list(assignment = "nonrandom"))$results[1:2]
```

## Steyn's (2009) sequence of tests

Steyn (2009) proposed a sequence of tests for these designs, in two parts:

- **Checks of internal validity.** Whether the pretested groups were
  equivalent before treatment; history and maturation; the effect of
  testing; the pretest-intervention interaction; reliability and
  instrumentation; regression to the mean; and attrition.
- **The effects of the treatments.** A one-way analysis of variance of all
  the posttests, then of the treatment groups, with post hoc tests.

`fit_solomon_steyn()` carries it out and labels each step. Its post hoc
tests are Scheffé's (1953) tests, which Steyn (2005, p. 153) used:

```{r steyn}
steyn <- fit_solomon_steyn(post_behavior, condition, pretested, pre_behavior,
                           control = "Control", data = mai2020)
steyn
```

The sequence is included because it is the published proposal for these
designs. It is not the package's recommended analysis, for three reasons:

- **It runs many tests.** Steyn (2009) notes that several analyses of
  variance capitalize on chance.
- **It tests sensitization one treatment at a time.** Step 4 tests the
  Pretest x Intervention interaction for each treatment against the
  control, in overlapping four-group analyses, each at the nominal level.
  Its later comparison of each treatment's pretested and unpretested groups
  (E4) mixes the effect of the pretest itself with sensitization. The joint
  model gives one Pretest x Condition test for all the treatments, and a
  Pretest x Treatment contrast for each with adjusted p-values.
- **Some of its tests treat paired scores as independent.** The help page
  lists these under "Cautions".

R. Steyn (personal communication, September 30, 2026) notes that the model
his article proposes is intended for situations with a large amount of data
and ample time. solomonR follows a pre-publication draft of Steyn's
article, which the author provided (R. Steyn, personal communication,
September 30, 2026). The published article was not available for
comparison. The help page lists every choice solomonR made where the
description leaves room.

## Reporting

`report_solomon()` writes the method and results with their references:

```{r report}
report_solomon(fit)
```

## How well the joint model performs

The joint model was checked in a simulation study with two or three
treatments and 10 to 50 participants per group, under a protocol posted
before the study was run. The article
[Designs With Several Treatments: Validating the Joint Model](ngroup-validation.html)
reports it. In summary:

- **Contrasts.** They were unbiased, and coverage of their 95% intervals
  was 0.939 to 0.967.
- **Adjusted comparisons.** The chance of at least one false rejection in a
  family of Holm-adjusted comparisons was at most 0.059.
- **Pretest x Condition test.** It rejected a true null hypothesis in 0.036
  to 0.055 of replications.
- **Two omnibus tests at the smallest size.** With three treatments and 10
  participants per group, the omnibus tests of Condition | pretested and
  Condition | unpretested rejected in 0.055 to 0.069 of replications at the
  .05 level. The study's rule for error control was therefore not met, and
  the analysis of designs with several treatments is labeled experimental.
  With groups that small, judge those two questions by the adjusted
  comparisons.
- **Overlapping four-group analyses.** When no treatment was sensitized,
  at least one of the separate interaction tests was significant in 8% to
  22% of replications.

## What needs a four-group design

The joint model, the design checks, the plots above, `baseline_solomon()`,
`equivalence_solomon()`, `solomon_from_summary()`, and `simulate_solomon()`
accept designs with several treatments. The other analyses take one
treatment and a control, and say so when given more: full-information
maximum likelihood, the historical test sequence, randomization tests,
marginal contrasts, structural equation models, multiple imputation, and
repeated posttests. To use one of them, analyze one treatment and the
control at a time:

```{r subset}
rp <- subset(mai2020, condition %in% c("RP", "Control") & !is.na(post_behavior))
rp$treat <- as.integer(rp$condition == "RP")
fit_solomon_ml(post_behavior, treat, pretested, pre_behavior, data = rp)$effects[, 1:5]
```

## References

Edmonds, W. A., & Kennedy, T. D. (2017). *An applied guide to research
designs: Quantitative, qualitative, and mixed methods* (2nd ed.). SAGE
Publications. https://doi.org/10.4135/9781071802779

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics, 6*(2), 65–70.
https://www.jstor.org/stable/4615733

Mai, N. N., Takahashi, Y., & Oo, M. M. (2020). Testing the effectiveness of
transfer interventions using Solomon four-group designs. *Education
Sciences, 10*(4), Article 92. https://doi.org/10.3390/educsci10040092

McCarthy, A. M., & Tucker, M. L. (2002). Encouraging community service
through service learning. *Journal of Management Education, 26*(6), 629–647.
https://doi.org/10.1177/1052562902238322

Scheffé, H. (1953). A method for judging all contrasts in the analysis of
variance. *Biometrika, 40*(1–2), 87–104.
https://doi.org/10.1093/biomet/40.1-2.87

Steyn, R. (2005). *Self-evaluasie en die vorming van
selfdoeltreffendheidspersepsies* [Self-evaluation and the forming of
self-efficacy perceptions] [Doctoral thesis, University of South Africa].
Unisa Institutional Repository. https://hdl.handle.net/10500/1745

Steyn, R. (2009). Re-designing the Solomon four-group: Can we improve on
this exemplary model? *Design Principles and Practices: An International
Journal—Annual Review, 3*(1), 383–394. https://doi.org/10.18848/1833-1874/CGP/v03i01/37588
