---
title: "Should I Use a Solomon Design?"
description: >
  When pretest sensitization is a real threat, what the Solomon four-group
  design costs, and what the alternatives are.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(solomonR)
```

A Solomon four-group design answers one question that a pretest-posttest
trial cannot: does taking the pretest change the effect of the treatment?
This article helps you decide whether that question is worth the design's
cost. Each recommendation is traced to a source. Where a point is solomonR's
own synthesis, it says so.

## What the design protects against

Solomon (1949) proposed the extra groups because a pretest can interact with
the treatment. It can change how participants attend to the treatment or
how they receive it (p. 141). In his own demonstration, the pretest reduced
the effect of a spelling lesson, so an ordinary two-group design would have
underrated the lesson (p. 145).

- **External validity.** Campbell and Stanley (1963/1966) treated this
  interaction of testing and treatment as a threat to external validity,
  first presented by Solomon (p. 18).
- **When it matters.** How much it matters depends on whether pretests are
  part of the setting the results should generalize to. They expected
  little harm in classrooms where testing is routine and more in attitude
  research, where pretests introduce unusual content (p. 18).
- **What the design determines.** The Solomon design determines both the
  main effect of testing and its interaction with the treatment (p. 25).

The design earns its cost only if sensitization is plausible. Sawilowsky
and Markman (1990b) laid out four possible outcomes: a treatment effect with
or without pretest sensitization, and no treatment effect with or without
it. When no sensitization is expected, they argued, the Solomon design has
no advantage over a two-group design (p. 424).

## How often sensitization occurs

The evidence suggests that sensitization is uncommon but not negligible.

- **Willson and Putnam (1982).** Their meta-analysis of 32 studies found
  that pretesting raised posttest scores by about a fifth of a standard
  deviation on average. The evidence for a pretest-by-treatment interaction
  was weak.
- **McCambridge et al. (2011).** Their review of Solomon studies of behavior
  change found sparse and inconsistent evidence of pretest effects, from too
  few studies to settle the question.

Sensitization can still change a study's conclusions:

- **Kvalem et al. (1996).** In a sex-education trial, condom use differed
  between intervention and control students who had been pretested (70%
  against 51%). It did not differ among students who had not been pretested
  (43% against 52%; p. 42; see `kvalem1996`).
- **El Karkri et al. (2025a).** Their classroom study reported a
  significant Pretest x Treatment interaction (pp. 11–12; see
  `elkarkri2025a`).

## When to worry

The MERIT study (French et al., 2021a, 2021b) developed recommendations for
bias from reactions to measurement in trials. Its "red flag" features mark
trials at higher risk (French et al., 2021b, Table 4, p. 24). Among them:

- **Participants.** Participants selected on their motivation to take part,
  restrictive eligibility criteria, or more highly educated participants.
- **Outcomes.** Participants aware of the outcome of interest; outcomes
  focused on behavior or anxiety; outcomes with strong social norms.
- **Measurement.** Measurement that is:
  - repeated, or followed up soon after the pretest;
  - self-reported, or collected specifically for the study;
  - burdensome or complex;
  - framed around goals;
  - different across trial arms;
  - disclosed to participants;
  - closely similar to the intervention's behavior change techniques.
- **Intervention and context.**
  - Behavioral or self-monitoring components, or a lack of blinding to arm.
  - Process measures of the intervention's mechanisms, or process data
    collected before or during outcome assessment.
  - A laboratory setting.

MERIT's Recommendation 7 asks whether these concerns warrant further
action. The options run from none to a Solomon design (French et al.,
2021b, p. 30):

```{r merit, fig.width = 8, fig.height = 4.6, fig.alt = "Decision chart adapted from the MERIT study: the number of red-flag features leads to no further action, to pilot work, studies within a trial, and design changes, or to a Solomon four-group design, all weighed against other concerns."}
boxes <- data.frame(
  x = c(2, 0.2, 2, 3.8, 2),
  y = c(4, 2.5, 2.5, 2.5, 1),
  label = c("Does the trial have red-flag features?",
            "Many present, and\nreactions could seriously\nbias the effect:\nconsider a Solomon design",
            "Several present:\npilot or feasibility work,\nstudies within a trial,\ndesign changes that\nreduce reactivity",
            "Most absent,\nor further data show\nlow risk:\nno further action",
            "Weigh against the research question, recruitment and retention,\nother sources of bias, resources, and ethics")
)
arrows <- data.frame(x = c(2, 2, 2), y = c(3.8, 3.8, 3.8), xend = c(0.3, 2, 3.7), yend = c(3.02, 3.1, 3.02))
links <- data.frame(x = c(0.2, 2, 3.8), y = c(1.98, 1.9, 1.98), xend = c(1.4, 2, 2.6), yend = c(1.28, 1.28, 1.28))
ggplot2::ggplot() +
  ggplot2::geom_segment(data = arrows, ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
                        arrow = ggplot2::arrow(length = ggplot2::unit(0.15, "cm")), colour = "grey40") +
  ggplot2::geom_segment(data = links, ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
                        linetype = "dashed", colour = "grey55") +
  ggplot2::geom_label(data = boxes, ggplot2::aes(x = x, y = y, label = label), size = 3,
                      label.padding = ggplot2::unit(0.35, "lines"), fill = "#eef4f9") +
  ggplot2::scale_x_continuous(limits = c(-0.9, 4.9)) +
  ggplot2::scale_y_continuous(limits = c(0.5, 4.4)) +
  ggplot2::theme_void()
```

*Adapted from Figure 3 of French, D. P., Miles, L. M., Elbourne, D., Farmer,
A., Gulliford, M., Locock, L., Sutton, S., McCambridge, J., & MERIT
Collaborative Group (2021b), "Reducing bias in trials from reactions to
measurement: The MERIT study including developmental work and expert
workshop," Health Technology Assessment, 25(55), NIHR Journals Library,
https://doi.org/10.3310/hta25550, p. 31. The original figure first appeared
in French et al. (2021a). It is licensed under CC BY 4.0
(<https://creativecommons.org/licenses/by/4.0/>). Changes: redrawn, wording
condensed, examples and color coding omitted.*

MERIT's report is cautious about the Solomon design itself. It is
"likely" warranted only when several indicators suggest measurement
reactivity is a major concern (French et al., 2021b, p. 32). Its costs are
the logistics of four arms, a larger sample, and a more complex analysis.
The report also suggests an alternative: a large simple trial with no
baseline measurement, relying on randomization for the equivalence of the
arms (p. 32). That is Campbell and Stanley's (1963/1966) Design 6, which
they considered "greatly underused" and, for many purposes, a better
investment than the Solomon design (p. 26).

## What it costs

The Solomon design asks more of the sample than a two-group trial does. The
sensitization contrast compares two treatment effects, so its estimate is
noisier than either one. `plan_solomon()` shows the consequence. The
comparison assumes a treatment effect of 0.4 standard deviations in the
unpretested groups, a pretest–posttest correlation of 0.5, 80% power, and a
two-sided alpha of .05:

```{r cost}
ate <- plan_solomon(power = 0.80, delta = 0.4, sens = 0, rho = 0.5, estimand = "ate")
sens <- plan_solomon(power = 0.80, delta = 0.4, sens = 0.4, rho = 0.5, estimand = "sensitization")
knitr::kable(data.frame(
  Question = c("Is there a treatment effect of 0.4 SD (averaged over pretesting)?",
               "Does the pretest change that effect by 0.4 SD?"),
  `Participants per group` = c(ate$n1, sens$n1),
  `Total` = c(ate$total_n, sens$total_n),
  check.names = FALSE
), align = "lrr")
```

Detecting sensitization as large as the treatment effect takes about
`r round(sens$total_n / ate$total_n)` times as many participants as
detecting the treatment effect. The article "Planning a Solomon Study"
explains the calculation and gives planning values from the literature.

## Deciding

**solomonR's synthesis of the sources above:** a Solomon design is worth
considering when all three of these hold.

1. **Sensitization is plausible.** The pretest is reactive for your
   participants, measures, and setting (the MERIT red flags).
2. **You need the answer.** Whether the pretest changes the effect matters
   for your conclusions or for later studies. It is not just a nuisance you
   could design away.
3. **You can afford the sample** to estimate sensitization with useful
   precision (`plan_solomon()`).

**If sensitization is unlikely:**
- a pretest-posttest trial analyzed with the pretest as a covariate; or
- a posttest-only randomized trial (Campbell & Stanley's Design 6) answers
  the treatment question at lower cost.

**If sensitization is possible but you cannot afford a Solomon design:**
- MERIT's intermediate options are pilot or feasibility work, a study
  within a trial, and design changes that make the measurement less
  reactive (French et al., 2021b, p. 30);
- a large trial without baseline measurement is another (p. 32).

**If you run a Solomon design,** decide before the analysis how
sensitization will be judged, and report it as the MERIT recommendations
ask. `report_solomon()` includes a design statement for this. The
contemporary analysis is in `vignette("glm-solomon")`, and the historical
procedures are in "A History of the Solomon Design and Its Analysis".

## References

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025a). The dual impact of
pretest sensitisation and the cognitive acceleration through science
education programme in the Solomon four-group design. *Brain Sciences,
16*(1), Article 64. https://doi.org/10.3390/brainsci16010064

French, D. P., Miles, L. M., Elbourne, D., Farmer, A., Gulliford, M.,
Locock, L., Sutton, S., McCambridge, J., & MERIT Collaborative Group.
(2021a). Reducing bias in trials due to reactions to measurement: Experts
produced recommendations informed by evidence. *Journal of Clinical
Epidemiology, 139*, 130–139. https://doi.org/10.1016/j.jclinepi.2021.06.028

French, D. P., Miles, L. M., Elbourne, D., Farmer, A., Gulliford, M.,
Locock, L., Sutton, S., McCambridge, J., & MERIT Collaborative Group.
(2021b). Reducing bias in trials from reactions to measurement: The MERIT
study including developmental work and expert workshop. *Health Technology
Assessment, 25*(55), 1–72. https://doi.org/10.3310/hta25550

Kvalem, I. L., Sundet, J. M., Rivø, K. I., Eilertsen, D. E., & Bakketeig,
L. S. (1996). The effect of sex education on adolescents' use of condoms:
Applying the Solomon four-group design. *Health Education Quarterly,
23*(1), 34–47. https://doi.org/10.1177/109019819602300103

McCambridge, J., Butor-Bhavsar, K., Witton, J., & Elbourne, D. (2011). Can
research assessments themselves cause bias in behaviour change trials? A
systematic review of evidence from Solomon 4-group studies. *PLoS ONE,
6*(10), Article e25223. https://doi.org/10.1371/journal.pone.0025223

Sawilowsky, S. S., & Markman, B. S. (1990b). Rejoinder to Braver and Walton
Braver. *Perceptual and Motor Skills, 71*(2), 424–426.
https://doi.org/10.2466/pms.1990.71.2.424

Solomon, R. L. (1949). An extension of control group design. *Psychological
Bulletin, 46*(2), 137–150. https://doi.org/10.1037/h0062958

Willson, V. L., & Putnam, R. R. (1982). A meta-analysis of pretest
sensitization effects in experimental design. *American Educational Research
Journal, 19*(2), 249–258. https://doi.org/10.3102/00028312019002249
