---
title: "Reanalysis and Synthesis"
description: >
  Reanalyzing a published Solomon study from its summary statistics, and
  computing effect sizes for a meta-analysis.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

Most published Solomon studies report cell means and standard deviations,
not individual data. This article shows two uses of those summaries:
- **Reanalysis.** Checking or extending a study's analysis.
- **Synthesis.** Computing effect sizes that combine across studies.

## Reanalyzing a published study

El Karkri et al. (2025a) report the posttest sample size, mean, and
standard deviation of their four Solomon groups (Table 8, p. 11). The
package bundles them as `elkarkri2025a`. `solomon_from_summary()` fits the
two-by-two posttest model from these numbers:

```{r reanalysis}
elkarkri2025a[, c("group", "n", "mean", "sd")]
fit <- with(elkarkri2025a, solomon_from_summary(n, mean, sd))
fit
```

The published two-way ANOVA is reproduced within rounding. The interaction
F is 11.46, against the published 11.482 (pp. 11–12). Treatment and
pretesting agree just as closely.

The reanalysis adds what the published tests do not show: the simple
treatment effects with confidence intervals. In this study, the treatment
effect is clear in the pretested classes and absent in the unpretested ones.

Summary statistics limit the reanalysis:
- **Variances.** The pooled error variance assumes equal variances in the
  four groups.
- **Individual data.** Robust standard errors, covariate adjustment, and
  the analyses of the pretested groups need the individual data.
- **The design.** Each group of El Karkri et al. is one intact class, so the
  class and the condition are confounded (see `?elkarkri2025a` and
  `baseline_solomon()`).

## Effect sizes for meta-analysis

`solomon_effect_sizes()` computes a standardized treatment effect for each
pair of the design, with its sampling variance in the `yi`/`vi` form that
meta-analysis software reads:

- **The pretested pair.** Morris's (2008) effect size for
  pretest-posttest-control designs, d<sub>ppc2</sub>. It is the difference
  in mean change, divided by the pooled pretest standard deviation, with a
  small-sample correction. Its variance (Eq. 25) needs the pre-post
  correlation. El Karkri et al. do not report it, so it has to come from
  elsewhere, such as the study's authors, a similar study, or a planning
  value.
- **The unpretested pair.** Hedges's g on the posttest (Hedges, 1981).

The correlation matters, so compute the effect sizes under several values:

```{r effect-sizes}
d <- elkarkri2025a
es <- lapply(c(0.3, 0.5, 0.7), function(r) {
  x <- solomon_effect_sizes(n = d$n, mean_post = d$mean, sd_post = d$sd,
                            mean_pre = d$pre_mean[1:2], sd_pre = d$pre_sd[1:2], r = r)
  cbind(r = r, x)
})
do.call(rbind, es)
```

The correlation changes the variance of the pretested effect size, and so
its weight in a meta-analysis.

Two cautions from Morris (2008) and the package's documentation apply:
- **Unreliable variances.** Eq. 25 can underestimate the true variance
  when the treatment inflates posttest variance (p. 380).
- **Different standardizers.** The pretested and unpretested effect sizes
  use different standardizers, so their difference is not a clean measure
  of pretest sensitization. To estimate sensitization itself, use the
  Pretest x Treatment contrast from `solomon_from_summary()`.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025a). The dual impact of
pretest sensitisation and the cognitive acceleration through science
education programme in the Solomon four-group design. *Brain Sciences,
16*(1), Article 64. https://doi.org/10.3390/brainsci16010064

Hedges, L. V. (1981). Distribution theory for Glass's estimator of effect
size and related estimators. *Journal of Educational Statistics, 6*(2),
107–128. https://doi.org/10.3102/10769986006002107

Morris, S. B. (2008). Estimating effect sizes from pretest-posttest-control
group designs. *Organizational Research Methods, 11*(2), 364–386.
https://doi.org/10.1177/1094428106291059
