---
title: "Teaching with solomonR"
description: >
  Six lessons with exercises and solutions for teaching the Solomon
  four-group design: what each group contributes, sensitization, the
  design's history, the historical test sequence, absence of evidence, and
  real studies.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

This article is for instructors who teach the Solomon four-group design in
a research-methods, experimental-design, or statistics course. Each lesson
has three parts:

1. a goal;
2. a short demonstration;
3. an exercise, with a solution you can reveal.

The lessons build on each other but can be taught separately.

The lessons use two kinds of data:

- **Simulated data with known effects.** `simulate_solomon()` generates a
  study with the treatment, pretest, and sensitization effects you choose,
  and records the true values, so students can check whether an analysis
  recovers them.
- **Published studies.** The package bundles several:
  - Solomon's (1949) own spelling experiment (`solomon1949`);
  - a published data set with individual scores (`mai2020`);
  - two published summaries (`elkarkri2025a` and `kvalem1996`).

## Lesson 1: What each group contributes

**Goal.** Students can say what each of the four groups adds, and which
comparison estimates which effect.

Campbell and Stanley (1963/1966, p. 25) summarized the design's strengths:

- it estimates both the main effect of testing and the interaction of
  testing with the treatment;
- it replicates the treatment effect in four comparisons;
- the unpretested control group, compared with the pretests, shows the
  combined effect of maturation and history.

```{r lesson1-design, fig.height = 3.5, fig.alt = "The four groups of the Solomon design: pretest, treatment, and posttest for each."}
plot_solomon_design()
```

**Exercise 1.** Using `solomon_example`, compute the four posttest means.
Then say which difference of means estimates each of these:

1. the treatment effect without a pretest;
2. the treatment effect with a pretest;
3. the effect of the pretest among controls;
4. the Pretest x Treatment interaction.

<details>
<summary>Solution</summary>

```{r lesson1-solution}
m <- with(solomon_example, tapply(y_post, list(treat = treat, pretested = pretested), mean))
m
c(
  treatment_unpretested = m["1", "0"] - m["0", "0"],
  treatment_pretested   = m["1", "1"] - m["0", "1"],
  pretest_among_control = m["0", "1"] - m["0", "0"],
  interaction           = (m["1", "1"] - m["0", "1"]) - (m["1", "0"] - m["0", "0"])
)
```

The interaction is the difference between the two treatment effects. It is
also the difference between the two pretest effects, among treated
participants and among controls, so sensitization can be read either way.
`solomon_example` was simulated with known values (`?solomon_example`): a
treatment effect of 5 points in both conditions, so a true interaction of
0, and a pretest effect of 2 points. The sample differences vary around
those values.

`fit_solomon_glm()` estimates the same contrasts in one model. Read them
from its effects table, where each has its own row:

```{r lesson1-model}
fit <- fit_solomon_glm(y_post, treat, pretested, y_pre, data = solomon_example)
fit$effects[, c("contrast", "estimate", "conf.low", "conf.high")]
```

The model adjusts the pretested groups for the pretest, so its pretest
effects compare the pretested groups, at their mean pretest score, with
the unpretested groups. With random assignment, the unpretested groups'
expected pretest equals that of the pretested groups, and the pretested
groups' mean estimates it, so that is the fair point of comparison
(Solomon & Lessac, 1968, pp. 146–147). Because the mean is an estimate,
the standard errors of the pretest effects include its sampling variance.
Here the two pretested groups started near that mean, so the adjusted
pretest effect among controls is close to the raw difference above.

</details>

## Lesson 2: Building in sensitization

**Goal.** Students see what sensitization looks like, and that the
analysis recovers an effect they built in.

Here the treatment works in both conditions, but less well after a pretest.
That is a negative interaction, the pattern Solomon (1949, p. 145) found:

```{r lesson2-simulate}
d <- simulate_solomon(n = 60, delta = 0.6, sens = -0.4, rho = 0.6, seed = 2)
attr(d, "truth")
fit <- with(d, fit_solomon_glm(y_post, treat, pretested, y_pre))
fit
```

```{r lesson2-plot, fig.height = 4, fig.alt = "Treatment effects among pretested and unpretested participants, with the interaction."}
plot_sensitization(fit)
```

**Exercise 2.** Simulate the same study with `rho = 0` and with
`rho = 0.8`, keeping the seed. Compare the standard errors of the four
contrasts. Which ones change, and why?

<details>
<summary>Solution</summary>

```{r lesson2-solution}
se <- function(rho) {
  d <- simulate_solomon(n = 60, delta = 0.6, sens = -0.4, rho = rho, seed = 2)
  f <- with(d, fit_solomon_glm(y_post, treat, pretested, y_pre))
  # The four treatment contrasts are the first four rows.
  stats::setNames(f$effects$std.error[1:4], f$effects$contrast[1:4])
}
round(rbind(`rho = 0` = se(0), `rho = 0.8` = se(0.8)), 3)
```

The standard error of the effect among unpretested participants hardly
changes, because those participants have no pretest to adjust for; the
small difference comes from the simulated scores themselves. The effect
among pretested participants becomes more precise as `rho` grows: the model adjusts for the
pretest, and a correlated pretest removes part of the error variance. The
interaction and the average effect combine both, so they improve less.

</details>

## Lesson 3: Where the analysis began

**Goal.** Students can reproduce the first published analysis of the
design, and say why it was replaced.

Solomon's (1949) first design had three groups. He inferred the pretest
mean of the unpretested group from the two pretested groups. He then
computed each group's improvement and an interaction term,
I = d1 − (d2 + d3) (p. 143). His fifth-grade spelling experiment
(Table II, p. 144):

```{r lesson3-solomon}
g5 <- solomon1949[solomon1949$grade == 5, ]
g5
fit_solomon_1949(post_mean = g5$mean, pre_mean = g5$pre_mean[1:2], n = g5$n)
```

Points for discussion:

- the groups were "roughly equated" by teachers' judgments, not randomized
  (p. 144);
- there were 10 pupils per group;
- Solomon gave no test for I.

Campbell and Stanley (1963/1966, p. 25) judged his gain-score suggestions
unacceptable, and recommended an analysis of variance of the posttests.

**Exercise 3.** Apply Solomon's four-group formula to `solomon_example`.
Show that I equals the posttest interaction contrast minus the pretest
difference between the two pretested groups. Why does the inferred pretest
drop out?

<details>
<summary>Solution</summary>

```{r lesson3-solution}
s <- with(solomon_example, fit_solomon_1949(y_post, treat, pretested, y_pre))
c(I = s$I, posttest_contrast = s$posttest_contrast,
  pretest_difference = s$pretest_difference,
  check = s$posttest_contrast - s$pretest_difference)
```

Both unpretested groups get the same inferred pretest, i. It is subtracted
once, in d3 = b3 − i, and added back once, in −d4 = −(b4 − i), so it
cancels. With random assignment the pretest difference averages zero, so
Solomon's I and the modern interaction contrast estimate the same effect.

</details>

## Lesson 4: The historical test sequence and its error rate

**Goal.** Students can follow the historical Test A–I sequence, and
explain why a sequence of tests inflates the chance of a false finding.

Walton Braver and Braver (1988) proposed a sequence of tests that ends in a
combination of the pretested and unpretested results. Sawilowsky et al.
(1994) and Sawilowsky (1996) showed by simulation that the sequence, run
when nothing is going on, declares some effect in about 12% to 14% of
studies rather than 5%. The package's replication found 13.5% for the 1995
version of the sequence with normal data and 30 per group (see the article
"Historical Tests: Replicating the Published Error Rates").

```{r lesson4-classic}
classic <- with(solomon_example, fit_solomon_classic(y_post, treat, pretested, y_pre,
                                                     flow = "1995"))
classic$path_string
classic$conclusion
```

**Exercise 4.** Simulate 200 studies in which nothing is going on (every
effect is zero). Count how often the 1995 sequence declares an effect at
any step.

<details>
<summary>Solution</summary>

```{r lesson4-solution}
declares_effect <- function(seed) {
  d <- simulate_solomon(n = 30, seed = seed)
  f <- with(d, fit_solomon_classic(y_post, treat, pretested, y_pre, flow = "1995"))
  p <- vapply(f$tests[f$path], function(t) t$result$p.value[1], numeric(1))
  any(p < 0.05)
}
rate <- mean(vapply(1:200, declares_effect, logical(1)))
c(rate = rate, monte_carlo_se = sqrt(rate * (1 - rate) / 200))
```

With 200 studies the Monte Carlo standard error is about 2.4 percentage
points, so different seeds give rates between about 9% and 18%, all well
above 5%.
Every test in the sequence uses alpha = .05. Each extra test is another
chance for a false positive, and the chances add up. The recommended
analysis fits one model and tests each estimand once.

</details>

## Lesson 5: Absence of evidence

**Goal.** Students can explain why a nonsignificant interaction is not
evidence that sensitization is absent.

Here is a small study with real sensitization of 0.3 standard deviations:

```{r lesson5-small}
d <- simulate_solomon(n = 20, delta = 0.5, sens = 0.3, rho = 0.5, seed = 11)
fit <- with(d, fit_solomon_glm(y_post, treat, pretested, y_pre))
fit$effects[fit$effects$contrast == "Pretest x Treatment", ]
```

The interval is wide. It is compatible with no sensitization and with
sensitization larger than the treatment effect itself. Claiming that
sensitization is negligible needs an equivalence test against a smallest
effect of interest fixed in advance (Lakens, 2017; Lakens et al., 2018):

```{r lesson5-equivalence}
equivalence_solomon(fit, bounds = 0.2)
```

**Exercise 5.** How many participants per group would a study need for
80% power to detect a Pretest x Treatment interaction of 0.3 standard
deviations, with a pretest-posttest correlation of 0.5?

<details>
<summary>Solution</summary>

```{r lesson5-solution}
plan_solomon(power = 0.80, delta = 0.5, sens = 0.3, rho = 0.5, estimand = "sensitization")
```

Interactions need far larger samples than main effects. Willson and Putnam
(1982) found little evidence of pretest-by-treatment interactions across
32 studies, so a planned Solomon study should say which effect it is
powered for. The article "Planning a Solomon Study" covers planning values.

</details>

## Lesson 6: Real studies and their limits

**Goal.** Students can say what a published Solomon study can and cannot
support.

Three bundled studies each teach one limit:

- **`mai2020`: attrition.** A randomized design with published individual
  data (Mai et al., 2020). Between 29% and 46% of each group did not answer
  the posttest. The article "Worked Example: A Published Solomon Study"
  carries it through the package.
- **`elkarkri2025a`: intact classes.** One intact class per group (El Karkri
  et al., 2025a). Class and condition are confounded.
- **`kvalem1996`: clustering.** Binary outcomes from a school trial that
  randomized whole classes but analyzed individuals (Kvalem et al., 1996).

**Exercise 6.** Using `elkarkri2025a`, reproduce the published
Pretest x Treatment F test. Then explain why a significant interaction in
this study does not by itself show pretest sensitization.

<details>
<summary>Solution</summary>

```{r lesson6-solution}
with(elkarkri2025a, solomon_from_summary(n, mean, sd))
pre <- elkarkri2025a[elkarkri2025a$pretested == 1, ]
baseline_solomon(n = pre$n, mean = pre$pre_mean, sd = pre$pre_sd)
```

The interaction F is 11.46, against the published 11.482 (pp. 11–12).
Each group was a single intact class, so any difference between groups can
come from the classes as well as from the treatment and the pretest. The
two pretested classes already differed at pretest. The unpretested classes
have no pretest, so their baseline cannot be checked at all.

</details>

## Using the lessons

- **Order.** Lessons 1, 2, and 5 form a short unit on the design's logic.
  Lessons 3 and 4 add its history, and Lesson 6 its practice.
- **Randomness.** Every simulation takes a seed, so a class sees the same
  numbers as the solutions. Changing the seed shows the sampling
  variability.
- **Further reading.** The article "A History of the Solomon Design and Its
  Analysis" gives the history in full.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

El Karkri, M., Quesada, A., & Romero-Ariza, M. (2025a). The dual impact of
pretest sensitisation and the cognitive acceleration through science
education programme in the Solomon four-group design. *Brain Sciences,
16*(1), Article 64. https://doi.org/10.3390/brainsci16010064

Kvalem, I. L., Sundet, J. M., Rivø, K. I., Eilertsen, D. E., & Bakketeig,
L. S. (1996). The effect of sex education on adolescents' use of condoms:
Applying the Solomon four-group design. *Health Education Quarterly,
23*(1), 34–47. https://doi.org/10.1177/109019819602300103

Lakens, D. (2017). Equivalence tests: A practical primer for t tests,
correlations, and meta-analyses. *Social Psychological and Personality
Science, 8*(4), 355–362. https://doi.org/10.1177/1948550617697177

Lakens, D., Scheel, A. M., & Isager, P. M. (2018). Equivalence testing for
psychological research: A tutorial. *Advances in Methods and Practices in
Psychological Science, 1*(2), 259–269.
https://doi.org/10.1177/2515245918770963

Mai, N. N., Takahashi, Y., & Oo, M. M. (2020). Testing the effectiveness of
transfer interventions using Solomon four-group designs. *Education
Sciences, 10*(4), Article 92. https://doi.org/10.3390/educsci10040092

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
https://digitalcommons.wayne.edu/coe_tbf/29/

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S. (1994).
Meta-analysis and the Solomon four-group design. *The Journal of
Experimental Education, 62*(4), 361–376.
https://doi.org/10.1080/00220973.1994.9944140

Solomon, R. L. (1949). An extension of control group design. *Psychological
Bulletin, 46*(2), 137–150. https://doi.org/10.1037/h0062958

Solomon, R. L., & Lessac, M. S. (1968). A control group design for
experimental studies of developmental processes. *Psychological Bulletin,
70*(3, Pt. 1), 145–150. https://doi.org/10.1037/h0026147

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150

Willson, V. L., & Putnam, R. R. (1982). A meta-analysis of pretest
sensitization effects in experimental design. *American Educational Research
Journal, 19*(2), 249–258. https://doi.org/10.3102/00028312019002249
