---
title: "Validation Evidence"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)

studies <- utils::read.csv("validation-evidence/studies.csv", stringsAsFactors = FALSE)
benchmarks <- utils::read.csv("validation-evidence/benchmarks.csv", stringsAsFactors = FALSE)

pct <- function(x) sprintf("%.0f%%", 100 * x)
num <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

This page gathers the simulation studies that check `solomonR`'s methods.
Each study followed the ADEMP structure of Morris, White and Crowther
(2019):

- its aims, data-generating mechanisms, estimands, methods, performance
  measures, and tolerances were posted to a GitHub issue before any results
  were examined;
- its script, scenario definitions, and results are committed with its
  article.

The evidence covers the scenarios each study examined. It is not a
validation of every Solomon design: non-normal continuous errors and
incidental missing data are outside every study below unless a study says
otherwise. Binary and count outcomes, and cluster-randomized designs, are
covered by their own studies.

## Studies

```{r studies}
knitr::kable(
  data.frame(
    Study = sprintf("[%s](%s)", studies$title, studies$article),
    Protocol = sprintf("[%s](%s)", studies$issues, studies$protocol),
    Scenarios = studies$scenarios,
    Replications = studies$replications,
    Methods = gsub("; ", "<br>", studies$methods),
    `Failed fits` = studies$failed_fits,
    check.names = FALSE
  ),
  escape = FALSE
)
```

## Shared results format

Every study contributes to two tables, which are rebuilt from the study
folders by `validation-evidence/build-benchmarks.R` and can be downloaded
from the
[package repository](https://github.com/JUhalt/solomonR/tree/master/vignettes/articles/validation-evidence):

- **`studies.csv`** has one row per study:
  - the protocol and article;
  - scenario and replication counts;
  - the methods and estimands;
  - the package commit and R version;
  - the run dates;
  - the total number of failed fits.
- **`benchmarks.csv`** has one row per scenario, method, estimand, and
  performance measure:
  - the scenario's design settings;
  - the value and its Monte Carlo standard error (MCSE);
  - the numbers of successful and failed fits;
  - a note.

Two conventions keep the tables honest:

- **Failed fits** are counted separately from the performance measures and
  never silently dropped.
- **A measure a method cannot have** is listed with an empty value and a note
  giving the reason, rather than omitted. For example, the historical Test I
  has no analytic power benchmark, because it combines one-sided p-values.

## Maximum-likelihood inference

`fit_solomon_ml()` was checked with its default Wald inference, its
small-sample option, and the unified GLM for comparison. The registered
tolerances:

- coverage between 0.940 and 0.960;
- Type I error for the sensitization test between 0.040 and 0.060;
- bias within 2 MCSE.

```{r ml}
ml <- benchmarks[benchmarks$study == "ml-inference", ]
coverage <- ml[ml$measure == "coverage", ]
null <- ml[ml$measure == "rejection_rate" & ml$null_effect &
             ml$estimand == "Pretest x Treatment", ]

methods <- unique(ml$method)
knitr::kable(
  data.frame(
    Method = methods,
    `Mean coverage` = vapply(methods, function(m) num(mean(coverage$value[coverage$method == m])), ""),
    `Lowest coverage` = vapply(methods, function(m) num(min(coverage$value[coverage$method == m])), ""),
    `Coverage within 0.940-0.960` = vapply(methods, function(m) {
      v <- coverage$value[coverage$method == m]
      pct(mean(v >= 0.94 & v <= 0.96))
    }, ""),
    `Mean Type I error` = vapply(methods, function(m) num(mean(null$value[null$method == m])), ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Across all scenarios and contrasts; results by cell size are in the full article."
)
```

The default Wald intervals are too narrow in small samples, which is why
`fit_solomon_ml()` warns below 40 participants per cell and offers the
small-sample option. See the
[full article](`r studies$article[studies$study == "ml-inference"]`).

## Power simulation

`power_solomon()` was checked against normal-theory benchmarks. The
registered tolerances:

- Type I error between 0.040 and 0.060 wherever the true effect is zero;
- power within 0.02 of the analytic benchmark, or within 2 MCSE, whichever
  is larger.

```{r power}
pw <- benchmarks[benchmarks$study == "power-simulation", ]
type1 <- pw[pw$measure == "rejection_rate" & pw$null_effect, ]
difference <- pw[pw$measure == "difference_from_benchmark" & !is.na(pw$value) &
                   !pw$null_effect, ]
difference$within <- abs(difference$value) <= pmax(0.02, 2 * difference$mcse)

methods <- unique(pw$method)
knitr::kable(
  data.frame(
    Method = methods,
    `Mean Type I error` = vapply(methods, function(m) {
      v <- type1$value[type1$method == m]
      if (length(v)) num(mean(v)) else "not assessed"
    }, ""),
    `Type I error within 0.040-0.060` = vapply(methods, function(m) {
      v <- type1$value[type1$method == m]
      if (length(v)) pct(mean(v >= 0.04 & v <= 0.06)) else "not assessed"
    }, ""),
    `Power within tolerance of the benchmark` = vapply(methods, function(m) {
      v <- difference$within[difference$method == m]
      if (length(v)) pct(mean(v)) else "no analytic benchmark"
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Across all scenarios; results by allocation are in the full article."
)
```

The 2x2 ANOVA interaction matched its benchmark in every scenario. The GLM
tests are conservative with 20 or fewer participants per cell, following
the HC3 standard errors the package uses by default. See the
[full article](`r studies$article[studies$study == "power-simulation"]`).

## Sample-size planning

`plan_solomon()` was checked by re-simulating the designs it returns with
the package's own GLM test. The registered criterion was that, from 30
participants per cell, re-simulated power falls within 0.02 of the target.

```{r plan}
pl <- benchmarks[benchmarks$study == "plan-resimulation" &
                   benchmarks$measure == "difference_from_target", ]
cells <- sub("^.*cells=", "", pl$design)
smallest <- vapply(strsplit(cells, "/"), function(v) min(as.integer(v)), integer(1))
large <- smallest >= 30

knitr::kable(
  data.frame(
    `Smallest cell` = c("30 or more", "under 30"),
    Plans = c(sum(large), sum(!large)),
    `Within 0.02 of the target` = c(sum(abs(pl$value[large]) <= 0.02),
                                    sum(abs(pl$value[!large]) <= 0.02)),
    `Mean difference` = sprintf("%+.3f", c(mean(pl$value[large]), mean(pl$value[!large]))),
    `Failed fits` = c(sum(pl$n_failed[large]), sum(pl$n_failed[!large])),
    check.names = FALSE
  )
)
```

Below 30 participants per cell, analytic plans run slightly short, following
the HC3 conservatism found for `power_solomon()`. For small designs,
`plan_solomon(method = "simulation")` plans for the package's own test
directly. See the
[full article](`r studies$article[studies$study == "plan-resimulation"]`).

## Binary outcomes

`marginal_solomon()` was checked on the risk-difference, risk-ratio, and
odds-ratio scales, with bootstrap and delta-method intervals, against the
same registered tolerances as above plus a model standard error within 10%
of the empirical one. A scenario in which more than 5% of datasets failed is
outside a method's supported range.

```{r binary}
bi <- benchmarks[benchmarks$study == "binary-outcomes" &
                   grepl("^marginal_solomon", benchmarks$method), ]
reps <- 2000
key <- c("method", "design", "estimand")
wide <- stats::reshape(bi[, c(key, "measure", "value", "mcse", "n_failed", "null_effect")],
                       idvar = key, timevar = "measure", direction = "wide")
wide$supported <- wide$n_failed.coverage <= 0.05 * reps
wide$ok <- abs(wide$value.bias) <= 2 * wide$mcse.bias &
  wide$value.coverage >= 0.94 & wide$value.coverage <= 0.96 &
  abs(wide$value.relative_se_error_pct) <= 10 &
  ifelse(wide$null_effect.coverage, wide$value.rejection_rate >= 0.04 &
           wide$value.rejection_rate <= 0.06, TRUE)
wide$scale <- sub("^.*\\((.*)\\)$", "\\1", wide$estimand)
s <- wide[wide$supported, ]
groups <- unique(s[, c("method", "scale")])
knitr::kable(
  data.frame(
    Method = groups$method,
    Scale = groups$scale,
    `Supported contrasts meeting every tolerance` = vapply(seq_len(nrow(groups)), function(i) {
      v <- s$ok[s$method == groups$method[i] & s$scale == groups$scale[i]]
      sprintf("%d of %d", sum(v), length(v))
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Across supported scenarios; results by design are in the full article."
)
```

Risk differences are validated across the supported range. Risk ratios and
odds ratios are conservative with 20 to 50 participants per cell. With 20
per cell and a control risk of 0.1, too many datasets have a cell without
events for any method. See the
[full article](`r studies$article[studies$study == "binary-outcomes"]`).

## Count outcomes

Poisson fits of `fit_solomon_glm()` and the rate contrasts of
`marginal_solomon()` were checked with and without overdispersion, against
the same tolerances as the binary study.

```{r counts}
co <- benchmarks[benchmarks$study == "count-outcomes", ]
key <- c("method", "design", "estimand")
wide <- stats::reshape(co[, c(key, "measure", "value", "mcse", "n_failed", "null_effect")],
                       idvar = key, timevar = "measure", direction = "wide")
wide$supported <- wide$n_failed.coverage <= 0.05 * 2000
wide$ok <- abs(wide$value.bias) <= 2 * wide$mcse.bias &
  wide$value.coverage >= 0.94 & wide$value.coverage <= 0.96 &
  abs(wide$value.relative_se_error_pct) <= 10 &
  ifelse(wide$null_effect.coverage, wide$value.rejection_rate >= 0.04 &
           wide$value.rejection_rate <= 0.06, TRUE)
wide$overdispersed <- grepl("overdispersion=1", wide$design)
wide$scale <- sub("^.*\\((.*)\\)$", "\\1", wide$estimand)
s <- wide[wide$supported, ]
groups <- unique(s[, c("method", "scale", "overdispersed")])
groups <- groups[order(groups$overdispersed, groups$method), ]
knitr::kable(
  data.frame(
    Counts = ifelse(groups$overdispersed, "overdispersed", "Poisson"),
    Method = groups$method,
    Scale = groups$scale,
    `Supported contrasts meeting every tolerance` = vapply(seq_len(nrow(groups)), function(i) {
      v <- s$ok[s$method == groups$method[i] & s$scale == groups$scale[i] &
                  s$overdispersed == groups$overdispersed[i]]
      sprintf("%d of %d", sum(v), length(v))
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Across supported scenarios; results by design are in the full article."
)
```

Model-based standard errors fail under overdispersion, which is why the
package's default is the robust covariance. See the
[full article](`r studies$article[studies$study == "count-outcomes"]`).

### The negative-binomial option

The NB2 fit of `fit_solomon_glm(family = "negative_binomial")` was checked
on the same datasets, against the same tolerances.

```{r negbin}
nbb <- benchmarks[benchmarks$study == "negative-binomial" & grepl("model contrast", benchmarks$estimand), ]
key <- c("method", "design", "estimand")
wide <- stats::reshape(nbb[, c(key, "measure", "value", "n_failed", "null_effect")],
                       idvar = key, timevar = "measure", direction = "wide")
wide <- wide[wide$n_failed.coverage <= 0.05 * 2000, ]
wide$ok <- wide$value.coverage >= 0.94 & wide$value.coverage <= 0.96 &
  ifelse(wide$null_effect.coverage, wide$value.rejection_rate >= 0.04 &
           wide$value.rejection_rate <= 0.06, TRUE)
wide$overdispersed <- grepl("overdispersion=1", wide$design)
groups <- unique(wide[, c("method", "overdispersed")])
groups <- groups[order(groups$overdispersed, groups$method), ]
knitr::kable(
  data.frame(
    Counts = ifelse(groups$overdispersed, "overdispersed", "Poisson"),
    Method = groups$method,
    `Contrasts within the coverage and Type I tolerances` = vapply(seq_len(nrow(groups)), function(i) {
      v <- wide$ok[wide$method == groups$method[i] & wide$overdispersed == groups$overdispersed[i]]
      sprintf("%d of %d", sum(v), length(v))
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Log rate-ratio contrasts across supported scenarios."
)
```

The NB2 fit with HC3 did slightly better than robust Poisson with strong
overdispersion but fell short of the threshold fixed for recommending it,
so robust Poisson remains the recommendation. Its model-based standard
errors should not be used. See the
[full article](`r studies$article[studies$study == "negative-binomial"]`).

## Clustered designs

The cluster-level randomization tests of `perm_solomon()` were checked for
Type I error in cluster-randomized Solomon designs, with equal and unequal
numbers of treated and control clusters, and compared with the CR2 tests of
`fit_solomon_glm()`. The tolerance is a Type I error from 0.040 to 0.060.

```{r clusters}
cl <- benchmarks[benchmarks$study == "clustered-designs" & benchmarks$null_effect, ]
cl$balanced <- grepl("allocation=balanced", cl$design)
groups <- unique(cl[, c("method", "balanced")])
groups <- groups[order(!groups$balanced, groups$method), ]
knitr::kable(
  data.frame(
    Allocation = ifelse(groups$balanced, "equal", "unequal"),
    Method = groups$method,
    `Within 0.040 to 0.060` = vapply(seq_len(nrow(groups)), function(i) {
      v <- cl$value[cl$method == groups$method[i] & cl$balanced == groups$balanced[i]]
      sprintf("%d of %d", sum(v >= 0.04 & v <= 0.06), length(v))
    }, ""),
    `Largest Type I error` = vapply(seq_len(nrow(groups)), function(i) {
      num(max(cl$value[cl$method == groups$method[i] & cl$balanced == groups$balanced[i]]))
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Null contrasts across both assignment mechanisms, both outcomes, and all variance ratios."
)
```

The permutation tests are exact under the sharp null hypothesis; with four
clusters per arm they are conservative, because few allocations exist. With
unequal numbers of clusters and more variable treated clusters, the
unstudentized difference rejects far too often, as Gail et al. (1996) found;
the studentized statistic reduces but does not remove that excess. CR2 tests
exceed the nominal level with four clusters per arm. See the
[full article](`r studies$article[studies$study == "clustered-designs"]`).

### Marginal risk contrasts from clustered fits

`marginal_solomon()` on logistic fits with CR2 covariance was checked in the
same two cluster-randomized designs, against the coverage and Type I
tolerances.

```{r cluster-marginal}
cm <- benchmarks[benchmarks$study == "cluster-marginal", ]
key <- c("method", "design", "estimand")
wide <- stats::reshape(cm[, c(key, "measure", "value", "null_effect")],
                       idvar = key, timevar = "measure", direction = "wide")
wide$ok <- wide$value.coverage >= 0.94 & wide$value.coverage <= 0.96 &
  ifelse(wide$null_effect.coverage, wide$value.rejection_rate >= 0.04 &
           wide$value.rejection_rate <= 0.06, TRUE)
wide$scale <- sub(".*\\((.*)\\)$", "\\1", wide$estimand)
groups <- unique(wide[, c("method", "scale")])
groups <- groups[order(groups$scale, groups$method), ]
knitr::kable(
  data.frame(
    Scale = groups$scale,
    Method = groups$method,
    `Contrasts within the coverage and Type I tolerances` = vapply(seq_len(nrow(groups)), function(i) {
      v <- wide$ok[wide$method == groups$method[i] & wide$scale == groups$scale[i]]
      sprintf("%d of %d", sum(v), length(v))
    }, ""),
    check.names = FALSE,
    row.names = NULL
  ),
  caption = "Both designs, all allocations, both intracluster correlations."
)
```

With Satterthwaite degrees of freedom, the intervals met the tolerances for
risk differences and odds ratios in almost every scenario and were
conservative for risk ratios with four clusters per cell or arm; a normal
reference was too liberal. See the
[full article](`r studies$article[studies$study == "cluster-marginal"]`).

## Historical Tests A-I

`fit_solomon_classic()` was checked against the published Type I error
rates of the historical sequence (Sawilowsky et al., 1994; Sawilowsky,
1996). Every condition is a complete null, so each rate is a Type I error
rate.

```{r historical}
hc <- benchmarks[benchmarks$study == "historical-tests" &
                   benchmarks$estimand == "any rejection (experiment-wise)" &
                   grepl("distribution=normal; n per group=30; pretest-posttest r=0$", benchmarks$design) &
                   !grepl("alpha_allocation|1990", benchmarks$method), ]
knitr::kable(
  data.frame(Method = hc$method, `Experiment-wise Type I error` = num(hc$value, 3),
             check.names = FALSE, row.names = NULL),
  caption = "Normal data, 30 per group; nominal alpha .05 at each step."
)
```

Tests A to H agreed with the published rates for normal data. Test I
rejected far more often than published, which a post hoc investigation
traced to the published studies' computation of Test I. As defined, the
sequence falsely declares an effect nearly three times as often as its
nominal level. See the
[full article](`r studies$article[studies$study == "historical-tests"]`).

## Latent contrasts: measurement invariance

`invariance_solomon()` reports two published criteria for each step of the
invariance sequence:

- the scaled chi-square difference test (Satorra & Bentler, 2001);
- the change in fit, judged against Chen's (2007) cutoffs.

The study asked whether either criterion is reliable enough in
Solomon-sized groups for `fit_solomon_sem_latent()` to refuse latent mean
contrasts when it flags noninvariance. The table gives the largest
false-rejection rate of each criterion when invariance holds.

```{r invariance}
iv <- benchmarks[benchmarks$study == "invariance-criteria" &
                   benchmarks$measure == "rejection_rate" &
                   grepl("pattern=no noninvariance", benchmarks$design), ]
iv$n <- as.integer(sub(".*n per group=([0-9]+).*", "\\1", iv$design))
iv$criterion <- sub("invariance_solomon(), ", "", iv$method, fixed = TRUE)
tab <- stats::aggregate(value ~ criterion + n, iv, max)
tab <- stats::reshape(tab, idvar = "criterion", timevar = "n", direction = "wide")
names(tab) <- c("Criterion", paste(sub("value.", "", names(tab)[-1], fixed = TRUE), "per group"))
tab[-1] <- lapply(tab[-1], num, digits = 3)
knitr::kable(tab, row.names = FALSE,
             caption = "Largest false-rejection rate across 3, 4, and 6 indicators and both steps; tolerance .060.")
```

No criterion kept false rejections at or below .060, so, under the rules
posted before the run, `fit_solomon_sem_latent()` warns rather than
refuses when a criterion flags noninvariance. The study also found that a
pretest-induced shift common to both pretested groups leaves the
sensitization contrast unbiased, while a shift in one group biases it. See
the
[full article](`r studies$article[studies$study == "invariance-criteria"]`).

## Missing posttests

`fit_solomon_mi()` multiply imputes missing posttests and shifts the
imputed values of each group by an offset, the pattern-mixture sensitivity
analysis of Carpenter et al. (2023, section 10.3). The study generated
posttests missing at random and posttests whose missing values were shifted
by known offsets, and analyzed them with the true offsets. The table gives
the coverage of 95% intervals by group size.

```{r missing-posttests}
mp <- benchmarks[benchmarks$study == "missing-posttests" & benchmarks$measure == "coverage" &
                   grepl("true offsets|delta = 0", benchmarks$method), ]
mp$n <- as.integer(sub(".*n per group=([0-9]+);.*", "\\1", mp$design))
# The analysis with the true offsets: delta = 0 when the posttests are missing at random.
mp <- mp[grepl("true offsets", mp$method) |
           (grepl("delta = 0", mp$method) & grepl("offsets=none", mp$design)), ]
tab <- stats::aggregate(value ~ n, mp, function(v) c(min = min(v), max = max(v)))
knitr::kable(data.frame(`Per group` = tab$n,
                        `Lowest coverage` = num(tab$value[, "min"], digits = 3),
                        `Highest coverage` = num(tab$value[, "max"], digits = 3),
                        check.names = FALSE),
             row.names = FALSE, caption = "Coverage of 95% intervals with the true offsets, over 4 contrasts and 8 scenarios per size; tolerance .940 to .960.")
```

With 60 or more per group, coverage and Type I error were within tolerance
in every scenario; with 30 per group, intervals were conservative. The
pre-specified rule for validation was not met (84 of 96 cells, against 90%),
so the functions stay experimental. The analyses that assume missing at
random biased the sensitization contrast by 0.10 to 0.16 SD when the
departure was confined to one pretested group. See the
[full article](`r studies$article[studies$study == "missing-posttests"]`).

## Adding a study

A new study joins this page by:

1. posting its protocol to a GitHub issue before running;
2. committing its script, `performance.csv`, and `run-information.csv` in a
   folder under `vignettes/articles/`;
3. adding a section for it to `validation-evidence/build-benchmarks.R`.

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Carpenter, J. R., Bartlett, J. W., Morris, T. P., Wood, A. M., Quartagno,
M., & Kenward, M. G. (2023). *Multiple imputation and its application* (2nd
ed.). Wiley. https://doi.org/10.1002/9781119756118

Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of
measurement invariance. *Structural Equation Modeling: A Multidisciplinary
Journal, 14*(3), 464–504. https://doi.org/10.1080/10705510701301834

Gail, M. H., Mark, S. D., Carroll, R. J., Green, S. B., & Pee, D. (1996). On
design considerations and randomization-based inference for community
intervention trials. *Statistics in Medicine, 15*(11), 1069–1092.
<https://doi.org/10.1002/(SICI)1097-0258(19960615)15:11%3C1069::AID-SIM220%3E3.0.CO;2-Q>

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine, 38*(11),
2074–2102. https://doi.org/10.1002/sim.8086

Satorra, A., & Bentler, P. M. (2001). A scaled difference chi-square test
statistic for moment structure analysis. *Psychometrika, 66*(4), 507–514.
https://doi.org/10.1007/BF02296192

Sawilowsky, S. S. (1996, June 23). *Controlling experiment-wise Type I error
of meta-analysis in the Solomon four-group design* [Paper presentation].
First International Conference on Multiple Comparisons, Tel Aviv, Israel.
https://digitalcommons.wayne.edu/coe_tbf/29/

Sawilowsky, S. S., Kelley, D. L., Blair, R. C., & Markman, B. S. (1994).
Meta-analysis and the Solomon four-group design. *The Journal of
Experimental Education, 62*(4), 361–376.
https://doi.org/10.1080/00220973.1994.9944140
