---
title: "Worked Example: A Published Solomon Study"
description: >
  The full analysis path, from checking the design to drafting the report,
  on the published individual data of Mai et al. (2020), checked at each
  step against the published results.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

Mai et al. (2020) tested two interventions meant to help students apply
what they learned in a time-management workshop:

- relapse prevention (RP);
- proximal plus distal goal setting (GS).

They published the individual data with the article under a Creative
Commons Attribution license, and solomonR bundles them as `mai2020` (see
`?mai2020` for the design, the variables, and the license). Because the
published results can be recomputed, each step below can be checked against
the article. This example follows the path of the site's menus: check the
design, reproduce the published analysis, fit the recommended model, ask how
much the missing posttests could matter, and draft the report.

**Acknowledgment.** We thank Nu Nu Mai, Yoshi Takahashi, and Mon Mon Oo
for making their data publicly available with their article. This example
reanalyzes their data to teach the package. For the study itself, its
interventions, its measures, and the authors' conclusions, read the
published article (Mai et al., 2020).

## The design and the data

Final-year students at a management college took a three-hour workshop.

- **Pretesting.** Before the workshop they were "randomly assigned into
  either the pretested or unpretested groups based on their roll numbers"
  (Mai et al., 2020, p. 5). The article does not say how roll numbers were
  used.
- **Interventions.** After the workshop, participants in both pretest
  conditions were randomly assigned to RP, GS, or control (p. 6).
- **Measures.** Self-reported time-management behavior was measured before
  the workshop in the pretested groups and after it in all groups (p. 7).

Pretesting crossed with three conditions gives six groups. The authors
analyzed three Solomon four-group designs, one for each pair of conditions,
reusing groups across them (Table 2, pp. 4–5). This example analyzes the
first pair, RP against control.

More than a third of the participants did not answer the posttest, and
the share differed by group:

```{r data}
with(mai2020, table(condition, pretested, posttest = !is.na(post_behavior)))
rp <- subset(mai2020, condition %in% c("RP", "Control"))
rp$treat <- as.integer(rp$condition == "RP")
```

## Check the design

`validate_solomon()` checks that the data have the structure of a Solomon
design, and `check_solomon_missing()` separates the pretests that are absent
by design from the scores that were lost:

```{r check}
with(rp, check_solomon_missing(post_behavior, treat, pretested, pre_behavior))
```

Two points follow for this study:

- **Posttest attrition.** It ranges from 29% to 46% across the four groups.
  A complete-case analysis is unbiased when missingness is unrelated to the
  outcome given the variables in the model (Little & Rubin, 2019). Attrition
  this uneven makes that assumption worth stating, and the numbers lost in
  each group belong in the report. The section "Missing posttests" below
  asks how much the conclusions depend on it.
- **Baseline.** Only the pretested arms can be compared at baseline:

```{r baseline}
with(rp, baseline_solomon(pre_behavior, treat, pretested))
```

The two arms hardly differed at pretest. Mai et al. (2020, p. 7) compared
all three pretested groups with a one-way ANOVA, F = 0.78, p = .46, which
the data reproduce:

```{r baseline-anova}
summary(aov(pre_behavior ~ condition, data = subset(mai2020, pretested == 1)))
```

The article prints F(2, 117). The data give 115 error degrees of freedom,
which matches the 118 pretested participants of its Table 3 (p. 8).

## Reproduce the published analysis

The authors followed the historical approach: a two-by-two ANOVA of the
posttests to check for pretest sensitization, a check for history effects,
and then an ANCOVA of the pretested groups (pp. 7–9). `fit_solomon_classic()`
runs these procedures, together with the rest of the Test A–I sequence of
Walton Braver and Braver (1988):

```{r classic}
classic <- with(rp, fit_solomon_classic(post_behavior, treat, pretested, pre_behavior))
classic
```

Three published results are reproduced exactly:

- **Test A** is the published sensitization check: F(1, 91) = 3.46,
  p = .066 (Table 4, p. 8).
- **Test E** is the published ANCOVA: its F of 3.00 is the square of the
  t for the mean difference of −0.180 (SE = 0.104, p = .090; Table 7, p. 9).
- **The history check** compares the unpretested control posttest (O6) with
  the pretests of the two pretested groups (O1 and O3). It reproduces the
  differences, t statistics, degrees of freedom, and p values of the first
  and third rows of Table 5 (p. 8). The published signs are not consistent:
  the first row subtracts O6 from the pretest, and the third subtracts the
  pretest from O6. The package always reports O6 minus the pretest.

The second row of Table 5, the only significant history check in the
article, comes from the goal-setting pair:

```{r history-gs}
gs <- subset(mai2020, condition %in% c("GS", "Control"))
with(gs, fit_solomon_classic(post_behavior, as.integer(condition == "GS"),
                             pretested, pre_behavior))$history
```

A difference between O6 and a pretest estimates the combined effect of
history and maturation between the two occasions, if the groups were
equivalent to begin with (Campbell & Stanley, 1963/1966, p. 25). Here the
pretests include every pretested participant, but O6 only the 22 of 35
unpretested controls who answered the posttest. A difference can therefore
also come from who was lost.

## The recommended analysis

The historical sequence tests parts of the design one at a time. The
recommended analysis fits one model to all four groups, as Newman et al.
(1990) proposed (see `vignette("glm-solomon")`). It estimates the Pretest x Treatment
interaction and both simple treatment effects together, adjusts for the
pretest among pretested participants (Lin, 2013), and uses standard errors
that do not assume equal variances:

```{r glm}
fit <- with(rp, fit_solomon_glm(post_behavior, treat, pretested, pre_behavior))
fit
```

The interaction test gives nearly the same p as the published ANOVA (.068
against .066). The model adds the estimate and its interval, which the
published tests do not show:

```{r sd}
sd(mai2020$post_behavior, na.rm = TRUE)
```

The interval for the interaction runs from −0.60 to 0.02 scale points, and
the posttest standard deviation is 0.38. The data are therefore compatible
with no sensitization, and also with an interaction of more than one and a
half standard deviations. A nonsignificant test is not evidence that
sensitization was absent. That claim needs an equivalence test against a
smallest effect of interest fixed in advance (see `equivalence_solomon()`).

Which sign is an improvement depends on how the scale is coded, and the
article and the data file describe the coding differently (see
`?mai2020`). The results are therefore reported here as signed differences,
without calling them better or worse.

## Missing posttests

```{r missing-counts, include = FALSE}
n_missing <- sum(is.na(rp$post_behavior))
lost <- function(t, p) {
  with(rp, c(missing = sum(is.na(post_behavior[treat == t & pretested == p])),
             n = sum(treat == t & pretested == p)))
}
lost_rp_pre <- lost(1, 1)
lost_ctl_pre <- lost(0, 1)
```

Posttests are missing for `r n_missing` of the `r nrow(rp)` participants in
this comparison. The model above is valid if they are missing at random:
if, within each group and, in the pretested groups, at a given pretest
score, whether a posttest is missing does not depend on what it would have
been (Little & Rubin, 2019). Because whether a posttest is missing may then
depend on the pretest, the model also needs the posttest to depend on the
pretest in the same way in both pretested groups, as it assumes. The data
cannot show whether missingness is at random. A sensitivity analysis asks
instead how far the missing posttests would have to depart from that
assumption before a conclusion changed (White et al., 2011).

`fit_solomon_mi()` and `tipping_point_solomon()`, used below, are
experimental: their simulation study did not meet its pre-specified rule
for validation (see `?fit_solomon_mi`). It studied 30, 60, and 120
participants per group with 10% to 30% of posttests missing. With 30 per
group, the size closest to this study's, the intervals were conservative,
and this study's attrition, 29% to 46%, goes beyond the proportions
studied.

### The same assumption, by multiple imputation

`fit_solomon_mi()` imputes the missing posttests in each group, from the
pretest in the pretested groups, and combines the analyses of the completed
data sets. The analyses below use 2,000 imputations rather than the default
100, with a fixed seed so the results can be reproduced. With this much
missing information, 100 imputations leave noticeable Monte Carlo error.
With 100 imputations, the tipping point for sensitization found below
ranged from 0.1 to 0.5 standard deviations across 40 seeds; with 2,000, it
was 0.3 standard deviations for each of the 30 seeds tried.

```{r mi}
mar <- with(rp, fit_solomon_mi(post_behavior, treat, pretested, pre_behavior,
                               m = 2000, seed = 82))
mar
```

```{r mi-compare, include = FALSE}
cc <- fit$effects
mi <- mar$effects
largest_gap <- max(abs(cc$estimate - mi$estimate))
```

Under the same assumption, multiple imputation agrees with the
complete-case model. Theory predicts this when the imputation model is the
same as the analysis model (Carpenter et al., 2023, pp. 255–256). Here the
imputation model lets the pretest slope differ between the two pretested
groups, so the agreement also suggests that the model's common slope does
no harm: the largest difference between the estimates is
`r sprintf("%.3f", largest_gap)` scale points. The fraction of missing
information (`fmi`) of `r sprintf("%.2f", min(mi$fmi))` to
`r sprintf("%.2f", max(mi$fmi))` shows how much the missing posttests weigh
on each contrast.

### The treatment effect

`tipping_point_solomon()` repeats the imputation with the imputed posttests
of chosen groups shifted by an offset, over a range of offsets: by default,
21 offsets from minus to plus one pooled within-group standard deviation of
the observed posttests. For the average effect of relapse prevention, the
offsets go to both treatment groups:

```{r tipping-ate}
tp <- with(rp, tipping_point_solomon(post_behavior, treat, pretested, pre_behavior,
                                     groups = "treatment", m = 2000, seed = 82))
tp$tipping_sd
```

```{r tipping-ate-text, include = FALSE}
ate_text <- if (all(is.na(tp$tipping_sd))) {
  paste("Neither value is reached (NA): the conclusion did not change within",
        "the range tried. The average effect stays nonsignificant even if the",
        "missing posttests of both relapse-prevention groups were a full",
        "standard deviation higher, or lower, than the imputation model predicts.")
} else {
  sprintf("The conclusion about the average effect changes at %s standard deviations.",
          paste(stats::na.omit(round(tp$tipping_sd, 1)), collapse = " and "))
}
```

`r ate_text`

### Sensitization

The Pretest x Treatment contrast is closer to the line. An offset confined
to any one group moves it by about the offset times the share of that
group's posttests that are missing, so the group with the most missing
posttests can change the conclusion at the smallest offset. Here the offsets
go to the pretested relapse-prevention group alone (`groups = 1`), where
`r lost_rp_pre[["missing"]]` of `r lost_rp_pre[["n"]]` posttests are
missing:

```{r tipping-sens, fig.width = 7, fig.height = 4.5, fig.alt = "Pooled Pretest x Treatment contrast with its confidence band as the offset added to the imputed posttests of the pretested relapse-prevention group goes from minus to plus one pooled within-group standard deviation."}
tp_sens <- with(rp, tipping_point_solomon(post_behavior, treat, pretested, pre_behavior,
                                          contrast = "Pretest x Treatment", groups = 1,
                                          m = 2000, seed = 82))
tp_sens$tipping_sd
plot_tipping_point(tp_sens)
```

```{r tipping-sens-values, include = FALSE}
sens_mar <- tp_sens$results[abs(tp_sens$results$delta) < 1e-12, ]
neg <- tp_sens$tipping_sd[["negative"]]
pos <- tp_sens$tipping_sd[["positive"]]
tp_ctl <- with(rp, tipping_point_solomon(post_behavior, treat, pretested, pre_behavior,
                                         contrast = "Pretest x Treatment", groups = 2,
                                         m = 2000, seed = 82))
ctl <- stats::na.omit(tp_ctl$tipping_sd)
```

Assuming missing at random, the contrast is not significant
(p = `r sub("^0", "", sprintf("%.3f", sens_mar$p.value))`).
`r if (!is.na(neg)) sprintf("It would be significant at the .05 level if the missing posttests of the pretested relapse-prevention group were about %.1f standard deviations lower on the scale than the imputation model predicts.", abs(neg)) else "It stays nonsignificant for every lower offset tried."`
`r if (!is.na(pos)) sprintf("It would also be significant if they were about %.1f standard deviations higher.", pos) else "No higher offset in the range changes the conclusion."`
`r if (length(ctl)) sprintf("The same analysis for the pretested control group (groups = 2), where %d of %d posttests are missing, changes the conclusion at %s standard deviations.", lost_ctl_pre[["missing"]], lost_ctl_pre[["n"]], paste(sprintf("%+.1f", ctl), collapse = " and ")) else sprintf("The same analysis for the pretested control group (groups = 2), where %d of %d posttests are missing, does not change the conclusion within the range.", lost_ctl_pre[["missing"]], lost_ctl_pre[["n"]])`

Whether departures of these sizes are plausible is a substantive judgment,
and White et al. (2011) advise specifying sensitivity analyses in detail
before the unblinded data are seen. Two facts bear on it here. Many
posttests are missing in both pretested groups: almost a third in the
relapse-prevention group and nearly half in the control group. And because
the article and the data file describe the scale's direction differently
(see `?mai2020`), "lower" and "higher" cannot be read as better or worse. A report would state the
conclusion under missing at random and add how large a departure, in which
group, would change it.

`report_solomon()` writes the method, the offsets explored, and the tipping
points, with their references:

```{r report-tipping}
report_solomon(tp_sens)
```

## Draft the report

`report_solomon()` drafts the method and results. Given the numbers in each
group, it also reports attrition by group. The numbers are given in the
order pretested treatment, pretested control, unpretested treatment, and
unpretested control (see the first table):

```{r report}
report_solomon(fit, design = list(randomized = c(35, 50, 31, 35),
                                  assignment = "random"))
```

Here the "randomized" numbers are the participants in the published data
file. The article reports that 310 students attended the workshops and that
data were collected from 210 (pp. 5–6), so the file does not show how many
were assigned to each group. A report of a new study would give the numbers
assigned. It would also state whether the sensitization analysis was
pre-specified, which the report never infers.

## What the example shows

The package reproduces the published tests, and the recommended model adds
what those tests leave out:

- estimates with confidence intervals;
- the interaction and both simple effects in one model;
- attrition reported by group.

The experimental functions `fit_solomon_mi()` and `tipping_point_solomon()`
add a sensitivity analysis of how much the conclusions about the average
effect and about sensitization depend on why posttests are missing.

The data also show limits that no analysis removes:

- **Attrition.** The posttest was missing for 29% to 46% of each group.
  The sensitivity analysis shows how far the conclusions about the average
  effect and sensitization depend on why; it cannot remove that dependence.
- **Pretest assignment.** The article does not describe how roll numbers
  were used.
- **Scale direction.** It is described differently in the article and in
  the data file.

Two points about the article remain:

- **Table 6.** Its cell statistics (p. 9) are not reproduced: they do not
  match the data ([#53](https://github.com/JUhalt/solomonR/issues/53)).
- **Overlapping comparisons.** The three four-group comparisons reuse
  groups, so they are not independent. `fit_solomon_glm(control = "Control")`
  fits a single model to all six groups; see "Designs With Several
  Treatments" ([#45](https://github.com/JUhalt/solomonR/issues/45)).

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

Carpenter, J. R., Bartlett, J. W., Morris, T. P., Wood, A. M., Quartagno,
M., & Kenward, M. G. (2023). *Multiple imputation and its application* (2nd
ed.). Wiley. https://doi.org/10.1002/9781119756118

Lin, W. (2013). Agnostic notes on regression adjustments to experimental
data: Reexamining Freedman's critique. *The Annals of Applied Statistics,
7*(1), 295–318. https://doi.org/10.1214/12-AOAS583

Little, R. J. A., & Rubin, D. B. (2019). *Statistical analysis with missing
data* (3rd ed.). Wiley. https://doi.org/10.1002/9781119482260

Mai, N. N., Takahashi, Y., & Oo, M. M. (2020). Testing the effectiveness of
transfer interventions using Solomon four-group designs. *Education
Sciences, 10*(4), Article 92. https://doi.org/10.3390/educsci10040092

Newman, I., Benz, C., & Williams, J. D. (1990). Alternatives in analyzing the
Solomon four group design. *Multiple Linear Regression Viewpoints, 17*(2),
91–103. https://ojs.lib.ua.edu/glmj/article/view/125

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150

White, I. R., Horton, N. J., Carpenter, J., & Pocock, S. J. (2011). Strategy
for intention to treat analysis in randomised trials with missing outcome
data. *BMJ, 342*, Article d40. https://doi.org/10.1136/bmj.d40
