---
title: "Worked Example: A Published Solomon Study"
description: >
  The full analysis path, from checking the design to drafting the report,
  on the published individual data of Mai et al. (2020), checked at each
  step against the published results.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(solomonR)
```

Mai et al. (2020) tested two interventions meant to help students apply
what they learned in a time-management workshop:

- relapse prevention (RP);
- proximal plus distal goal setting (GS).

They published the individual data with the article under a Creative
Commons Attribution license, and solomonR bundles them as `mai2020` (see
`?mai2020` for the design, the variables, and the license). Because the
published results can be recomputed, each step below can be checked against
the article. This example follows the path of the site's menus: check the
design, reproduce the published analysis, fit the recommended model, and
draft the report.

## The design and the data

Final-year students at a management college took a three-hour workshop.

- **Pretesting.** Before the workshop they were "randomly assigned into
  either the pretested or unpretested groups based on their roll numbers"
  (Mai et al., 2020, p. 5). The article does not say how roll numbers were
  used.
- **Interventions.** After the workshop, participants in both pretest
  conditions were randomly assigned to RP, GS, or control (p. 6).
- **Measures.** Self-reported time-management behavior was measured before
  the workshop in the pretested groups and after it in all groups (p. 7).

Pretesting crossed with three conditions gives six groups. The authors
analyzed three Solomon four-group designs, one for each pair of conditions,
reusing groups across them (Table 2, pp. 4–5). This example analyzes the
first pair, RP against control.

More than a third of the participants did not answer the posttest, and
the share differed by group:

```{r data}
with(mai2020, table(condition, pretested, posttest = !is.na(post_behavior)))
rp <- subset(mai2020, condition %in% c("RP", "Control"))
rp$treat <- as.integer(rp$condition == "RP")
```

## Check the design

`validate_solomon()` checks that the data have the structure of a Solomon
design, and `check_solomon_missing()` separates the pretests that are absent
by design from the scores that were lost:

```{r check}
with(rp, check_solomon_missing(post_behavior, treat, pretested, pre_behavior))
```

Two points follow for this study:

- **Posttest attrition.** It ranges from 29% to 46% across the four groups.
  A complete-case analysis is unbiased when missingness is unrelated to the
  outcome given the variables in the model (Little & Rubin, 2019). Attrition
  this uneven makes that assumption worth stating, and the numbers lost in
  each group belong in the report.
- **Baseline.** Only the pretested arms can be compared at baseline:

```{r baseline}
with(rp, baseline_solomon(pre_behavior, treat, pretested))
```

The two arms hardly differed at pretest. Mai et al. (2020, p. 7) compared
all three pretested groups with a one-way ANOVA, F = 0.78, p = .46, which
the data reproduce:

```{r baseline-anova}
summary(aov(pre_behavior ~ condition, data = subset(mai2020, pretested == 1)))
```

The article prints F(2, 117). The data give 115 error degrees of freedom,
which matches the 118 pretested participants of its Table 3 (p. 8).

## Reproduce the published analysis

The authors followed the historical approach: a two-by-two ANOVA of the
posttests to check for pretest sensitization, a check for history effects,
and then an ANCOVA of the pretested groups (pp. 7–9). `fit_solomon_classic()`
runs these procedures, together with the rest of the Test A–I sequence of
Walton Braver and Braver (1988):

```{r classic}
classic <- with(rp, fit_solomon_classic(post_behavior, treat, pretested, pre_behavior))
classic
```

Three published results are reproduced exactly:

- **Test A** is the published sensitization check: F(1, 91) = 3.46,
  p = .066 (Table 4, p. 8).
- **Test E** is the published ANCOVA: its F of 3.00 is the square of the
  t for the mean difference of −0.180 (SE = 0.104, p = .090; Table 7, p. 9).
- **The history check** compares the unpretested control posttest (O6) with
  the pretests of the two pretested groups (O1 and O3). It reproduces the
  differences, t statistics, degrees of freedom, and p values of the first
  and third rows of Table 5 (p. 8). The published signs are not consistent:
  the first row subtracts O6 from the pretest, and the third subtracts the
  pretest from O6. The package always reports O6 minus the pretest.

The second row of Table 5, the only significant history check in the
article, comes from the goal-setting pair:

```{r history-gs}
gs <- subset(mai2020, condition %in% c("GS", "Control"))
with(gs, fit_solomon_classic(post_behavior, as.integer(condition == "GS"),
                             pretested, pre_behavior))$history
```

A difference between O6 and a pretest estimates the combined effect of
history and maturation between the two occasions, if the groups were
equivalent to begin with (Campbell & Stanley, 1963/1966, p. 25). Here the
pretests include every pretested participant, but O6 only the 22 of 35
unpretested controls who answered the posttest. A difference can therefore
also come from who was lost.

## The recommended analysis

The historical sequence tests parts of the design one at a time. The
recommended analysis fits one model to all four groups
(`vignette("glm-solomon")`). It estimates the Pretest x Treatment
interaction and both simple treatment effects together, adjusts for the
pretest among pretested participants (Lin, 2013), and uses standard errors
that do not assume equal variances:

```{r glm}
fit <- with(rp, fit_solomon_glm(post_behavior, treat, pretested, pre_behavior))
fit
```

The interaction test gives nearly the same p as the published ANOVA (.068
against .066). The model adds the estimate and its interval, which the
published tests do not show:

```{r sd}
sd(mai2020$post_behavior, na.rm = TRUE)
```

The interval for the interaction runs from −0.60 to 0.02 scale points, and
the posttest standard deviation is 0.38. The data are therefore compatible
with no sensitization, and also with an interaction of more than one and a
half standard deviations. A nonsignificant test is not evidence that
sensitization was absent. That claim needs an equivalence test against a
smallest effect of interest fixed in advance (see `equivalence_solomon()`).

Which sign is an improvement depends on how the scale is coded, and the
article and the data file describe the coding differently (see
`?mai2020`). The results are therefore reported here as signed differences,
without calling them better or worse.

## Draft the report

`report_solomon()` drafts the method and results. Given the numbers in each
group, it also reports attrition by group. The numbers are given in the
order pretested treatment, pretested control, unpretested treatment, and
unpretested control (see the first table):

```{r report}
report_solomon(fit, design = list(randomized = c(35, 50, 31, 35),
                                  assignment = "random"))
```

Here the "randomized" numbers are the participants in the published data
file. The article reports that 310 students attended the workshops and that
data were collected from 210 (pp. 5–6), so the file does not show how many
were assigned to each group. A report of a new study would give the numbers
assigned. It would also state whether the sensitization analysis was
pre-specified, which the report never infers.

## What the example shows

The package reproduces the published tests, and the recommended model adds
what those tests leave out:

- estimates with confidence intervals;
- the interaction and both simple effects in one model;
- attrition reported by group.

The data also show limits that no analysis removes:

- **Attrition.** The posttest was missing for 29% to 46% of each group.
- **Pretest assignment.** The article does not describe how roll numbers
  were used.
- **Scale direction.** It is described differently in the article and in
  the data file.

Two parts of the article are not reproduced:

- **Table 6.** Its cell statistics (p. 9) do not match the data
  ([#53](https://github.com/JUhalt/solomonR/issues/53)).
- **Overlapping comparisons.** The three four-group comparisons reuse
  groups, so they are not independent. A single model for all six groups
  is planned ([#45](https://github.com/JUhalt/solomonR/issues/45)).

All works cited in solomonR are listed, with notes on how the package uses
them, on the [References](references.html) page.

## References

Campbell, D. T., & Stanley, J. C. (1966). *Experimental and
quasi-experimental designs for research*. Rand McNally. (Original work
published 1963)

Lin, W. (2013). Agnostic notes on regression adjustments to experimental
data: Reexamining Freedman's critique. *The Annals of Applied Statistics,
7*(1), 295–318. https://doi.org/10.1214/12-AOAS583

Little, R. J. A., & Rubin, D. B. (2019). *Statistical analysis with missing
data* (3rd ed.). Wiley. https://doi.org/10.1002/9781119482260

Mai, N. N., Takahashi, Y., & Oo, M. M. (2020). Testing the effectiveness of
transfer interventions using Solomon four-group designs. *Education
Sciences, 10*(4), Article 92. https://doi.org/10.3390/educsci10040092

Walton Braver, M. C., & Braver, S. L. (1988). Statistical treatment of the
Solomon four-group design: A meta-analytic approach. *Psychological
Bulletin, 104*(1), 150–154. https://doi.org/10.1037/0033-2909.104.1.150
